---
title: "Introduction to R - Interactive Workshop"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome! 👋

Welcome to your first session "Introduction to R" in the Institute Ruder Boskovic (IRB) workshop. 

### Authors & Affiliations

**dr.sc. Paula Štancl** - Bioinformatics group, Faculty of Science  
**dr.sc. Andrea Gelemanović** - MedILS, UNIST

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## R as Calculator

R can be used as a basic calculator to perform arithmetic operations.

### Try It Out!

Run this code to see basic calculations:

```{r calc-demo, exercise=TRUE}
# Addition
(9 + 2) * (3 - 1)

# Square root
sqrt(2)^2

# Division
5 / 2
```

### Exercise: Integer Division

Try these operations with the number 17:

- Integer division by 5 (`%/%`)
- Remainder (modulo) with 5 (`%%`)

```{r int-div, exercise=TRUE}
# Write your code here


```

```{r int-div-hint-1}
# Use %/% for integer division
17 %/% 5
```

```{r int-div-hint-2}
# Use %% for remainder
17 %% 5
```

```{r int-div-solution}
# Integer division
17 %/% 5

# Remainder
17 %% 5
```

---

## 🎯 Task: Stem Cell Division

A scientist has 48 stem cells. Each one divides into 2 daughter cells.

Calculate:
1. Total cells after division
2. Full groups of 10 cells
3. Leftover cells

### Your Turn!

```{r stem-cells, exercise=TRUE}
# Number of initial cells
initial_cells <- 48

# Calculate total cells after division


# Calculate full groups of 10


# Calculate leftover cells


```

```{r stem-cells-hint-1}
# Multiply by 2 for division
total_cells <- initial_cells * 2
```

```{r stem-cells-hint-2}
# Use %/% for full groups
full_groups <- total_cells %/% 10
```

```{r stem-cells-hint-3}
# Use %% for leftovers
leftover <- total_cells %% 10
```

```{r stem-cells-solution}
# Number of initial cells
initial_cells <- 48

# Calculate total cells after division
total_cells <- initial_cells * 2
total_cells

# Calculate full groups of 10
full_groups <- total_cells %/% 10
full_groups

# Calculate leftover cells
leftover <- total_cells %% 10
leftover
```

---

## Variables

### Creating Variables

To create a variable, use the assignment operator `<-`:

```{r var-demo, exercise=TRUE}
# Create a variable
vec <- 5

# Print it
vec

# Create and print in one step
(vec2 <- 10)
```

### Exercise: Create Your Own Variables

Create three variables:
- `my_age` with your age
- `my_name` with your name (in quotes)
- `is_student` with TRUE or FALSE

```{r create-vars, exercise=TRUE}
# Create your variables here




```

```{r create-vars-solution}
# Create your variables
my_age <- 25
my_name <- "Andrea"
is_student <- TRUE

# Print them
my_age
my_name
is_student
```

---

## Vectors

Vectors are the most common data type in R. Use `c()` to combine values:

```{r vector-demo, exercise=TRUE}
# Numeric vector
numbers <- c(1, 2, 5)
numbers

# Character vector
animals <- c("human", "mouse", "dog", "cat")
animals

# Sequence
sequence <- 1:10
sequence
```

### Exercise: Create Your Own Vectors

Create these vectors:
1. `temperatures` with values: 18, 22, 30, 15
2. `days` with: "Mon", "Tue", "Wed", "Thu"
3. `numbers` from 5 to 20 using `:`

```{r create-vectors, exercise=TRUE}
# Create your vectors here




```

```{r create-vectors-solution}
# Create vectors
temperatures <- c(18, 22, 30, 15)
days <- c("Mon", "Tue", "Wed", "Thu")
numbers <- 5:20

# Print them
temperatures
days
numbers
```

---

## Vector Operations

R performs operations on entire vectors at once!

```{r vector-ops-demo, exercise=TRUE}
# Create vectors
vec1 <- c(1, 2, 3)
vec2 <- c(10, 20, 30)

# Add them
vec1 + vec2

# Multiply
vec1 * vec2

# Add a single number (recycling)
vec1 + 100
```

### Exercise: Temperature Conversion

Convert these Celsius temperatures to Fahrenheit using the formula: `F = C * 9/5 + 32`

```{r temp-convert, exercise=TRUE}
celsius <- c(0, 10, 20, 30, 40)

# Convert to Fahrenheit
fahrenheit <- 

fahrenheit
```

```{r temp-convert-hint}
# Formula: F = C * 9/5 + 32
fahrenheit <- celsius * 9/5 + 32
```

```{r temp-convert-solution}
celsius <- c(0, 10, 20, 30, 40)

# Convert to Fahrenheit
fahrenheit <- celsius * 9/5 + 32
fahrenheit
```

---

## Subsetting Vectors

Access specific elements using `[]`:

```{r subset-demo, exercise=TRUE}
animals <- c("dog", "cat", "bird", "fish", "hamster")

# First element
animals[1]

# Elements 2 to 4
animals[2:4]

# Multiple specific elements
animals[c(1, 3, 5)]

# Exclude element 2
animals[-2]
```

### Exercise: Extract Elements

Given this vector, extract:
1. The 3rd element
2. Elements 2, 4, and 5
3. All except the first element

```{r subset-exercise, exercise=TRUE}
scores <- c(85, 92, 78, 95, 88)

# Extract 3rd element


# Extract elements 2, 4, and 5


# All except first


```

```{r subset-exercise-solution}
scores <- c(85, 92, 78, 95, 88)

# Extract 3rd element
scores[3]

# Extract elements 2, 4, and 5
scores[c(2, 4, 5)]

# All except first
scores[-1]
```

---

## Logical Subsetting

Use logical conditions to filter vectors:

```{r logical-demo, exercise=TRUE}
numbers <- c(5, 12, 8, 20, 3)

# Find numbers greater than 10
numbers > 10

# Extract numbers greater than 10
numbers[numbers > 10]

# Multiple conditions
numbers[numbers > 5 & numbers < 15]
```

### Exercise: Filter Temperatures

Find all temperatures that are "comfortable" (between 20 and 25 degrees):

```{r filter-temps, exercise=TRUE}
temps <- c(18, 22, 30, 15, 24, 21, 28)

# Extract comfortable temperatures (20-25)


```

```{r filter-temps-hint}
# Use & to combine conditions
temps[temps >= 20 & temps <= 25]
```

```{r filter-temps-solution}
temps <- c(18, 22, 30, 15, 24, 21, 28)

# Extract comfortable temperatures
comfortable <- temps[temps >= 20 & temps <= 25]
comfortable
```

---

## Useful Functions

R has many built-in functions for vectors:

```{r functions-demo, exercise=TRUE}
data <- c(5, 12, 8, 20, 3, 15)

# Summary statistics
mean(data)    # Average
sum(data)     # Total
sd(data)      # Standard deviation
length(data)  # Number of elements

# Summary of all
summary(data)
```

### Exercise: Analyze Gene Expression

Calculate statistics for these gene expression values:

```{r gene-stats, exercise=TRUE}
gene_expression <- c(145, 203, 187, 156, 198, 177, 165, 189)

# Calculate:
# 1. Mean expression
# 2. Maximum value
# 3. Minimum value
# 4. Standard deviation





```

```{r gene-stats-solution}
gene_expression <- c(145, 203, 187, 156, 198, 177, 165, 189)

# Mean expression
mean(gene_expression)

# Maximum value
max(gene_expression)

# Minimum value
min(gene_expression)

# Standard deviation
sd(gene_expression)
```

---

## 🎊 Congratulations!

You've completed the Introduction to R workshop!

### What You've Learned:

✅ Using R as a calculator  
✅ Creating and manipulating variables  
✅ Working with vectors  
✅ Subsetting and filtering data  
✅ Using built-in functions  

### Next Steps:

Continue to **Day 1 Part 2: Functions and Conditions** to learn about:
- Creating custom functions
- Conditional statements (if/else)
- Apply family functions

---

## Additional Resources

- [R Documentation](https://www.r-project.org/other-docs.html)
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)
- [R for Data Science](https://r4ds.had.co.nz/)
