---
title: "Data.table in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Server setup (empty - download handled by JavaScript)
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
ddf <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
ddf_dt <- as.data.table(ddf)
iris_dt <- as.data.table(iris)
```

## Welcome

Welcome to **Data.table in R**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

<div class="callout callout-info">
<div class="callout-title">How to Use This Tutorial</div>

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises
</div>

<div class="callout callout-success">
<div class="callout-title">Why data.table?</div>

- **Extremely fast** - optimized for speed with large datasets  
- **Concise syntax** - less typing, more doing  
- **Powerful** - complex operations in single expressions  
- **Memory efficient** - works by reference, not copies
</div>

---

## Getting Started

<div class="callout callout-info">
<div class="callout-title">What is data.table?</div>

`data.table` is a high-performance package for working with tabular data in R. It's **extremely fast** and uses concise syntax: `DT[i, j, by]` where `i` = rows, `j` = columns, `by` = groups.
</div>

### Introduction

### Sample Data

We'll use this dataset for examples:

```{r data-preview, exercise=TRUE, exercise.setup="shared-data"}
ddf
```

And you'll practice on the `iris` dataset:

```{r iris-preview, exercise=TRUE}
iris
```

---

### Loading the Library

### Installation

If you don't have the package, install it (only once):

```{r install-dt, exercise=TRUE, eval=FALSE}
install.packages("data.table")
```

### Loading

Load it every session:

```{r load-dt, exercise=TRUE}
library(data.table)
```

### Converting to data.table

If you already have a data frame:

```{r convert-dt, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt <- as.data.table(ddf)
ddf_dt
```

---

### data.table Syntax

### All You Need to Know:

### **DT[i, j, by]**

- **i:** select those rows
- **j:** do this to them
- **by:** do it per groups

---

## Basic Operations

<div class="callout callout-concept">
<div class="callout-title">The DT[i, j, by] Syntax</div>

**The core syntax:** `DT[i, j, by]`
- **i** = which rows (row filter/subset)
- **j** = what to do with columns (select, compute, modify)
- **by** = group by (aggregation groups)

This unified syntax makes data.table incredibly powerful and consistent!
</div>

### Selecting Rows

### Example: Height > 180

How many people are taller than 180?

```{r select-rows, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[Height > 180]
nrow(ddf_dt[Height > 180])
```

<div class="callout callout-tip">
<div class="callout-title">Familiar Syntax!</div>

You select rows similar to vector subsetting - no need for `ddf_dt$Height`!

data.table knows to look for column names within the data.table.
</div>

---

### Selecting Columns

### Show Age and Continent of First 5 People

<div class="callout callout-concept">
<div class="callout-title">The Magic Dot: .()</div>

In data.table use **`.`** instead of `c()`

`.()` is shorthand for `list()` - it creates a list of columns to select.
</div>

In data.table use `.` instead of `c()` (`.` is short for `list()`):

```{r select-cols, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[1:5, .(Age, Continent)]
```

### Using Column Names as Variables

<div class="callout callout-tip">
<div class="callout-title">Two Dots = Variable!</div>

Use the `..` prefix to reference column names stored in variables. The two dots tell data.table: "look up one level in the calling environment".
</div>

```{r select-cols-vars, exercise=TRUE, exercise.setup="shared-data"}
cnames <- c("Age", "Continent")
ddf_dt[1:5, ..cnames]
```

---

### More Column Selection

### Range of Columns

```{r select-range, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[1:3, Age:Continent]
```

### Exclude Columns

```{r select-exclude, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[1:3, -(Age:Continent)]
```

### Negate Selection

```{r select-negate, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[1:3, !(Age:Gender)]
```

---

### Exercise 1: iris Basics

Practice with the `iris` dataset!

**Tasks:**
1. Convert `iris` to data.table - call it `iris_dt`
2. Select all rows with Sepal.Length < 6.7
3. Select as before, but show only Sepal.Length and Species columns

```{r exercise-iris-1, exercise=TRUE}
# Convert iris to data.table
iris_dt <- as.data.table(iris)

# Select rows with Sepal.Length < 6.7


# Show only specific columns

```

```{r exercise-iris-1-hint}
# Step 2: iris_dt[Sepal.Length < 6.7]
# Step 3: iris_dt[Sepal.Length < 6.7, .(Sepal.Length, Species)]
```

```{r exercise-iris-1-solution}
iris_dt <- as.data.table(iris)
iris_dt[Sepal.Length < 6.7]
iris_dt[Sepal.Length < 6.7, .(Sepal.Length, Species)]
```

---

## Aggregations and Groups

<div class="callout callout-success">
<div class="callout-title">Group Operations</div>

The `by` parameter is where data.table truly shines. Compute statistics **by groups** with lightning-fast performance - essential for data analysis!
</div>

### Operations on Columns

### Calculate Mean

```{r calc-mean, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, mean(Height)]
```

### Mean by Groups

```{r mean-by-group, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, mean(Height), by = Gender]
```

---

### Multiple Statistics by Group

Calculate mean, sd, max, min for Height, and count per Gender, ordered by mean:

```{r multiple-stats, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, 
       .(mean = mean(Height),
         sd = sd(Height),
         min_x = min(Height),
         max_x = max(Height),
         N = .N),
       by = Gender][order(mean)]
```

---

### Special Symbols: .N

<div class="callout callout-concept">
<div class="callout-title">Special Symbol: .N</div>

`.N` gives the **number of observations** in the current group. It's incredibly useful for counting and is one of data.table's most frequently used special symbols.
</div>

### .N gives number of observations

Use `.N` to count rows in your data or within groups:

```{r dot-n, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .N]
ddf_dt[, .N, by = Gender]
```

### Group by Multiple Variables

<div class="callout callout-tip">
<div class="callout-title">Dynamic Grouping</div>

You can group by **expressions**, not just column names! This allows for flexible on-the-fly categorization like `Height > 180` without creating new columns.
</div>

```{r multiple-groups, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .N, by = .(Gender, Height > 180)]
```

### Exercise: Count by Group

How many men and women are there from different continents?

```{r exercise-count, exercise=TRUE, exercise.setup="shared-data"}
# Count by Gender and Continent

```

```{r exercise-count-solution}
ddf_dt[, .N, by = .(Gender, Continent)]
```

---

### Exercise 2: iris Operations

**Tasks:**
1. Select all rows where Sepal.Length < 6.7 AND Species == "virginica"
2. For those, use chaining `[][]` to calculate mean Petal.Width
3. How many flowers have Sepal.Width > 3 and how many have Sepal.Width < 3?

```{r exercise-iris-2, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Task 1: Filter


# Task 2: Calculate mean with chaining


# Task 3: Count by Sepal.Width

```

```{r exercise-iris-2-hint}
# Task 1: iris_dt[Sepal.Length < 6.7 & Species == "virginica"]
# Task 2: Add [, mean(Petal.Width)] after Task 1
# Task 3: iris_dt[Sepal.Length < 6.7 & Species == "virginica", .N, by = .(Sepal.Width > 3)]
```

```{r exercise-iris-2-solution}
iris_dt <- as.data.table(iris)

# Task 1
iris_dt[Sepal.Length < 6.7 & Species == "virginica"]

# Task 2
iris_dt[Sepal.Length < 6.7 & Species == "virginica"][, mean(Petal.Width)]

# Task 3
iris_dt[Sepal.Length < 6.7 & Species == "virginica", .N, by = .(Sepal.Width > 3)]
```

---

## Modifying Data

<div class="callout callout-concept">
<div class="callout-title">The Walrus Operator: :=</div>

The `:=` operator is data.table's **assignment by reference** operator. It modifies data tables **in-place** without making copies, which is:
- **Fast** - no copying overhead
- **Memory efficient** - works on the original object
- **Permanent** - changes the original data table
</div>

<div class="callout callout-warning">
<div class="callout-title">The := Operator</div>

The `:=` operator modifies data tables **by reference** (in-place) without making copies. This is incredibly fast and memory-efficient, but be careful - it changes the original object!
</div>

### Adding New Columns

### Use := to Add a Column

```{r add-column, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, N := .N]
ddf_dt[1:3]
```

### Add Multiple Columns

<div class="callout callout-tip">
<div class="callout-title">Multiple Assignments</div>

Use `:=` as a **function** to assign multiple columns at once. Wrap assignments in `':='(col1 = val1, col2 = val2)` for efficient batch updates.
</div>

Use `:=` as a function:

```{r add-multiple, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, 
       ':='(N_grp = .N, mean = mean(Height)),
       by = .(Gender, Continent)]
ddf_dt
```

---

### Exercise 3: Add Columns to iris

**Tasks:**
1. Add columns to `iris_dt` for mean and sd of Sepal.Width grouped by Species
2. Use `uniqueN()` to check how many unique mean Sepal Widths there are

```{r exercise-add-cols, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Add mean and sd columns


# Check unique means

```

```{r exercise-add-cols-solution}
iris_dt <- as.data.table(iris)

iris_dt[, ':='(mean_sw = mean(Sepal.Width), 
               sd_sw = sd(Sepal.Width)), 
        by = Species]

uniqueN(iris_dt$mean_sw)
```

---

### Special Symbols: .I and .GRP

<div class="callout callout-info">
<div class="callout-title">More Special Symbols</div>

data.table has several built-in symbols for advanced operations:
- `.I` - row numbers (indices) 
- `.GRP` - unique group number
- `.N` - count of rows in group
- `.SD` - subset of data (all columns except grouping ones)

These make complex operations simple and readable!
</div>

### .I holds row numbers

```{r dot-i, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .(row_id = .I, Gender, Continent)]
```

### .GRP holds unique group number

```{r dot-grp, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .GRP, by = .(Gender, Continent)]
```

---

### Exercise 4: Advanced Operations

**Tasks:**
1. Add column to `iris_dt` with number of observations where Petal.Length < 6.5, grouped by Species
2. Select all rows with Species == "virginica" and rename those Species entries to "new_virginica"

```{r exercise-advanced, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Task 1: Add column with counts


# Task 2: Sub-assign by reference

```

```{r exercise-advanced-solution}
iris_dt <- as.data.table(iris)

# Task 1
iris_dt[Petal.Length < 6.5, N_count := .N, by = Species]

# Task 2
iris_dt[Species == "virginica", Species := "new_virginica"]
iris_dt
```

---

## Advanced Features

<div class="callout callout-tip">
<div class="callout-title">Advanced Techniques</div>

data.table includes powerful features like **keys** (for fast lookups), **.SD** (subset of data), **rolling joins**, and **reshaping** - these are what make it enterprise-grade!
</div>

### Keys for Faster Operations

Set a key for faster manipulation (especially merging)!

```{r keys, exercise=TRUE, exercise.setup="shared-data"}
setkey(ddf_dt, Gender, Continent, Height)
ddf_dt
```

---

### .SD and .SDcols

### .SD selects all columns or subset

```{r sd-demo, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .SD, .SDcols = c(2, 4, 5)]
```

### Apply function to multiple columns

```{r sd-apply, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, lapply(.SD, mean), by = Gender, .SDcols = c(2, 4, 5)]
```

### Select first and last row per group

```{r sd-firstlast, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, .SD[c(1, .N)], by = Gender]
```

---

### Exercise 5: Advanced .SD

Do the following in a single command:
1. Order results by Petal.Width and select first three (smallest) observations
2. Calculate mean of first four columns for those observations

```{r exercise-sd, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Order, select, and calculate means

```

```{r exercise-sd-hint}
# Use setorder() or order(), then .SD with .SDcols
iris_dt[order(Petal.Width)][1:3][, lapply(.SD, mean), .SDcols = 1:4]
```

```{r exercise-sd-solution}
iris_dt <- as.data.table(iris)
iris_dt[order(Petal.Width)][1:3][, lapply(.SD, mean), .SDcols = 1:4]
```

---

### Reshape with dcast & melt

```{r reshape, exercise=TRUE, exercise.setup="shared-data"}
molten <- melt(ddf_dt, id.vars = c("Gender", "Continent"))
dcast(ddf_dt, Gender + Age ~ Continent)
dcast(ddf_dt, Gender + Age ~ Continent, fun.aggregate = mean)
```

---

### Suppress Output with {}

Calculate BMI without saving intermediate steps!

**BMI = Weight / (Height in meters)²**

```{r bmi, exercise=TRUE, exercise.setup="shared-data"}
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    heightinmeterssquared <- heightinmeters^2
    Weight / heightinmeterssquared
}]
ddf_dt
```

### Exercise: How many people are obese (BMI > 25)?

```{r exercise-bmi, exercise=TRUE, exercise.setup="shared-data"}
# Calculate BMI first if needed
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    Weight / (heightinmeters^2)
}]

# Count obese people

```

```{r exercise-bmi-solution}
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    Weight / (heightinmeters^2)
}]
nrow(ddf_dt[BMI > 25])
```

---

### Merging Data Tables

### Define Dummy Data

```{r merge-setup, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt1
dt2
```

### Merge by Variable

```{r merge-demo, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))

merge(dt1, dt2, by = "x")
```

### Other Join Types

```{r merge-types, exercise=TRUE, eval=FALSE}
merge(dt1, dt2, by = "x", all.x = TRUE)  # Left join
merge(dt1, dt2, by = "x", all.y = TRUE)  # Right join
```

### data.table Join Syntax

```{r dt-join, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))

dt1[dt2, on = .(x)]
```

---

### Rolling Joins

### Forward Join

Merge on closest smaller numerical value in dt2:

```{r roll-forward, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))

dt2[dt1, on = .(y), roll = TRUE]
```

### Backward Join

Merge on closest larger numerical value:

```{r roll-backward, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))

dt2[dt1, on = .(y), roll = -4]
```

---

### Overlapping Joins

```{r foverlaps, exercise=TRUE}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt3 <- data.table(min_y = c(0, 10, 15, 20), max_y = c(10, 15, 20, 30))

setkey(dt3, min_y, max_y)
dt1[, `:=`(dt1_y_end = c(13, 25, 10, 22), dt1_y = y)]
setkey(dt1, dt1_y, dt1_y_end)

foverlaps(dt1, dt3, type = "any")
```

---

## Congratulations!

You've completed the **Data.table in R** tutorial!

### What You've Learned:

- data.table syntax: DT[i, j, by]  
- Selecting rows and columns  
- Operations and aggregations by group  
- Special symbols: .N, .I, .GRP, .SD  
- Adding and modifying columns with :=  
- Keys for faster operations  
- Reshaping with melt and dcast  
- Merging and joining data tables  
- Rolling and overlapping joins

### Next Steps:

Continue with **ggplot2** for data visualization!

---

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

### Additional Resources

- [data.table Documentation](https://rdatatable.gitlab.io/data.table/)
- [data.table Cheat Sheet](https://github.com/Rdatatable/data.table/wiki/Getting-started)
- [Advanced data.table](https://rdatatable.gitlab.io/data.table/articles/datatable-intro.html)
