---
title: "Data.table in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()

# data.table datasets
ddf <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
ddf_dt <- as.data.table(ddf)
iris_dt <- as.data.table(iris)

# ggplot2 datasets
tidy <- ddf
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))

# IntroToR variables
myFirstNumber <- 0.1
myFirstVector <- c("some words","p","awesome", "last one")
myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
myvector <- c(15, 16, 17, 18, 20)
```

## Welcome

Welcome to **Data.table in R**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

<div class="callout callout-info">
<div class="callout-title">How to Use This Tutorial</div>

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises
</div>

<div class="callout callout-success">
<div class="callout-title">Why data.table?</div>

- **Extremely fast** - optimized for speed with large datasets  
- **Concise syntax** - less typing, more doing  
- **Powerful** - complex operations in single expressions  
- **Memory efficient** - works by reference, not copies
</div>

<div class="callout callout-success">
<div class="callout-title">What You Will Learn</div>

In this tutorial, you will master data.table for high-performance data manipulation:

1. **Getting Started**  
   - Introduction to data.table  
   - Installing and loading data.table  
   - Converting to data.table  
   - Understanding DT[i, j, by] syntax  
   
2. **Basic Operations**  
   - Selecting rows  
   - Selecting columns  
   - Column selection with variables  
   - Range and exclusion operations  
   
3. **Aggregations and Groups**  
   - Operations on columns  
   - Grouping with by  
   - Multiple statistics by group  
   - Using .N for counting  
   - Grouping by multiple variables  
   
4. **Modifying Data**  
   - The := operator  
   - Adding new columns  
   - Adding multiple columns  
   - Sub-assignment by reference  
   
5. **Advanced Features**  
   - Keys for faster operations  
   - .SD and .SDcols  
   - .I and .GRP special symbols  
   - Reshaping with melt and dcast  
   - Merging and joining data tables  
   - Rolling and overlapping joins  

</div>

---

## Getting Started

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started**  
   → Introduction to data.table  
   - Installing and loading data.table  
   - Converting to data.table  
   - Understanding DT[i, j, by] syntax  

</div>

<div class="callout callout-info">
<div class="callout-title">What is data.table?</div>

`data.table` is a high-performance package for working with tabular data in R. It's **extremely fast** and uses concise syntax: `DT[i, j, by]` where `i` = rows, `j` = columns, `by` = groups.
</div>

### Introduction

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started**  
   → Introduction to data.table  
   - Installing and loading data.table  
   - Converting to data.table  
   - Understanding DT[i, j, by] syntax  

</div>

### Sample Data

We'll use this dataset for examples:

Display the `ddf` dataset by typing its name:

```{r data-preview, exercise=TRUE, exercise.setup="shared-data"}
# Display the ddf dataset


```

```{r data-preview-hint}
ddf
```

```{r data-preview-solution}
ddf
```

And you'll practice on the `iris` dataset:

Display the built-in `iris` dataset:

```{r iris-preview, exercise=TRUE}
# Display the iris dataset


```

```{r iris-preview-hint}
iris
```

```{r iris-preview-solution}
iris
```

---

### Loading the Library

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started**  
   - Introduction to data.table  
   → Installing and loading data.table  
   - Converting to data.table  
   - Understanding DT[i, j, by] syntax  

</div>

### Installation

If you don't have the package, install it (only once):

Use `install.packages()` to install data.table:

```{r install-dt, exercise=TRUE, eval=FALSE}
# Install data.table


```

```{r install-dt-hint, eval=FALSE}
install.packages("data.table")
```

```{r install-dt-solution, eval=FALSE}
install.packages("data.table")
```

### Loading

Load it every session:

Load the data.table library using `library()`:

```{r load-dt, exercise=TRUE}
# Load data.table


```

```{r load-dt-hint}
library(data.table)
```

```{r load-dt-solution}
library(data.table)
```

### Converting to data.table

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started**  
   - Introduction to data.table  
   - Installing and loading data.table  
   → Converting to data.table  
   - Understanding DT[i, j, by] syntax  

</div>

If you already have a data frame:

Convert `ddf` to a data.table using `as.data.table()` and display it:

```{r convert-dt, exercise=TRUE, exercise.setup="shared-data"}
# Convert ddf to data.table


```

```{r convert-dt-hint}
ddf_dt <- as.data.table(ddf)
ddf_dt
```

```{r convert-dt-solution}
ddf_dt <- as.data.table(ddf)
ddf_dt
```

---

### data.table Syntax

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started**  
   - Introduction to data.table  
   - Installing and loading data.table  
   - Converting to data.table  
   → Understanding DT[i, j, by] syntax  

</div>

### All You Need to Know:

### **DT[i, j, by]**

- **i:** select those rows
- **j:** do this to them
- **by:** do it per groups

---

## Basic Operations

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Basic Operations**  
   → Selecting rows  
   - Selecting columns  
   - Column selection with variables  
   - Range and exclusion operations  

</div>

<div class="callout callout-concept">
<div class="callout-title">The DT[i, j, by] Syntax</div>

**The core syntax:** `DT[i, j, by]`
- **i** = which rows (row filter/subset)
- **j** = what to do with columns (select, compute, modify)
- **by** = group by (aggregation groups)

This unified syntax makes data.table incredibly powerful and consistent!
</div>

### Selecting Rows

### Example: Height > 180

How many people are taller than 180?

Filter `ddf_dt` for rows where `Height > 180`, then count how many there are using `nrow()`:

```{r select-rows, exercise=TRUE, exercise.setup="shared-data"}
# Select rows where Height > 180 and count them


```

```{r select-rows-hint}
ddf_dt[Height > 180]
nrow(ddf_dt[Height > 180])
```

```{r select-rows-solution}
ddf_dt[Height > 180]
nrow(ddf_dt[Height > 180])
```

<div class="callout callout-tip">
<div class="callout-title">Familiar Syntax!</div>

You select rows similar to vector subsetting - no need for `ddf_dt$Height`!

data.table knows to look for column names within the data.table.
</div>

---

### Selecting Columns

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Basic Operations**  
   - Selecting rows  
   → Selecting columns  
   - Column selection with variables  
   - Range and exclusion operations  

</div>

### Show Age and Continent of First 5 People

<div class="callout callout-concept">
<div class="callout-title">The Magic Dot: .()</div>

In data.table use **`.`** instead of `c()`

`.()` is shorthand for `list()` - it creates a list of columns to select.
</div>

In data.table use `.` instead of `c()` (`.` is short for `list()`):

Select the first 5 rows and only the `Age` and `Continent` columns using `.()` syntax:

```{r select-cols, exercise=TRUE, exercise.setup="shared-data"}
# Select Age and Continent for first 5 rows


```

```{r select-cols-hint}
ddf_dt[1:5, .(Age, Continent)]
```

```{r select-cols-solution}
ddf_dt[1:5, .(Age, Continent)]
```

### Using Column Names as Variables

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Basic Operations**  
   - Selecting rows  
   - Selecting columns  
   → Column selection with variables  
   - Range and exclusion operations  

</div>

<div class="callout callout-tip">
<div class="callout-title">Two Dots = Variable!</div>

Use the `..` prefix to reference column names stored in variables. The two dots tell data.table: "look up one level in the calling environment".
</div>

Store column names in a variable `cnames` and use `..` prefix to select those columns for the first 5 rows:

```{r select-cols-vars, exercise=TRUE, exercise.setup="shared-data"}
# Use variable with .. prefix to select columns


```

```{r select-cols-vars-hint}
cnames <- c("Age", "Continent")
ddf_dt[1:5, ..cnames]
```

```{r select-cols-vars-solution}
cnames <- c("Age", "Continent")
ddf_dt[1:5, ..cnames]
```

---

### More Column Selection

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Basic Operations**  
   - Selecting rows  
   - Selecting columns  
   - Column selection with variables  
   → Range and exclusion operations  

</div>

### Range of Columns

Select a range of columns from `Age` to `Continent` for the first 3 rows:

```{r select-range, exercise=TRUE, exercise.setup="shared-data"}
# Select range of columns


```

```{r select-range-hint}
ddf_dt[1:3, Age:Continent]
```

```{r select-range-solution}
ddf_dt[1:3, Age:Continent]
```

### Exclude Columns

Exclude the columns from `Age` to `Continent` using `-()` for the first 3 rows:

```{r select-exclude, exercise=TRUE, exercise.setup="shared-data"}
# Exclude a range of columns


```

```{r select-exclude-hint}
ddf_dt[1:3, -(Age:Continent)]
```

```{r select-exclude-solution}
ddf_dt[1:3, -(Age:Continent)]
```

### Negate Selection

Negate a selection using `!()` to exclude columns from `Age` to `Gender` for the first 3 rows:

```{r select-negate, exercise=TRUE, exercise.setup="shared-data"}
# Negate column selection


```

```{r select-negate-hint}
ddf_dt[1:3, !(Age:Gender)]
```

```{r select-negate-solution}
ddf_dt[1:3, !(Age:Gender)]
```

---

### Exercise 1: iris Basics

Practice with the `iris` dataset!

**Tasks:**
1. Convert `iris` to data.table - call it `iris_dt`
2. Select all rows with Sepal.Length < 6.7
3. Select as before, but show only Sepal.Length and Species columns

```{r exercise-iris-1, exercise=TRUE}
# Convert iris to data.table
iris_dt <- as.data.table(iris)

# Select rows with Sepal.Length < 6.7


# Show only specific columns

```

```{r exercise-iris-1-hint}
# Step 2: iris_dt[Sepal.Length < 6.7]
# Step 3: iris_dt[Sepal.Length < 6.7, .(Sepal.Length, Species)]
```

```{r exercise-iris-1-solution}
iris_dt <- as.data.table(iris)
iris_dt[Sepal.Length < 6.7]
iris_dt[Sepal.Length < 6.7, .(Sepal.Length, Species)]
```

---

## Aggregations and Groups

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Aggregations and Groups**  
   → Operations on columns  
   - Grouping with by  
   - Multiple statistics by group  
   - Using .N for counting  
   - Grouping by multiple variables  

</div>

<div class="callout callout-success">
<div class="callout-title">Group Operations</div>

The `by` parameter is where data.table truly shines. Compute statistics **by groups** with lightning-fast performance - essential for data analysis!
</div>

### Operations on Columns

### Calculate Mean

Calculate the mean of `Height` from `ddf_dt`:

```{r calc-mean, exercise=TRUE, exercise.setup="shared-data"}
# Calculate mean Height


```

```{r calc-mean-hint}
ddf_dt[, mean(Height)]
```

```{r calc-mean-solution}
ddf_dt[, mean(Height)]
```

### Mean by Groups

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Aggregations and Groups**  
   - Operations on columns  
   → Grouping with by  
   - Multiple statistics by group  
   - Using .N for counting  
   - Grouping by multiple variables  

</div>

Calculate the mean of `Height` grouped by `Gender`:

```{r mean-by-group, exercise=TRUE, exercise.setup="shared-data"}
# Mean Height by Gender


```

```{r mean-by-group-hint}
ddf_dt[, mean(Height), by = Gender]
```

```{r mean-by-group-solution}
ddf_dt[, mean(Height), by = Gender]
```

---

### Multiple Statistics by Group

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Aggregations and Groups**  
   - Operations on columns  
   - Grouping with by  
   → Multiple statistics by group  
   - Using .N for counting  
   - Grouping by multiple variables  

</div>

Calculate mean, sd, max, min for Height, and count per Gender, ordered by mean:

Calculate mean, sd, min, max of Height and count (`.N`) by Gender, then order by mean:

```{r multiple-stats, exercise=TRUE, exercise.setup="shared-data"}
# Multiple statistics by Gender, ordered by mean


```

```{r multiple-stats-hint}
ddf_dt[, 
       .(mean = mean(Height),
         sd = sd(Height),
         min_x = min(Height),
         max_x = max(Height),
         N = .N),
       by = Gender][order(mean)]
```

```{r multiple-stats-solution}
ddf_dt[, 
       .(mean = mean(Height),
         sd = sd(Height),
         min_x = min(Height),
         max_x = max(Height),
         N = .N),
       by = Gender][order(mean)]
```

---

### Special Symbols: .N

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Aggregations and Groups**  
   - Operations on columns  
   - Grouping with by  
   - Multiple statistics by group  
   → Using .N for counting  
   - Grouping by multiple variables  

</div>

<div class="callout callout-concept">
<div class="callout-title">Special Symbol: .N</div>

`.N` gives the **number of observations** in the current group. It's incredibly useful for counting and is one of data.table's most frequently used special symbols.
</div>

### .N gives number of observations

Use `.N` to count rows in your data or within groups:

Use `.N` to count total rows and count rows by `Gender`:

```{r dot-n, exercise=TRUE, exercise.setup="shared-data"}
# Count total rows and by Gender


```

```{r dot-n-hint}
ddf_dt[, .N]
ddf_dt[, .N, by = Gender]
```

```{r dot-n-solution}
ddf_dt[, .N]
ddf_dt[, .N, by = Gender]
```

### Group by Multiple Variables

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Aggregations and Groups**  
   - Operations on columns  
   - Grouping with by  
   - Multiple statistics by group  
   - Using .N for counting  
   → Grouping by multiple variables  

</div>

<div class="callout callout-tip">
<div class="callout-title">Dynamic Grouping</div>

You can group by **expressions**, not just column names! This allows for flexible on-the-fly categorization like `Height > 180` without creating new columns.
</div>

Count rows grouped by `Gender` and whether `Height > 180`:

```{r multiple-groups, exercise=TRUE, exercise.setup="shared-data"}
# Count by Gender and Height > 180


```

```{r multiple-groups-hint}
ddf_dt[, .N, by = .(Gender, Height > 180)]
```

```{r multiple-groups-solution}
ddf_dt[, .N, by = .(Gender, Height > 180)]
```

### Exercise: Count by Group

How many men and women are there from different continents?

```{r exercise-count, exercise=TRUE, exercise.setup="shared-data"}
# Count by Gender and Continent

```

```{r exercise-count-solution}
ddf_dt[, .N, by = .(Gender, Continent)]
```

---

### Exercise 2: iris Operations

**Tasks:**
1. Select all rows where Sepal.Length < 6.7 AND Species == "virginica"
2. For those, use chaining `[][]` to calculate mean Petal.Width
3. How many flowers have Sepal.Width > 3 and how many have Sepal.Width < 3?

```{r exercise-iris-2, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Task 1: Filter


# Task 2: Calculate mean with chaining


# Task 3: Count by Sepal.Width

```

```{r exercise-iris-2-hint}
# Task 1: iris_dt[Sepal.Length < 6.7 & Species == "virginica"]
# Task 2: Add [, mean(Petal.Width)] after Task 1
# Task 3: iris_dt[Sepal.Length < 6.7 & Species == "virginica", .N, by = .(Sepal.Width > 3)]
```

```{r exercise-iris-2-solution}
iris_dt <- as.data.table(iris)

# Task 1
iris_dt[Sepal.Length < 6.7 & Species == "virginica"]

# Task 2
iris_dt[Sepal.Length < 6.7 & Species == "virginica"][, mean(Petal.Width)]

# Task 3
iris_dt[Sepal.Length < 6.7 & Species == "virginica", .N, by = .(Sepal.Width > 3)]
```

---

## Modifying Data

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Modifying Data**  
   → The := operator  
   - Adding new columns  
   - Adding multiple columns  
   - Sub-assignment by reference  

</div>

<div class="callout callout-concept">
<div class="callout-title">The Walrus Operator: :=</div>

The `:=` operator is data.table's **assignment by reference** operator. It modifies data tables **in-place** without making copies, which is:
- **Fast** - no copying overhead
- **Memory efficient** - works on the original object
- **Permanent** - changes the original data table
</div>

<div class="callout callout-warning">
<div class="callout-title">The := Operator</div>

The `:=` operator modifies data tables **by reference** (in-place) without making copies. This is incredibly fast and memory-efficient, but be careful - it changes the original object!
</div>

### Adding New Columns

### Use := to Add a Column

Add a column `N` with the total count using `:=`:

```{r add-column, exercise=TRUE, exercise.setup="shared-data"}
# Add N column with total count


```

```{r add-column-hint}
ddf_dt[, N := .N]
ddf_dt[1:3]
```

```{r add-column-solution}
ddf_dt[, N := .N]
ddf_dt[1:3]
```

### Add Multiple Columns

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Modifying Data**  
   - The := operator  
   → Adding new columns  
   - Adding multiple columns  
   - Sub-assignment by reference  

</div>

<div class="callout callout-tip">
<div class="callout-title">Multiple Assignments</div>

Use `:=` as a **function** to assign multiple columns at once. Wrap assignments in `':='(col1 = val1, col2 = val2)` for efficient batch updates.
</div>

Use `:=` as a function:

Add multiple columns (`N_grp` and `mean`) using `:=` as a function, grouped by Gender and Continent:

```{r add-multiple, exercise=TRUE, exercise.setup="shared-data"}
# Add N_grp and mean columns by Gender and Continent


```

```{r add-multiple-hint}
ddf_dt[, 
       ':='(N_grp = .N, mean = mean(Height)),
       by = .(Gender, Continent)]
ddf_dt
```

```{r add-multiple-solution}
ddf_dt[, 
       ':='(N_grp = .N, mean = mean(Height)),
       by = .(Gender, Continent)]
ddf_dt
```

---

### Exercise 3: Add Columns to iris

**Tasks:**
1. Add columns to `iris_dt` for mean and sd of Sepal.Width grouped by Species
2. Use `uniqueN()` to check how many unique mean Sepal Widths there are

```{r exercise-add-cols, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Add mean and sd columns


# Check unique means

```

```{r exercise-add-cols-solution}
iris_dt <- as.data.table(iris)

iris_dt[, ':='(mean_sw = mean(Sepal.Width), 
               sd_sw = sd(Sepal.Width)), 
        by = Species]

uniqueN(iris_dt$mean_sw)
```

---

### Special Symbols: .I and .GRP

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Modifying Data**  
   - The := operator  
   - Adding new columns  
   - Adding multiple columns  
   → Sub-assignment by reference  

</div>

<div class="callout callout-info">
<div class="callout-title">More Special Symbols</div>

data.table has several built-in symbols for advanced operations:
- `.I` - row numbers (indices) 
- `.GRP` - unique group number
- `.N` - count of rows in group
- `.SD` - subset of data (all columns except grouping ones)

These make complex operations simple and readable!
</div>

### .I holds row numbers

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Advanced Features**  
   - Keys for faster operations  
   - .SD and .SDcols  
   → .I and .GRP special symbols  
   - Reshaping with melt and dcast  
   - Merging and joining data tables  
   - Rolling and overlapping joins  

</div>

Use `.I` to show row numbers alongside `Gender` and `Continent`:

```{r dot-i, exercise=TRUE, exercise.setup="shared-data"}
# Show row numbers with .I


```

```{r dot-i-hint}
ddf_dt[, .(row_id = .I, Gender, Continent)]
```

```{r dot-i-solution}
ddf_dt[, .(row_id = .I, Gender, Continent)]
```

### .GRP holds unique group number

Use `.GRP` to get unique group numbers for each Gender-Continent combination:

```{r dot-grp, exercise=TRUE, exercise.setup="shared-data"}
# Show group numbers with .GRP


```

```{r dot-grp-hint}
ddf_dt[, .GRP, by = .(Gender, Continent)]
```

```{r dot-grp-solution}
ddf_dt[, .GRP, by = .(Gender, Continent)]
```

---

### Exercise 4: Advanced Operations

**Tasks:**
1. Add column to `iris_dt` with number of observations where Petal.Length < 6.5, grouped by Species
2. Select all rows with Species == "virginica" and rename those Species entries to "new_virginica"

```{r exercise-advanced, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Task 1: Add column with counts


# Task 2: Sub-assign by reference

```

```{r exercise-advanced-solution}
iris_dt <- as.data.table(iris)

# Task 1
iris_dt[Petal.Length < 6.5, N_count := .N, by = Species]

# Task 2
iris_dt[Species == "virginica", Species := "new_virginica"]
iris_dt
```

---

## Advanced Features

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Advanced Features**  
   → Keys for faster operations  
   - .SD and .SDcols  
   - .I and .GRP special symbols  
   - Reshaping with melt and dcast  
   - Merging and joining data tables  
   - Rolling and overlapping joins  

</div>

<div class="callout callout-tip">
<div class="callout-title">Advanced Techniques</div>

data.table includes powerful features like **keys** (for fast lookups), **.SD** (subset of data), **rolling joins**, and **reshaping** - these are what make it enterprise-grade!
</div>

### Keys for Faster Operations

Set a key for faster manipulation (especially merging)!

Set keys on `Gender`, `Continent`, and `Height` using `setkey()` and display the result:

```{r keys, exercise=TRUE, exercise.setup="shared-data"}
# Set keys for faster operations


```

```{r keys-hint}
setkey(ddf_dt, Gender, Continent, Height)
ddf_dt
```

```{r keys-solution}
setkey(ddf_dt, Gender, Continent, Height)
ddf_dt
```

---

### .SD and .SDcols

### .SD selects all columns or subset

Use `.SD` with `.SDcols` to select columns 2, 4, and 5:

```{r sd-demo, exercise=TRUE, exercise.setup="shared-data"}
# Select specific columns using .SD and .SDcols


```

```{r sd-demo-hint}
ddf_dt[, .SD, .SDcols = c(2, 4, 5)]
```

```{r sd-demo-solution}
ddf_dt[, .SD, .SDcols = c(2, 4, 5)]
```

### Apply function to multiple columns

Apply `mean` to columns 2, 4, and 5 grouped by `Gender` using `lapply(.SD, mean)`:

```{r sd-apply, exercise=TRUE, exercise.setup="shared-data"}
# Apply mean to multiple columns by Gender


```

```{r sd-apply-hint}
ddf_dt[, lapply(.SD, mean), by = Gender, .SDcols = c(2, 4, 5)]
```

```{r sd-apply-solution}
ddf_dt[, lapply(.SD, mean), by = Gender, .SDcols = c(2, 4, 5)]
```

### Select first and last row per group

Select first and last row per `Gender` group using `.SD[c(1, .N)]`:

```{r sd-firstlast, exercise=TRUE, exercise.setup="shared-data"}
# Get first and last row per group


```

```{r sd-firstlast-hint}
ddf_dt[, .SD[c(1, .N)], by = Gender]
```

```{r sd-firstlast-solution}
ddf_dt[, .SD[c(1, .N)], by = Gender]
```

---

### Exercise 5: Advanced .SD

Do the following in a single command:
1. Order results by Petal.Width and select first three (smallest) observations
2. Calculate mean of first four columns for those observations

```{r exercise-sd, exercise=TRUE}
iris_dt <- as.data.table(iris)

# Order, select, and calculate means

```

```{r exercise-sd-hint}
# Use setorder() or order(), then .SD with .SDcols
iris_dt[order(Petal.Width)][1:3][, lapply(.SD, mean), .SDcols = 1:4]
```

```{r exercise-sd-solution}
iris_dt <- as.data.table(iris)
iris_dt[order(Petal.Width)][1:3][, lapply(.SD, mean), .SDcols = 1:4]
```

---

### Reshape with dcast & melt

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Advanced Features**  
   - Keys for faster operations  
   - .SD and .SDcols  
   - .I and .GRP special symbols  
   → Reshaping with melt and dcast  
   - Merging and joining data tables  
   - Rolling and overlapping joins  

</div>

Use `melt()` to reshape `ddf_dt` by Gender and Continent, then use `dcast()` to pivot it back:

```{r reshape, exercise=TRUE, exercise.setup="shared-data"}
# Reshape data with melt and dcast


```

```{r reshape-hint}
molten <- melt(ddf_dt, id.vars = c("Gender", "Continent"))
dcast(ddf_dt, Gender + Age ~ Continent)
dcast(ddf_dt, Gender + Age ~ Continent, fun.aggregate = mean)
```

```{r reshape-solution}
molten <- melt(ddf_dt, id.vars = c("Gender", "Continent"))
dcast(ddf_dt, Gender + Age ~ Continent)
dcast(ddf_dt, Gender + Age ~ Continent, fun.aggregate = mean)
```

---

### Suppress Output with {}

Calculate BMI without saving intermediate steps!

**BMI = Weight / (Height in meters)²**

Calculate BMI (`Weight / (Height/100)^2`) using `{}` to suppress intermediate output, and add it as a column:

```{r bmi, exercise=TRUE, exercise.setup="shared-data"}
# Calculate and add BMI column


```

```{r bmi-hint}
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    heightinmeterssquared <- heightinmeters^2
    Weight / heightinmeterssquared
}]
ddf_dt
```

```{r bmi-solution}
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    heightinmeterssquared <- heightinmeters^2
    Weight / heightinmeterssquared
}]
ddf_dt
```

### Exercise: How many people are obese (BMI > 25)?

```{r exercise-bmi, exercise=TRUE, exercise.setup="shared-data"}
# Calculate BMI first if needed
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    Weight / (heightinmeters^2)
}]

# Count obese people

```

```{r exercise-bmi-solution}
ddf_dt[, BMI := {
    heightinmeters <- Height / 100
    Weight / (heightinmeters^2)
}]
nrow(ddf_dt[BMI > 25])
```

---

### Merging Data Tables

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Advanced Features**  
   - Keys for faster operations  
   - .SD and .SDcols  
   - .I and .GRP special symbols  
   - Reshaping with melt and dcast  
   → Merging and joining data tables  
   - Rolling and overlapping joins  

</div>

### Define Dummy Data

Create two data.tables `dt1` and `dt2` and display them:

```{r merge-setup, exercise=TRUE}
# Create dt1 and dt2


```

```{r merge-setup-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt1
dt2
```

```{r merge-setup-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt1
dt2
```

### Merge by Variable

Merge `dt1` and `dt2` by the `x` column using `merge()`:

```{r merge-demo, exercise=TRUE}
# Create the data and merge by x


```

```{r merge-demo-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
merge(dt1, dt2, by = "x")
```

```{r merge-demo-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
merge(dt1, dt2, by = "x")
```

### Other Join Types

Try a left join (`all.x = TRUE`) and a right join (`all.y = TRUE`):

```{r merge-types, exercise=TRUE, eval=FALSE}
# Left and right joins


```

```{r merge-types-hint, eval=FALSE}
merge(dt1, dt2, by = "x", all.x = TRUE)  # Left join
merge(dt1, dt2, by = "x", all.y = TRUE)  # Right join
```

```{r merge-types-solution, eval=FALSE}
merge(dt1, dt2, by = "x", all.x = TRUE)  # Left join
merge(dt1, dt2, by = "x", all.y = TRUE)  # Right join
```

### data.table Join Syntax

Use data.table's join syntax `dt1[dt2, on = .(x)]` to join the two tables:

```{r dt-join, exercise=TRUE}
# Join using data.table syntax


```

```{r dt-join-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt1[dt2, on = .(x)]
```

```{r dt-join-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt1[dt2, on = .(x)]
```

---

### Rolling Joins

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Advanced Features**  
   - Keys for faster operations  
   - .SD and .SDcols  
   - .I and .GRP special symbols  
   - Reshaping with melt and dcast  
   - Merging and joining data tables  
   → Rolling and overlapping joins  

</div>

### Forward Join

Merge on closest smaller numerical value in dt2:

Use a forward rolling join with `roll = TRUE` to match on the `y` column:

```{r roll-forward, exercise=TRUE}
# Forward rolling join


```

```{r roll-forward-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt2[dt1, on = .(y), roll = TRUE]
```

```{r roll-forward-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt2[dt1, on = .(y), roll = TRUE]
```

### Backward Join

Merge on closest larger numerical value:

Use a backward rolling join with `roll = -4` to match on the closest larger value:

```{r roll-backward, exercise=TRUE}
# Backward rolling join


```

```{r roll-backward-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt2[dt1, on = .(y), roll = -4]
```

```{r roll-backward-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt2 <- data.table(x = c("a", "b", "k"), y = c(10, 15, 20), z = c("one", "two", "three"))
dt2[dt1, on = .(y), roll = -4]
```

---

### Overlapping Joins

Use `foverlaps()` to find overlapping ranges between `dt1` and `dt3`:

```{r foverlaps, exercise=TRUE}
# Overlapping joins with foverlaps


```

```{r foverlaps-hint}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt3 <- data.table(min_y = c(0, 10, 15, 20), max_y = c(10, 15, 20, 30))
setkey(dt3, min_y, max_y)
dt1[, `:=`(dt1_y_end = c(13, 25, 10, 22), dt1_y = y)]
setkey(dt1, dt1_y, dt1_y_end)
foverlaps(dt1, dt3, type = "any")
```

```{r foverlaps-solution}
dt1 <- data.table(x = c("a", "b", "c", "d"), y = c(11.9, 21.4, 5.7, 18))
dt3 <- data.table(min_y = c(0, 10, 15, 20), max_y = c(10, 15, 20, 30))
setkey(dt3, min_y, max_y)
dt1[, `:=`(dt1_y_end = c(13, 25, 10, 22), dt1_y = y)]
setkey(dt1, dt1_y, dt1_y_end)
foverlaps(dt1, dt3, type = "any")
```

---

## Congratulations!

You've completed the **Data.table in R** tutorial!

### What You've Learned:

- data.table syntax: DT[i, j, by]  
- Selecting rows and columns  
- Operations and aggregations by group  
- Special symbols: .N, .I, .GRP, .SD  
- Adding and modifying columns with :=  
- Keys for faster operations  
- Reshaping with melt and dcast  
- Merging and joining data tables  
- Rolling and overlapping joins

### Next Steps:

Continue with **ggplot2** for data visualization!

---

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

## Please Give Feedback!

<div class="callout callout-success">
<div class="callout-title">💬 Your Feedback Matters</div>

Help us improve this tutorial! Share your experience and suggestions.
</div>

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/B7Wy7G29wQTuEKjU9" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```

---

### Additional Resources

- [data.table Documentation](https://rdatatable.gitlab.io/data.table/)
- [data.table Cheat Sheet](https://github.com/Rdatatable/data.table/wiki/Getting-started)
- [Advanced data.table](https://rdatatable.gitlab.io/data.table/articles/datatable-intro.html)
