---  
title: "Data Analysis with data.table and Categorical Tests"  
output:   
  learnr::tutorial:  
    progressive: true  
    allow_skip: true  
    css: "../assets/kuzman_theme.css"  
runtime: shiny_prerendered  
---  

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
library(ggplot2)
library(openxlsx)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()

# Main datasets
tidy <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
iris_dt <- as.data.table(iris)
mtcars_dt <- as.data.table(mtcars)

# Create additional datasets for statistical tests
set.seed(123)

# Dataset for paired t-test: before/after weight measurements
weight_data <- data.table(
  participant = 1:30,
  before = rnorm(30, mean = 75, sd = 10),
  after = rnorm(30, mean = 72, sd = 10) - runif(30, 0, 5)
)
weight_data[, difference := after - before]

# Prepare long format for plotting
weight_long <- melt(weight_data[, .(participant, before, after)],
                    id.vars = "participant",
                    variable.name = "time",
                    value.name = "weight")

# Dataset for ANOVA: plant growth with 3 fertilizer types
growth_data <- data.table(
  fertilizer = rep(c("Control", "TypeA", "TypeB"), each = 20),
  growth_cm = c(
    rnorm(20, mean = 15, sd = 3),
    rnorm(20, mean = 18, sd = 3),
    rnorm(20, mean = 21, sd = 3)
  )
)

# Dataset for chi-squared: smoking and disease
smoking_disease <- data.table(
  smoking = rep(c("Smoker", "Non-smoker"), times = c(120, 180)),
  disease = c(
    rep(c("Disease", "Healthy"), times = c(40, 80)),
    rep(c("Disease", "Healthy"), times = c(30, 150))
  )
)

# Small dataset for Fisher's exact test
small_study <- data.table(
  treatment = rep(c("Drug", "Placebo"), times = c(10, 10)),
  outcome = c(
    rep(c("Improved", "Not_Improved"), times = c(7, 3)),
    rep(c("Improved", "Not_Improved"), times = c(3, 7))
  )
)

# Dataset for one-sample t-test: blood pressure measurements
bp_data <- data.table(
  patient = 1:25,
  systolic_bp = rnorm(25, mean = 125, sd = 12)
)

# Prepare subset for exercises
iris_subset <- iris_dt[Species %in% c("setosa", "versicolor")]

# Calculate summary statistics for visualization
height_summary <- tidy[, .(
  mean_height = mean(Height),
  sd_height = sd(Height),
  n = .N
), by = Gender]

mean_height <- tidy[, .(mean_height = mean(Height)), by = Gender]

growth_summary <- growth_data[, .(
  mean_growth = mean(growth_cm),
  se_growth = sd(growth_cm) / sqrt(.N),
  n = .N
), by = fertilizer]
growth_summary[, ci_lower := mean_growth - 1.96 * se_growth]
growth_summary[, ci_upper := mean_growth + 1.96 * se_growth]
```

## 1. Welcome  

Welcome to **Data Analysis with data.table and Categorical Tests**  

### 1.1. Author  

**dr.sc. Maja Kuzman** - Kuzman Consulting d.o.o.  

<div class="callout callout-info">  
<div class="callout-title">How to Use This Tutorial</div>  

- Click **"Run Code"** to execute R code  
    - ctrl + enter (windows) or command + enter (macos)  
- Click **"Hint"** if you need help  
- Progress is tracked as you complete exercises  

</div>  

<div class="callout callout-success">  
<div class="callout-title">What You Will Learn</div>  

This tutorial combines data manipulation with categorical statistical analysis:  

1. **Welcome**  
   - Tutorial overview  
   - How to use this tutorial  

2. **Quick Review: Subsetting and Conditions**  
   - Logical subsetting  
   - Comparison operators  
   - Logical operators (AND, OR, NOT)  
   - Practical filtering examples  

3. **Introduction to data.table**  
   - Why data.table?  
   - Converting to data.table  
   - The DT[i, j, by] syntax  

4. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   - Grouping data  

5. **Categorical Data Analysis**  
   - Visualizing categorical data  
   - Chi-squared test  
   - Fisher's exact test  
   - Goodness of fit tests  

**Note:** For parametric tests (t-tests, ANOVA, regression), see the companion tutorial: "Parametric Tests and Regression with data.table"  

</div>  

---  

## 2. Quick Review: Subsetting and Conditions  

Before diving into data.table, let's review key concepts from **IntroToR** that are essential for data manipulation and filtering.  

### 2.1. Logical Subsetting  

You can subset vectors using TRUE/FALSE (logical) values:  

```{r logical-review, exercise=TRUE}
# Create a vector
numbers <- c(10, 25, 30, 15, 40)

# Create a logical vector
selection <- c(TRUE, FALSE, TRUE, FALSE, TRUE)

# Subset: keep only TRUE positions
numbers[selection]  # Returns: 10, 30, 40
```

<div class="callout callout-info">  
**💡 Key Concept:** When you put a logical vector inside `[ ]`, R keeps only elements where the value is TRUE.  
</div>  

---  

### 2.2. Comparison Operators  

**Comparison operators** automatically create TRUE/FALSE vectors:  

```{r comparison-review, exercise=TRUE}
ages <- c(18, 22, 35, 17, 40, 16, 25)

# Check which ages are greater than 18
ages > 18  # Returns: FALSE TRUE TRUE FALSE TRUE FALSE TRUE

# Check which ages equal 25
ages == 25  # Returns: FALSE FALSE FALSE FALSE FALSE FALSE TRUE

# Check which ages are NOT 18
ages != 18  # Returns: FALSE TRUE TRUE TRUE TRUE TRUE TRUE
```

<div class="callout callout-concept">  
<div class="callout-title">Comparison Operators</div>  

- `==` equals  
- `!=` not equals  
- `>` greater than  
- `<` less than  
- `>=` greater than or equal  
- `<=` less than or equal  

</div>  

**Your Turn:** Create a logical vector that checks which numbers are equal to 0.3, then display it:  

```{r comparison-exercise, exercise=TRUE}
numbers <- 0.1*c(1, 2, 3, 4)
numbers
# Check which numbers are equal to 0.3

```

```{r comparison-exercise-hint}
# Use the == operator
numbers == 0.3
```

```{r comparison-exercise-solution}
numbers <- 0.1*c(1, 2, 3, 4)
numbers
numbers == 0.3
# Returns:  WHAT?
```

---

### ⚠️ BE CAREFUL: Comparing Decimal Numbers with ==

<div class="callout callout-danger">
<div class="callout-title">⚠️ DANGER: Floating Point Precision Issues</div>

**Never use `==` to compare decimal numbers!**

Computers store decimal numbers with limited precision, which can cause unexpected results:

```r
0.1 + 0.2 == 0.3  # You'd expect TRUE, but it returns FALSE!
```

**Why?** Computers store numbers in binary (0s and 1s), and some decimals can't be represented exactly. This causes tiny rounding errors that make "equal" numbers appear unequal.

**The Problem in Filtering:**
```r
x <- c(0.1, 0.2, 0.3, 0.4)
x[x == 0.3]  # Might not find 0.3 due to precision errors!
```

</div>

---

### Demonstration: The Problem with == for Decimals

See the floating point precision problem in action:

```{r floating-point-problem, exercise=TRUE}
# Example 1: Simple arithmetic that "should" work
0.1 + 0.2 == 0.3

# Example 2: Why does this happen?
print(0.1 + 0.2, digits = 20)  # Shows the actual stored value
print(0.3, digits = 20)         # Compare to what we expect

# Example 3: Filtering problem
values <- c(0.1, 0.2, 0.3, 0.4, 0.5)
calculated <- c(0.05 + 0.05, 0.1 + 0.1, 0.15 + 0.15, 0.2 + 0.2, 0.25 + 0.25)

# This might not work as expected!
calculated == 0.3
```

---

### Solution: Use Tolerance or Other Comparisons

<div class="callout callout-success">
<div class="callout-title">✅ SAFE ALTERNATIVES</div>

**Instead of `==`, use one of these methods:**

**1. Use `all.equal()` for tolerance-based comparison:**
```r
all.equal(0.1 + 0.2, 0.3)  # Returns TRUE (within tolerance)
```

**2. Use comparison operators (>, <, >=, <=) when possible:**
```r
x[x > 0.29 & x < 0.31]  # Safer than x[x == 0.3]
```

**3. Round numbers first:**
```r
round(0.1 + 0.2, 10) == round(0.3, 10)  # TRUE
```

**4. Use as.numeric(as.character(x)):**
```r
as.numeric(as.character(0.1 + 0.2 ))== 0.3  # TRUE (built-in tolerance)
```

</div>

---

### Demonstration: Safe Comparison Methods

Compare the unsafe `==` with safer alternatives:

```{r safe-comparisons, exercise=TRUE}
# The problem
0.1 + 0.2 == 0.3  # FALSE (unexpected!)

# Solution 1: all.equal() with tolerance
all.equal(0.1 + 0.2, 0.3)  # TRUE (safe!)

# Solution 2: Use range instead of exact match
x <- 0.1 + 0.2
x > 0.29 & x < 0.31  # TRUE (safe!)

# Solution 3: Round first
round(0.1 + 0.2, 10) == round(0.3, 10)  # TRUE (safe!)

# Solution 4: Check absolute difference is small
abs((0.1 + 0.2) - 0.3) < 0.0000001  # TRUE (safe!)


```

<div class="callout callout-tip">
<div class="callout-title">💡 Pro Tip: When to Worry</div>

**You don't need to worry about this when:**
- Comparing integers (whole numbers): `5 == 5` is always safe  
- Using `>`, `<`, `>=`, `<=` with decimals (these work fine!)  

**You DO need to worry when:**
- Using `==` with any decimal calculations  
- Filtering datasets for specific decimal values  
- Working with financial data or precise measurements  

**Remember:** This is a fundamental limitation of how computers store numbers, not a bug in R!

</div>

---  

### 2.3. Conditional Subsetting (Filtering)  

Combine comparisons with subsetting to **filter data based on conditions**:  

```{r conditional-review, exercise=TRUE}
ages <- c(18, 22, 35, 17, 40, 16, 25)

# Get only ages greater than 18 (adults)
adults <- ages[ages > 18]
adults  # Returns: 22, 35, 40, 25

# Get only ages between 18 and 30
young_adults <- ages[ages >= 18 & ages <= 30]
young_adults  # Returns: 18, 22, 25
```

**What's happening:**  
1. `ages > 18` creates a TRUE/FALSE vector  
2. Use it as an index: `ages[ages > 18]`  
3. R keeps only values where the condition is TRUE  

**Your Turn:** From the `heights` vector, get only the heights that are 170 or greater:  

```{r filtering-exercise, exercise=TRUE}
heights <- c(165, 180, 155, 172, 190, 168, 175)

# Filter to keep only heights >= 170

```

```{r filtering-exercise-hint}
# Use heights >= 170 as the condition inside [ ]
heights[heights >= 170]
```

```{r filtering-exercise-solution}
heights <- c(165, 180, 155, 172, 190, 168, 175)
heights[heights >= 170]
# Returns: 180, 172, 190, 175
```

---  

### 2.4. Logical Operators  

Combine multiple conditions using logical operators:  

```{r logical-operators-review, exercise=TRUE}
scores <- c(45, 75, 82, 91, 68, 55, 88)

# AND (&): Both conditions must be TRUE
# Get scores between 70 and 85
scores[scores >= 70 & scores <= 85]  # Returns: 75, 82

# OR (|): At least one condition must be TRUE
# Get scores below 50 or above 90
scores[scores < 50 | scores > 90]  # Returns: 45, 91

# NOT (!): Flip TRUE/FALSE
# Get scores NOT between 60 and 80
scores[!(scores >= 60 & scores <= 80)]  # Returns: 45, 82, 91, 55, 88
```

<div class="callout callout-concept">  
<div class="callout-title">Logical Operators</div>  

- `&` (AND): Both must be TRUE  
- `|` (OR): At least one must be TRUE  
- `!` (NOT): Flips TRUE ↔ FALSE  

**Example:** `x > 10 & x < 20` means "greater than 10 AND less than 20"  

</div>  

**Your Turn:** From the `values` vector, get only values that are either less than 15 OR greater than 25:  

```{r logical-ops-exercise, exercise=TRUE}
values <- c(10, 20, 30, 12, 28, 18, 35)

# Filter for values < 15 OR > 25

```

```{r logical-ops-exercise-hint}
# Use the | (OR) operator
values[values < 15 | values > 25]
```

```{r logical-ops-exercise-solution}
values <- c(10, 20, 30, 12, 28, 18, 35)
values[values < 15 | values > 25]
# Returns: 10, 30, 12, 28, 35
```

---  

<div class="callout callout-tip">  
<div class="callout-title">Why This Matters for data.table</div>  

These concepts are **fundamental to data.table**:  
- The `i` in `DT[i, j, by]` uses these exact filtering techniques  
- You'll write conditions like: `DT[Age > 18 & Gender == "female"]`  
- Understanding logical subsetting makes data.table intuitive!  

</div>  

---  

## 3. Introduction to data.table  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   → Why data.table?  
   - Converting to data.table  
   - The DT[i, j, by] syntax  

</div>  

### 3.1. Why data.table?  

<div class="callout callout-info">  
<div class="callout-title">Why Use data.table?</div>  

**data.table** is a powerful package for working with data in R:  

- **Fast** - extremely efficient with large datasets  
- **Concise** - less typing, clearer code  
- **Powerful** - complex operations made simple  
- **Integrates well** - works seamlessly with ggplot2 and statistical functions  

</div>  

### 3.2. Installing and Loading  

Install data.table once:  

```{r install-dt, echo=TRUE, eval=FALSE}
# Install data.table (only once)
install.packages("data.table")
```

Load it every session:  

```{r load-dt, echo=TRUE, eval=FALSE}
# Load data.table
library(data.table)
library(ggplot2)  # We'll use this for visualization
```

---  

### 3.3. Converting to data.table  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   - Why data.table?  
   → Converting to data.table  
   - The DT[i, j, by] syntax  

</div>  

#### 3.3.1. Reading from Excel Files  

In practice, you'll often receive data in Excel format. Here's the typical workflow:  

```r
# Install openxlsx package if needed
# install.packages("openxlsx")

# Load library
library(openxlsx)
library(data.table)

# Read Excel file
my_data <- read.xlsx("path/to/your/file.xlsx")

# Convert to data.table
my_dt <- as.data.table(my_data)
```

<div class="callout callout-info">  
**💡 Tip:** The `read.xlsx()` function from the `openxlsx` package reads Excel files into a regular data.frame. Use `as.data.table()` to convert it to data.table format for efficient manipulation.  
</div>  

#### 3.3.2. Reading Text Files with fread()  

The `fread()` function (**f**ast **read**) is data.table's powerful function for reading text-based files. It automatically detects file format, separators, and data types, making it very convenient and much faster than base R functions.  

<div class="callout callout-concept">  
<div class="callout-title">🎯 Why use fread()?</div>  

- **Automatic detection**: Detects separators (comma, tab, semicolon, etc.), column types, and headers  
- **Speed**: Much faster than `read.csv()` or `read.table()` for large files  
- **Direct to data.table**: Returns a data.table object directly (no conversion needed)  
- **Flexible**: Works with CSV, TSV, TXT, and other delimited files  
- **URLs supported**: Can read directly from web URLs  

</div>  

```{r fread-example, exercise=TRUE, exercise.setup="shared-data"}
# Example 1: Reading from a URL (like we did with tidy dataset)
# fread automatically creates a data.table
tidy_data <- fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt")
head(tidy_data)

# Example 2: CSV file (comma-separated)
# fread("mydata.csv")

# Example 3: TSV file (tab-separated)
# fread("mydata.tsv")

# Example 4: TXT file with custom separator
# fread("mydata.txt", sep="|")  # pipe-separated

# Example 5: Specifying options explicitly
# fread("mydata.csv", 
#       header = TRUE,        # First row is column names
#       sep = ",",           # Comma separator (auto-detected usually)
#       na.strings = "NA")   # What represents missing values
```

<div class="callout callout-info">  
**💡 Common file formats:**  

- **CSV** (Comma-Separated Values): `mydata.csv` - values separated by commas  
- **TSV** (Tab-Separated Values): `mydata.tsv` or `mydata.tab` - values separated by tabs  
- **TXT** (Text file): `mydata.txt` - can use any delimiter (space, tab, comma, semicolon, pipe, etc.)  

`fread()` automatically detects the separator in most cases, so you usually just need: `fread("filename")`  
</div>  

#### 3.3.3. Converting Built-in Datasets  

You can also convert R's built-in datasets to data.table:  

```{r convert-example, exercise=TRUE}
# Look at regular iris
head(iris)

# Convert to data.table
iris_dt <- as.data.table(iris)
head(iris_dt)
```

---  

### 3.4. The DT[i, j, by] Syntax  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   - Why data.table?  
   - Converting to data.table  
   → The DT[i, j, by] syntax  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">The Magic Formula: DT[i, j, by]</div>  

Think of a data.table like a spreadsheet where you can:  

- **i** = which **rows** (filter/subset)  
- **j** = what to do with **columns** (select/compute)  
- **by** = grouped **by** what (grouping variable)  

**Example:** `DT[Gender == "Female", mean(Height), by = Continent]`  
→ For female rows, calculate mean height, separately for each continent  

</div>  

Let's see this in action:  

```{r syntax-demo, exercise=TRUE, exercise.setup="shared-data"}
# All rows, all columns
iris_dt

# Filter rows: only setosa
iris_dt[Species == "setosa"]

# Select columns
iris_dt[, .(Species, Sepal.Length)]

# Compute: mean sepal length
iris_dt[, mean(Sepal.Length)]

# Group by: mean sepal length per species
iris_dt[, mean(Sepal.Length), by = Species]
```

---  

## 4. Basic Data Operations  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   → Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   - Grouping data  

</div>  

### 4.1. Selecting Rows (Filtering)  

**About the `tidy` dataset:**  
This dataset contains information about people including their Gender ("male" or "female"), Height (in cm), Weight (in kg), Age (in years), and Continent.  

Use logical conditions in the **i** position to filter rows:  

**Example 1: Filter by a single condition**  

```{r filter-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Select only females
tidy[Gender == "female"]
```

**Example 2: Filter by numeric condition**  

```{r filter-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Select people taller than 180 cm
tidy[Height > 180]
```

**Example 3: Combine multiple conditions**  

```{r filter-demo3, exercise=TRUE, exercise.setup="shared-data"}
# Select females taller than 170 cm
tidy[Gender == "female" & Height > 170]
```

**Your Turn:** From the `iris_dt` data, select only rows where `Sepal.Length > 6`:  

```{r filter-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Filter iris for Sepal.Length > 6


```

```{r filter-exercise-hint}
iris_dt[Sepal.Length > 6]
```

```{r filter-exercise-solution}
iris_dt[Sepal.Length > 6]
```

---  

### 4.2. Selecting Columns  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   → Selecting columns  
   - Computing summaries  
   - Grouping data  

</div>  

Use `.()` or `list()` in the **j** position to select columns:  

**Example 1: Select one column**  

```{r select-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Select one column (returns a vector)
tidy[, Gender]
```

**Example 2: Select multiple columns**  

```{r select-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Select multiple columns (returns a data.table)
# Using .() which is shorthand for list()
tidy[, .(Gender, Age, Height)]
```

**Example 3: Alternative syntax**  

```{r select-demo3, exercise=TRUE, exercise.setup="shared-data"}
# Can also use list() directly
tidy[, list(Gender, Height)]
```

**Your Turn:** Select the `Species` and `Petal.Length` columns from `iris_dt`:  

```{r select-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Select Species and Petal.Length


```

```{r select-exercise-hint}
iris_dt[, .(Species, Petal.Length)]
```

```{r select-exercise-solution}
iris_dt[, .(Species, Petal.Length)]
```

---  

### 4.3. Computing Summaries  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   → Computing summaries  
   - Grouping data  

</div>  

In the **j** position, you can do more than just select existing columns - you can also **create new columns with computations**:  

**Creating New Columns:**  

```{r compute-new-cols, exercise=TRUE, exercise.setup="shared-data"}
# Create a new column: ratio of Height to Weight
tidy[, .(Gender, Height, Weight, Height_Weight_Ratio = Height / Weight)]

# Multiple new computed columns
tidy[, .(
  Gender,
  Height,
  Weight,
  Height_m = Height / 100,          # Convert cm to meters
  BMI = Weight / (Height/100)^2     # Calculate BMI
)]
```

<div class="callout callout-info">  
**💡 Key Concept:** Inside `.()` in the j position:  
- Column names alone select existing columns: `.(Gender, Height)`  
- `name = expression` creates new computed columns: `.(BMI = Weight / (Height/100)^2)`  
- You can mix both: select existing and create new columns together!  
</div>  

---  

**Computing Summary Statistics:**  

Now for something powerful: instead of just listing or modifying columns, we can **collapse the entire dataset into summary statistics**!  

<div class="callout callout-concept">  
<div class="callout-title">🎯 From Data to Summaries</div>  

When you use **functions** like `mean()`, `median()`, `sd()` in the j position, data.table:  
- **Aggregates** all rows into a single summary value  
- Returns the computed statistic(s), not the original data  
- Can compute **multiple summaries at once** inside `.()`   

This is fundamentally different from selecting or computing columns!  

</div>  

```{r compute-demo, exercise=TRUE, exercise.setup="shared-data"}
# Single summary: Mean height (returns one number)
tidy[, mean(Height)]

# Multiple summaries at once (returns a data.table with one row)
tidy[, .(
  mean_height = mean(Height),
  median_height = median(Height),
  sd_height = sd(Height),
  min_height = min(Height),
  max_height = max(Height)
)]

# Special variable .N = count number of rows
tidy[, .N]
```

<div class="callout callout-tip">  
**💡 Notice:**   
- **One summary:** `DT[, mean(col)]` returns a single value  
- **Multiple summaries:** `DT[, .(stat1 = mean(col), stat2 = sd(col))]` returns a data.table with one row containing all statistics  
- **Combining makes sense:** Get mean, median, sd, min, max all at once!  
</div>  

**Your Turn:** Calculate the mean and standard deviation of `Sepal.Length` from `iris_dt`:  

```{r compute-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Calculate mean and sd of Sepal.Length


```

```{r compute-exercise-hint}
iris_dt[, .(mean_sepal = mean(Sepal.Length), sd_sepal = sd(Sepal.Length))]
```

```{r compute-exercise-solution}
iris_dt[, .(
  mean_sepal = mean(Sepal.Length),
  sd_sepal = sd(Sepal.Length)
)]
```

---  

### 4.4. Grouping Data  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   → Grouping data  

</div>  

Use **by** to compute statistics for each group:  

```{r group-demo, exercise=TRUE, exercise.setup="shared-data"}
# Mean height by gender
tidy[, mean(Height), by = Gender]

# Multiple statistics by group
tidy[, .(
  mean_height = mean(Height),
  count = .N
), by = Gender]

# Group by multiple variables
tidy[, mean(Height), by = .(Gender, Continent)]
```

**Your Turn:** Calculate the mean `Petal.Length` for each `Species` in `iris_dt`:  

```{r group-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Mean Petal.Length by Species


```

```{r group-exercise-hint}
iris_dt[, mean(Petal.Length), by = Species]
```

```{r group-exercise-solution}
iris_dt[, .(mean_petal_length = mean(Petal.Length)), by = Species]
```

---  

### 4.5. Practice: Combining Operations  

Now combine filtering, selecting, and grouping:  

**Challenge:** From the `tidy` data:  
1. Filter for people with Height > 165  
2. Calculate mean Age and count  
3. Group by Gender  

```{r combine-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Your code here


```

```{r combine-exercise-hint}
# Remember: DT[i, j, by]
# i = Height > 165
# j = calculate mean Age and count
# by = Gender
```

```{r combine-exercise-solution}
tidy[Height > 165, .(mean_age = mean(Age), count = .N), by = Gender]
```

---  

## 5. Categorical Data Analysis  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   → Visualizing categorical data  
   - Chi-squared test  
   - Fisher's exact test  

</div>  

### 5.1. Visualizing Categorical Data  

**About the `smoking_disease` dataset:**  
This dataset contains data from 300 people, including their smoking status (Smoker or Non-smoker) and disease status (Disease or Healthy). We want to know if smoking is associated with disease.  

Before testing, always visualize your data!  

**Example 1: Single variable bar chart**  

```{r cat-viz-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Bar chart showing count of smokers vs non-smokers
ggplot(smoking_disease, aes(smoking, fill = smoking)) +
    geom_bar() +
    theme_light() +
    labs(title = "Smoking Status Distribution", y = "Count")
```

**Example 2: Stacked proportional bar chart**  

```{r cat-viz-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Stacked bar chart showing disease within each smoking group
ggplot(smoking_disease, aes(smoking, fill = disease)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(
        title = "Disease Prevalence by Smoking Status",
        y = "Proportion"
    )
```

**Your Turn:** Calculate the number of smokers/nonsmokers with/without disease using data.table:  

```{r cat-viz-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Calculate counts using data.table


```

```{r cat-viz-exercise-hint}
# Use .N to count rows, grouped by smoking and disease
smoking_disease[, .N, by = .(smoking, disease)]
```

```{r cat-viz-exercise-solution}
# Count by smoking status and disease status
smoking_disease[, .N, by = .(smoking, disease)]

# Alternative: Create a contingency table
table(smoking_disease[, smoking], smoking_disease[, disease])
```

---  

### 5.2. Chi-Squared Test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   - Visualizing categorical data  
   → Chi-squared test  
   - Fisher's exact test  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Chi-Squared Test</div>  

**Purpose:** Test if two categorical variables are independent  

**Question:** "Is there a relationship between smoking and disease?"  

**Requirements:**  
- Categorical variables (e.g., yes/no, groups)  
- Expected cell counts ≥ 5  
- Independent observations  

**Hypotheses:**  
- H₀: Variables are independent (no relationship)  
- H₁: Variables are related  

</div>  

First, create a contingency table:  

```{r chi-table, exercise=TRUE, exercise.setup="shared-data"}
# Count observations using .N
smoking_disease[, .N, by = .(smoking, disease)]

# Create contingency table for the test
smoking_table <- table(smoking_disease[,smoking], smoking_disease[,disease])
smoking_table
```

Now perform the chi-squared test:  

```{r chi-test, exercise=TRUE, exercise.setup="shared-data"}
# Chi-squared test
test_result <- chisq.test(smoking_disease[,smoking], smoking_disease[,disease])
test_result

# Extract p-value
test_result$p.value
```

<div class="callout callout-tip">  
<div class="callout-title">Interpreting Results</div>  

- **p-value < 0.05**: Reject H₀ → Variables are related  
- **p-value ≥ 0.05**: Fail to reject H₀ → No evidence of relationship  

In our example, if p < 0.05, smoking status and disease are related!  

</div>  

---  

### 5.3. Fisher's Exact Test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   - Visualizing categorical data  
   - Chi-squared test  
   → Fisher's exact test  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Fisher's Exact Test</div>  

**Purpose:** Test independence for **small sample sizes**  

**Use when:**  
- Sample size is small (< 5 expected counts in any cell)  
- Chi-squared assumptions are violated  

**Same hypotheses as chi-squared test**  

</div>  

**About the `small_study` dataset:**  
This dataset contains results from a small clinical trial with 20 patients. Half received a Drug and half received a Placebo. We measured whether each patient Improved or Not_Improved.  

**Visualize the data:**  

```{r fisher-viz, exercise=TRUE, exercise.setup="shared-data"}
# Visualize treatment outcomes
ggplot(small_study, aes(treatment, fill = outcome)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(
        title = "Treatment Outcomes (Small Study)",
        y = "Proportion"
    )

# Look at the table
table(small_study[,treatment], small_study[,outcome])
```

**Perform Fisher's exact test:**  

```{r fisher-test, exercise=TRUE, exercise.setup="shared-data"}
# Fisher's exact test
fisher_result <- fisher.test(small_study[,treatment], small_study[,outcome])
fisher_result
```

**Your Turn:** Test if there's a relationship between `Continent` and `Gender` in the `tidy` dataset:  

```{r chi-practice, exercise=TRUE, exercise.setup="shared-data"}
# 1. Visualize the relationship
# 2. Create a table
# 3. Perform chi-squared test


```

```{r chi-practice-hint}
# Visualization
ggplot(tidy, aes(Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light()

# Test
chisq.test(tidy[,Continent], tidy[,Gender])
```

```{r chi-practice-solution}
# Visualization
ggplot(tidy, aes(Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(title = "Gender Distribution by Continent", y = "Proportion")

# Table
table(tidy[,Continent], tidy[,Gender])

# Chi-squared test
chisq.test(tidy[,Continent], tidy[,Gender])
```

---  

## 6. Summary and Quick Reference  

### 6.1. Congratulations!  

You've learned how to combine data manipulation, visualization, and statistical testing!  

### 6.2. What You've Learned  

**Data Manipulation (data.table):**  
- `DT[i, j, by]` syntax for filtering, selecting, computing  
- Grouping with `by`  
- Calculating summaries with `.N`, `mean()`, `sd()`  

**Statistical Tests:**  

| Test | Purpose | When to Use | R Function |  
|------|---------|-------------|------------|  
| **Chi-squared** | Test independence of categorical variables | 2+ categorical variables, n ≥ 5 per cell | `chisq.test()` |  
| **Fisher's exact** | Test independence (small samples) | 2 categorical variables, small n | `fisher.test()` |  

**Visualization (ggplot2):**  
- Bar charts for categorical data  
- Histograms and density plots for distributions  
- Boxplots and violin plots for group comparisons  
- Always visualize before testing!  

### 6.3. Key Workflow  

1. **Explore**: Use data.table to understand your data  
2. **Visualize**: Create plots with ggplot2  
3. **Test**: Choose and perform appropriate statistical test  
4. **Interpret**: Look at p-values and effect sizes  
5. **Communicate**: Report findings clearly  

### 6.4. Best Practices  

<div class="callout callout-tip">  
<div class="callout-title">Analysis Checklist</div>  

✓ Always visualize your data first  
✓ Check assumptions before testing  
✓ Use appropriate test for your data type  
✓ Report effect sizes, not just p-values  
✓ Consider practical significance, not just statistical significance  
✓ Document your analysis workflow  

</div>  

---  

### 6.5. Download Your Work  

<div class="callout callout-tip">  
<div class="callout-title">Save Your Work</div>  

Save this completed tutorial as an HTML file for future reference.  
</div>  

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---  

### 6.6. Please Give Feedback!  

<div class="callout callout-success">  
<div class="callout-title">💬 Your Feedback Matters</div>  

Help us improve this tutorial! Share your experience and suggestions.  
</div>  

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/F3xueJLoTrVyPPLM7" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```

---  

### 6.7. Additional Resources  

**R Packages:**  
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)  
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)  

**Statistics Resources:**  
- [Quick-R: Basic Statistics](https://www.statmethods.net/stats/index.html)  
- [R for Data Science](https://r4ds.had.co.nz/)  

**Statistical Tests:**  
- [Choosing the Right Test](https://www.scribbr.com/statistics/statistical-tests/)  
- [Statistical Test Decision Tree](https://stats.oarc.ucla.edu/other/mult-pkg/whatstat/)  
