---  
title: "Basic Data Analysis with data.table and Statistics"  
output:   
  learnr::tutorial:  
    progressive: true  
    allow_skip: true  
    css: "../assets/kuzman_theme.css"  
runtime: shiny_prerendered  
---  

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
library(ggplot2)
library(openxlsx)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()

# Main datasets
tidy <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
iris_dt <- as.data.table(iris)
mtcars_dt <- as.data.table(mtcars)

# Create additional datasets for statistical tests
set.seed(123)

# Dataset for paired t-test: before/after weight measurements
weight_data <- data.table(
  participant = 1:30,
  before = rnorm(30, mean = 75, sd = 10),
  after = rnorm(30, mean = 72, sd = 10) - runif(30, 0, 5)
)
weight_data[, difference := after - before]

# Prepare long format for plotting
weight_long <- melt(weight_data[, .(participant, before, after)],
                    id.vars = "participant",
                    variable.name = "time",
                    value.name = "weight")

# Dataset for ANOVA: plant growth with 3 fertilizer types
growth_data <- data.table(
  fertilizer = rep(c("Control", "TypeA", "TypeB"), each = 20),
  growth_cm = c(
    rnorm(20, mean = 15, sd = 3),
    rnorm(20, mean = 18, sd = 3),
    rnorm(20, mean = 21, sd = 3)
  )
)

# Dataset for chi-squared: smoking and disease
smoking_disease <- data.table(
  smoking = rep(c("Smoker", "Non-smoker"), times = c(120, 180)),
  disease = c(
    rep(c("Disease", "Healthy"), times = c(40, 80)),
    rep(c("Disease", "Healthy"), times = c(30, 150))
  )
)

# Small dataset for Fisher's exact test
small_study <- data.table(
  treatment = rep(c("Drug", "Placebo"), times = c(10, 10)),
  outcome = c(
    rep(c("Improved", "Not_Improved"), times = c(7, 3)),
    rep(c("Improved", "Not_Improved"), times = c(3, 7))
  )
)

# Dataset for one-sample t-test: blood pressure measurements
bp_data <- data.table(
  patient = 1:25,
  systolic_bp = rnorm(25, mean = 125, sd = 12)
)

# Prepare subset for exercises
iris_subset <- iris_dt[Species %in% c("setosa", "versicolor")]

# Calculate summary statistics for visualization
height_summary <- tidy[, .(
  mean_height = mean(Height),
  sd_height = sd(Height),
  n = .N
), by = Gender]

mean_height <- tidy[, .(mean_height = mean(Height)), by = Gender]

growth_summary <- growth_data[, .(
  mean_growth = mean(growth_cm),
  se_growth = sd(growth_cm) / sqrt(.N),
  n = .N
), by = fertilizer]
growth_summary[, ci_lower := mean_growth - 1.96 * se_growth]
growth_summary[, ci_upper := mean_growth + 1.96 * se_growth]
```

## 1. Welcome  

Welcome to **Basic Data Analysis with data.table and Statistics**  

### 1.1. Author  

**dr.sc. Maja Kuzman** - Kuzman Consulting d.o.o.  

<div class="callout callout-info">  
<div class="callout-title">How to Use This Tutorial</div>  

- Click **"Run Code"** to execute R code  
    - ctrl + enter (windows) or command + enter (macos)  
- Click **"Hint"** if you need help  
- Progress is tracked as you complete exercises  

</div>  

<div class="callout callout-success">  
<div class="callout-title">What You Will Learn</div>  

This tutorial combines data manipulation with statistical analysis:  

1. **Introduction to data.table**  
   - Why data.table?  
   - Converting to data.table  
   - The DT[i, j, by] syntax  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   - Grouping data  

3. **Categorical Data Analysis**  
   - Visualizing categorical data  
   - Chi-squared test  
   - Fisher's exact test  

4. **Comparing Means: t-tests**  
   - One-sample t-test  
   - Two-sample t-test  
   - Paired samples t-test  
   - Visualizing group differences  

5. **Comparing Multiple Groups**  
   - ANOVA (Analysis of Variance)  
   - Visualizing multiple groups  
   - Post-hoc analysis  

6. **Integrated Workflow**  
   - Complete analysis examples  
   - From data manipulation to visualization to testing  

</div>  

---  

## 2. Quick Review: Subsetting and Conditions  

Before diving into data.table, let's review key concepts from **IntroToR** that are essential for data manipulation and filtering.  

### 2.1. Logical Subsetting  

You can subset vectors using TRUE/FALSE (logical) values:  

```{r logical-review, exercise=TRUE}
# Create a vector
numbers <- c(10, 25, 30, 15, 40)

# Create a logical vector
selection <- c(TRUE, FALSE, TRUE, FALSE, TRUE)

# Subset: keep only TRUE positions
numbers[selection]  # Returns: 10, 30, 40
```

<div class="callout callout-info">  
**💡 Key Concept:** When you put a logical vector inside `[ ]`, R keeps only elements where the value is TRUE.  
</div>  

---  

### 2.2. Comparison Operators  

**Comparison operators** automatically create TRUE/FALSE vectors:  

```{r comparison-review, exercise=TRUE}
ages <- c(18, 22, 35, 17, 40, 16, 25)

# Check which ages are greater than 18
ages > 18  # Returns: FALSE TRUE TRUE FALSE TRUE FALSE TRUE

# Check which ages equal 25
ages == 25  # Returns: FALSE FALSE FALSE FALSE FALSE FALSE TRUE

# Check which ages are NOT 18
ages != 18  # Returns: FALSE TRUE TRUE TRUE TRUE TRUE TRUE
```

<div class="callout callout-concept">  
<div class="callout-title">Comparison Operators</div>  

- `==` equals  
- `!=` not equals  
- `>` greater than  
- `<` less than  
- `>=` greater than or equal  
- `<=` less than or equal  

</div>  

**Your Turn:** Create a logical vector that checks which temperatures are below 20 degrees, then display it:  

```{r comparison-exercise, exercise=TRUE}
temperatures <- c(18, 22, 15, 25, 19, 30, 12)

# Check which temperatures are below 20

```

```{r comparison-exercise-hint}
# Use the < operator
temperatures < 20
```

```{r comparison-exercise-solution}
temperatures <- c(18, 22, 15, 25, 19, 30, 12)
temperatures < 20
# Returns: TRUE FALSE TRUE FALSE TRUE FALSE TRUE
```

---  

### 2.3. Conditional Subsetting (Filtering)  

Combine comparisons with subsetting to **filter data based on conditions**:  

```{r conditional-review, exercise=TRUE}
ages <- c(18, 22, 35, 17, 40, 16, 25)

# Get only ages greater than 18 (adults)
adults <- ages[ages > 18]
adults  # Returns: 22, 35, 40, 25

# Get only ages between 18 and 30
young_adults <- ages[ages >= 18 & ages <= 30]
young_adults  # Returns: 18, 22, 25
```

**What's happening:**  
1. `ages > 18` creates a TRUE/FALSE vector  
2. Use it as an index: `ages[ages > 18]`  
3. R keeps only values where the condition is TRUE  

**Your Turn:** From the `heights` vector, get only the heights that are 170 or greater:  

```{r filtering-exercise, exercise=TRUE}
heights <- c(165, 180, 155, 172, 190, 168, 175)

# Filter to keep only heights >= 170

```

```{r filtering-exercise-hint}
# Use heights >= 170 as the condition inside [ ]
heights[heights >= 170]
```

```{r filtering-exercise-solution}
heights <- c(165, 180, 155, 172, 190, 168, 175)
heights[heights >= 170]
# Returns: 180, 172, 190, 175
```

---  

### 2.4. Logical Operators  

Combine multiple conditions using logical operators:  

```{r logical-operators-review, exercise=TRUE}
scores <- c(45, 75, 82, 91, 68, 55, 88)

# AND (&): Both conditions must be TRUE
# Get scores between 70 and 85
scores[scores >= 70 & scores <= 85]  # Returns: 75, 82

# OR (|): At least one condition must be TRUE
# Get scores below 50 or above 90
scores[scores < 50 | scores > 90]  # Returns: 45, 91

# NOT (!): Flip TRUE/FALSE
# Get scores NOT between 60 and 80
scores[!(scores >= 60 & scores <= 80)]  # Returns: 45, 82, 91, 55, 88
```

<div class="callout callout-concept">  
<div class="callout-title">Logical Operators</div>  

- `&` (AND): Both must be TRUE  
- `|` (OR): At least one must be TRUE  
- `!` (NOT): Flips TRUE ↔ FALSE  

**Example:** `x > 10 & x < 20` means "greater than 10 AND less than 20"  

</div>  

**Your Turn:** From the `values` vector, get only values that are either less than 15 OR greater than 25:  

```{r logical-ops-exercise, exercise=TRUE}
values <- c(10, 20, 30, 12, 28, 18, 35)

# Filter for values < 15 OR > 25

```

```{r logical-ops-exercise-hint}
# Use the | (OR) operator
values[values < 15 | values > 25]
```

```{r logical-ops-exercise-solution}
values <- c(10, 20, 30, 12, 28, 18, 35)
values[values < 15 | values > 25]
# Returns: 10, 30, 12, 28, 35
```

---  

<div class="callout callout-tip">  
<div class="callout-title">Why This Matters for data.table</div>  

These concepts are **fundamental to data.table**:  
- The `i` in `DT[i, j, by]` uses these exact filtering techniques  
- You'll write conditions like: `DT[Age > 18 & Gender == "female"]`  
- Understanding logical subsetting makes data.table intuitive!  

</div>  

---  

## 3. Introduction to data.table  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   → Why data.table?  
   - Converting to data.table  
   - The DT[i, j, by] syntax  

</div>  

### 3.1. Why data.table?  

<div class="callout callout-info">  
<div class="callout-title">Why Use data.table?</div>  

**data.table** is a powerful package for working with data in R:  

- **Fast** - extremely efficient with large datasets  
- **Concise** - less typing, clearer code  
- **Powerful** - complex operations made simple  
- **Integrates well** - works seamlessly with ggplot2 and statistical functions  

</div>  

### 3.2. Installing and Loading  

Install data.table once:  

```{r install-dt, echo=TRUE, eval=FALSE}
# Install data.table (only once)
install.packages("data.table")
```

Load it every session:  

```{r load-dt, echo=TRUE, eval=FALSE}
# Load data.table
library(data.table)
library(ggplot2)  # We'll use this for visualization
```

---  

### 3.3. Converting to data.table  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   - Why data.table?  
   → Converting to data.table  
   - The DT[i, j, by] syntax  

</div>  

#### 3.3.1. Reading from Excel Files  

In practice, you'll often receive data in Excel format. Here's the typical workflow:  

```r
# Install openxlsx package if needed
# install.packages("openxlsx")

# Load library
library(openxlsx)
library(data.table)

# Read Excel file
my_data <- read.xlsx("path/to/your/file.xlsx")

# Convert to data.table
my_dt <- as.data.table(my_data)
```

<div class="callout callout-info">  
**💡 Tip:** The `read.xlsx()` function from the `openxlsx` package reads Excel files into a regular data.frame. Use `as.data.table()` to convert it to data.table format for efficient manipulation.  
</div>  

#### 3.3.2. Reading Text Files with fread()  

The `fread()` function (**f**ast **read**) is data.table's powerful function for reading text-based files. It automatically detects file format, separators, and data types, making it very convenient and much faster than base R functions.  

<div class="callout callout-concept">  
<div class="callout-title">🎯 Why use fread()?</div>  

- **Automatic detection**: Detects separators (comma, tab, semicolon, etc.), column types, and headers  
- **Speed**: Much faster than `read.csv()` or `read.table()` for large files  
- **Direct to data.table**: Returns a data.table object directly (no conversion needed)  
- **Flexible**: Works with CSV, TSV, TXT, and other delimited files  
- **URLs supported**: Can read directly from web URLs  

</div>  

```{r fread-example, exercise=TRUE, exercise.setup="shared-data"}
# Example 1: Reading from a URL (like we did with tidy dataset)
# fread automatically creates a data.table
tidy_data <- fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt")
head(tidy_data)

# Example 2: CSV file (comma-separated)
# fread("mydata.csv")

# Example 3: TSV file (tab-separated)
# fread("mydata.tsv")

# Example 4: TXT file with custom separator
# fread("mydata.txt", sep="|")  # pipe-separated

# Example 5: Specifying options explicitly
# fread("mydata.csv", 
#       header = TRUE,        # First row is column names
#       sep = ",",           # Comma separator (auto-detected usually)
#       na.strings = "NA")   # What represents missing values
```

<div class="callout callout-info">  
**💡 Common file formats:**  

- **CSV** (Comma-Separated Values): `mydata.csv` - values separated by commas  
- **TSV** (Tab-Separated Values): `mydata.tsv` or `mydata.tab` - values separated by tabs  
- **TXT** (Text file): `mydata.txt` - can use any delimiter (space, tab, comma, semicolon, pipe, etc.)  

`fread()` automatically detects the separator in most cases, so you usually just need: `fread("filename")`  
</div>  

#### 3.3.3. Converting Built-in Datasets  

You can also convert R's built-in datasets to data.table:  

```{r convert-example, exercise=TRUE}
# Look at regular iris
head(iris)

# Convert to data.table
iris_dt <- as.data.table(iris)
head(iris_dt)
```

---  

### 3.4. The DT[i, j, by] Syntax  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

1. **Introduction to data.table**  
   - Why data.table?  
   - Converting to data.table  
   → The DT[i, j, by] syntax  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">The Magic Formula: DT[i, j, by]</div>  

Think of a data.table like a spreadsheet where you can:  

- **i** = which **rows** (filter/subset)  
- **j** = what to do with **columns** (select/compute)  
- **by** = grouped **by** what (grouping variable)  

**Example:** `DT[Gender == "Female", mean(Height), by = Continent]`  
→ For female rows, calculate mean height, separately for each continent  

</div>  

Let's see this in action:  

```{r syntax-demo, exercise=TRUE, exercise.setup="shared-data"}
# All rows, all columns
iris_dt

# Filter rows: only setosa
iris_dt[Species == "setosa"]

# Select columns
iris_dt[, .(Species, Sepal.Length)]

# Compute: mean sepal length
iris_dt[, mean(Sepal.Length)]

# Group by: mean sepal length per species
iris_dt[, mean(Sepal.Length), by = Species]
```

---  

## 4. Basic Data Operations  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   → Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   - Grouping data  

</div>  

### 4.1. Selecting Rows (Filtering)  

**About the `tidy` dataset:**  
This dataset contains information about people including their Gender ("male" or "female"), Height (in cm), Weight (in kg), Age (in years), and Continent.  

Use logical conditions in the **i** position to filter rows:  

**Example 1: Filter by a single condition**  

```{r filter-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Select only females
tidy[Gender == "female"]
```

**Example 2: Filter by numeric condition**  

```{r filter-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Select people taller than 180 cm
tidy[Height > 180]
```

**Example 3: Combine multiple conditions**  

```{r filter-demo3, exercise=TRUE, exercise.setup="shared-data"}
# Select females taller than 170 cm
tidy[Gender == "female" & Height > 170]
```

**Your Turn:** From the `iris_dt` data, select only rows where `Sepal.Length > 6`:  

```{r filter-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Filter iris for Sepal.Length > 6


```

```{r filter-exercise-hint}
iris_dt[Sepal.Length > 6]
```

```{r filter-exercise-solution}
iris_dt[Sepal.Length > 6]
```

---  

### 4.2. Selecting Columns  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   → Selecting columns  
   - Computing summaries  
   - Grouping data  

</div>  

Use `.()` or `list()` in the **j** position to select columns:  

**Example 1: Select one column**  

```{r select-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Select one column (returns a vector)
tidy[, Gender]
```

**Example 2: Select multiple columns**  

```{r select-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Select multiple columns (returns a data.table)
# Using .() which is shorthand for list()
tidy[, .(Gender, Age, Height)]
```

**Example 3: Alternative syntax**  

```{r select-demo3, exercise=TRUE, exercise.setup="shared-data"}
# Can also use list() directly
tidy[, list(Gender, Height)]
```

**Your Turn:** Select the `Species` and `Petal.Length` columns from `iris_dt`:  

```{r select-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Select Species and Petal.Length


```

```{r select-exercise-hint}
iris_dt[, .(Species, Petal.Length)]
```

```{r select-exercise-solution}
iris_dt[, .(Species, Petal.Length)]
```

---  

### 4.3. Computing Summaries  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   → Computing summaries  
   - Grouping data  

</div>  

In the **j** position, you can do more than just select existing columns - you can also **create new columns with computations**:  

**Creating New Columns:**  

```{r compute-new-cols, exercise=TRUE, exercise.setup="shared-data"}
# Create a new column: ratio of Height to Weight
tidy[, .(Gender, Height, Weight, Height_Weight_Ratio = Height / Weight)]

# Multiple new computed columns
tidy[, .(
  Gender,
  Height,
  Weight,
  Height_m = Height / 100,          # Convert cm to meters
  BMI = Weight / (Height/100)^2     # Calculate BMI
)]
```

<div class="callout callout-info">  
**💡 Key Concept:** Inside `.()` in the j position:  
- Column names alone select existing columns: `.(Gender, Height)`  
- `name = expression` creates new computed columns: `.(BMI = Weight / (Height/100)^2)`  
- You can mix both: select existing and create new columns together!  
</div>  

---  

**Computing Summary Statistics:**  

Now for something powerful: instead of just listing or modifying columns, we can **collapse the entire dataset into summary statistics**!  

<div class="callout callout-concept">  
<div class="callout-title">🎯 From Data to Summaries</div>  

When you use **functions** like `mean()`, `median()`, `sd()` in the j position, data.table:  
- **Aggregates** all rows into a single summary value  
- Returns the computed statistic(s), not the original data  
- Can compute **multiple summaries at once** inside `.()`   

This is fundamentally different from selecting or computing columns!  

</div>  

```{r compute-demo, exercise=TRUE, exercise.setup="shared-data"}
# Single summary: Mean height (returns one number)
tidy[, mean(Height)]

# Multiple summaries at once (returns a data.table with one row)
tidy[, .(
  mean_height = mean(Height),
  median_height = median(Height),
  sd_height = sd(Height),
  min_height = min(Height),
  max_height = max(Height)
)]

# Special variable .N = count number of rows
tidy[, .N]
```

<div class="callout callout-tip">  
**💡 Notice:**   
- **One summary:** `DT[, mean(col)]` returns a single value  
- **Multiple summaries:** `DT[, .(stat1 = mean(col), stat2 = sd(col))]` returns a data.table with one row containing all statistics  
- **Combining makes sense:** Get mean, median, sd, min, max all at once!  
</div>  

**Your Turn:** Calculate the mean and standard deviation of `Sepal.Length` from `iris_dt`:  

```{r compute-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Calculate mean and sd of Sepal.Length


```

```{r compute-exercise-hint}
iris_dt[, .(mean_sepal = mean(Sepal.Length), sd_sepal = sd(Sepal.Length))]
```

```{r compute-exercise-solution}
iris_dt[, .(
  mean_sepal = mean(Sepal.Length),
  sd_sepal = sd(Sepal.Length)
)]
```

---  

### 4.4. Grouping Data  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

2. **Basic Data Operations**  
   - Selecting rows (filtering)  
   - Selecting columns  
   - Computing summaries  
   → Grouping data  

</div>  

Use **by** to compute statistics for each group:  

```{r group-demo, exercise=TRUE, exercise.setup="shared-data"}
# Mean height by gender
tidy[, mean(Height), by = Gender]

# Multiple statistics by group
tidy[, .(
  mean_height = mean(Height),
  count = .N
), by = Gender]

# Group by multiple variables
tidy[, mean(Height), by = .(Gender, Continent)]
```

**Your Turn:** Calculate the mean `Petal.Length` for each `Species` in `iris_dt`:  

```{r group-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Mean Petal.Length by Species


```

```{r group-exercise-hint}
iris_dt[, mean(Petal.Length), by = Species]
```

```{r group-exercise-solution}
iris_dt[, .(mean_petal_length = mean(Petal.Length)), by = Species]
```

---  

### 4.5. Practice: Combining Operations  

Now combine filtering, selecting, and grouping:  

**Challenge:** From the `tidy` data:  
1. Filter for people with Height > 165  
2. Calculate mean Age and count  
3. Group by Gender  

```{r combine-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Your code here


```

```{r combine-exercise-hint}
# Remember: DT[i, j, by]
# i = Height > 165
# j = calculate mean Age and count
# by = Gender
```

```{r combine-exercise-solution}
tidy[Height > 165, .(mean_age = mean(Age), count = .N), by = Gender]
```

---  

## 5. Categorical Data Analysis  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   → Visualizing categorical data  
   - Chi-squared test  
   - Fisher's exact test  

</div>  

### 5.1. Visualizing Categorical Data  

**About the `smoking_disease` dataset:**  
This dataset contains data from 300 people, including their smoking status (Smoker or Non-smoker) and disease status (Disease or Healthy). We want to know if smoking is associated with disease.  

Before testing, always visualize your data!  

**Example 1: Single variable bar chart**  

```{r cat-viz-demo1, exercise=TRUE, exercise.setup="shared-data"}
# Bar chart showing count of smokers vs non-smokers
ggplot(smoking_disease, aes(smoking, fill = smoking)) +
    geom_bar() +
    theme_light() +
    labs(title = "Smoking Status Distribution", y = "Count")
```

**Example 2: Stacked proportional bar chart**  

```{r cat-viz-demo2, exercise=TRUE, exercise.setup="shared-data"}
# Stacked bar chart showing disease within each smoking group
ggplot(smoking_disease, aes(smoking, fill = disease)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(
        title = "Disease Prevalence by Smoking Status",
        y = "Proportion"
    )
```

**Your Turn:** Create a bar chart showing the distribution of `Species` in `iris_dt`:  

```{r cat-viz-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Create bar chart of Species


```

```{r cat-viz-exercise-solution}
ggplot(iris_dt, aes(Species, fill = Species)) +
    geom_bar() +
    theme_light() +
    labs(title = "Iris Species Distribution")
```

---  

### 5.2. Chi-Squared Test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   - Visualizing categorical data  
   → Chi-squared test  
   - Fisher's exact test  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Chi-Squared Test</div>  

**Purpose:** Test if two categorical variables are independent  

**Question:** "Is there a relationship between smoking and disease?"  

**Requirements:**  
- Categorical variables (e.g., yes/no, groups)  
- Expected cell counts ≥ 5  
- Independent observations  

**Hypotheses:**  
- H₀: Variables are independent (no relationship)  
- H₁: Variables are related  

</div>  

First, create a contingency table:  

```{r chi-table, exercise=TRUE, exercise.setup="shared-data"}
# Count observations using .N
smoking_disease[, .N, by = .(smoking, disease)]

# Create contingency table for the test
smoking_table <- table(smoking_disease[,smoking], smoking_disease[,disease])
smoking_table
```

Now perform the chi-squared test:  

```{r chi-test, exercise=TRUE, exercise.setup="shared-data"}
# Chi-squared test
test_result <- chisq.test(smoking_disease[,smoking], smoking_disease[,disease])
test_result

# Extract p-value
test_result$p.value
```

<div class="callout callout-tip">  
<div class="callout-title">Interpreting Results</div>  

- **p-value < 0.05**: Reject H₀ → Variables are related  
- **p-value ≥ 0.05**: Fail to reject H₀ → No evidence of relationship  

In our example, if p < 0.05, smoking status and disease are related!  

</div>  

---  

### 5.3. Fisher's Exact Test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

3. **Categorical Data Analysis**  
   - Visualizing categorical data  
   - Chi-squared test  
   → Fisher's exact test  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Fisher's Exact Test</div>  

**Purpose:** Test independence for **small sample sizes**  

**Use when:**  
- Sample size is small (< 5 expected counts in any cell)  
- Chi-squared assumptions are violated  

**Same hypotheses as chi-squared test**  

</div>  

**About the `small_study` dataset:**  
This dataset contains results from a small clinical trial with 20 patients. Half received a Drug and half received a Placebo. We measured whether each patient Improved or Not_Improved.  

**Visualize the data:**  

```{r fisher-viz, exercise=TRUE, exercise.setup="shared-data"}
# Visualize treatment outcomes
ggplot(small_study, aes(treatment, fill = outcome)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(
        title = "Treatment Outcomes (Small Study)",
        y = "Proportion"
    )

# Look at the table
table(small_study[,treatment], small_study[,outcome])
```

**Perform Fisher's exact test:**  

```{r fisher-test, exercise=TRUE, exercise.setup="shared-data"}
# Fisher's exact test
fisher_result <- fisher.test(small_study[,treatment], small_study[,outcome])
fisher_result
```

**Your Turn:** Test if there's a relationship between `Continent` and `Gender` in the `tidy` dataset:  

```{r chi-practice, exercise=TRUE, exercise.setup="shared-data"}
# 1. Visualize the relationship
# 2. Create a table
# 3. Perform chi-squared test


```

```{r chi-practice-hint}
# Visualization
ggplot(tidy, aes(Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light()

# Test
chisq.test(tidy[,Continent], tidy[,Gender])
```

```{r chi-practice-solution}
# Visualization
ggplot(tidy, aes(Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(title = "Gender Distribution by Continent", y = "Proportion")

# Table
table(tidy[,Continent], tidy[,Gender])

# Chi-squared test
chisq.test(tidy[,Continent], tidy[,Gender])
```

---  

## 6. Comparing Means: t-tests  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

4. **Comparing Means: t-tests**  
   → One-sample t-test  
   - Two-sample t-test  
   - Paired samples t-test  
   - Visualizing group differences  

</div>  

### 6.1. One-Sample t-test  

<div class="callout callout-concept">  
<div class="callout-title">When to Use One-Sample t-test</div>  

**Purpose:** Test if a sample mean differs from a known value  

**Question:** "Is the average blood pressure different from 120?"  

**Requirements:**  
- Continuous variable  
- Approximately normally distributed  
- Independent observations  

**Hypotheses:**  
- H₀: μ = known value  
- H₁: μ ≠ known value  

</div>  

**About the `bp_data` dataset:**  
This dataset contains systolic blood pressure measurements from 25 patients. Normal systolic blood pressure is around 120 mmHg. We want to test if our sample mean differs from this normal value.  

**Step 1: Visualize and summarize the data**  

```{r onesample-viz, exercise=TRUE, exercise.setup="shared-data"}
# Summary statistics
bp_data[, .(
  mean_bp = mean(systolic_bp),
  sd_bp = sd(systolic_bp),
  n = .N
)]

# Histogram with reference line at normal BP (120)
ggplot(bp_data, aes(systolic_bp)) +
    geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "black") +
    geom_vline(xintercept = 120, color = "red", linetype = "dashed", size = 1) +
    annotate("text", x = 120, y = 5, label = "Normal BP = 120", color = "red", hjust = -0.1) +
    theme_light() +
    labs(title = "Blood Pressure Distribution", x = "Systolic BP (mmHg)", y = "Count")
```

**Step 2: Perform the one-sample t-test**  

```{r onesample-test, exercise=TRUE, exercise.setup="shared-data"}
# Test if mean BP differs from 120
t.test(bp_data[,systolic_bp], mu = 120)
```

**Your Turn:** Test if the mean `Sepal.Length` in iris differs from 6:  

```{r onesample-exercise, exercise=TRUE, exercise.setup="shared-data"}
# 1. Calculate mean Sepal.Length
# 2. Create a histogram with reference line at 6
# 3. Perform one-sample t-test


```

```{r onesample-exercise-hint}
# Mean
iris_dt[, mean(Sepal.Length)]

# Test
t.test(iris_dt[,Sepal.Length], mu = 6)
```

```{r onesample-exercise-solution}
# Summary
iris_dt[, .(mean_sepal = mean(Sepal.Length), sd_sepal = sd(Sepal.Length))]

# Visualization
ggplot(iris_dt, aes(Sepal.Length)) +
    geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7, color = "black") +
    geom_vline(xintercept = 6, color = "red", linetype = "dashed", size = 1) +
    theme_light() +
    labs(title = "Sepal Length Distribution", x = "Sepal Length")

# Test
t.test(iris_dt[,Sepal.Length], mu = 6)
```

---  

### 6.2. Two-Sample t-test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

4. **Comparing Means: t-tests**  
   - One-sample t-test  
   → Two-sample t-test  
   - Paired samples t-test  
   - Visualizing group differences  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Two-Sample t-test</div>  

**Purpose:** Compare means between **two independent groups**  

**Question:** "Do males and females have different average heights?"  

**Requirements:**  
- Continuous outcome variable  
- Categorical grouping variable (2 groups)  
- Independent observations  
- Approximately normal distributions (or large sample)  

**Hypotheses:**  
- H₀: μ₁ = μ₂ (means are equal)  
- H₁: μ₁ ≠ μ₂ (means differ)  

</div>  

First, explore the data with data.table:  

```{r twosample-explore, exercise=TRUE, exercise.setup="shared-data"}
# Summary by gender
tidy[, .(
  mean_height = mean(Height),
  sd_height = sd(Height),
  n = .N
), by = Gender]
```

Visualize the distributions:  

```{r twosample-viz, exercise=TRUE, exercise.setup="shared-data"}
# Boxplot
ggplot(tidy, aes(Gender, Height, fill = Gender)) +
    geom_boxplot(alpha = 0.7) +
    theme_light() +
    labs(title = "Height by Gender", y = "Height (cm)")

# Density plot
ggplot(tidy, aes(Height, fill = Gender)) +
    geom_density(alpha = 0.5) +
    theme_light() +
    labs(title = "Height Distribution by Gender")
```

Perform the two-sample t-test:  

```{r twosample-test, exercise=TRUE, exercise.setup="shared-data"}
# Two-sample t-test
t.test(Height ~ Gender, data = tidy)

# Alternative: split data manually
females <- tidy[Gender == "female", Height]
males <- tidy[Gender == "male", Height]
t.test(females, males)
```

**Your Turn:** Compare `Petal.Length` between `setosa` and `versicolor` species in iris:  

```{r twosample-exercise, exercise=TRUE, exercise.setup="shared-data"}
# 1. Filter for setosa and versicolor only
# 2. Calculate summary statistics by species
# 3. Create a boxplot
# 4. Perform two-sample t-test


```

```{r twosample-exercise-hint}
# Filter
iris_subset <- iris_dt[Species %in% c("setosa", "versicolor")]

# Summary
iris_subset[, .(mean_petal = mean(Petal.Length)), by = Species]

# Test
t.test(Petal.Length ~ Species, data = iris_subset)
```

```{r twosample-exercise-solution}
# Filter data
iris_subset <- iris_dt[Species %in% c("setosa", "versicolor")]

# Summary statistics
iris_subset[, .(
  mean_petal = mean(Petal.Length),
  sd_petal = sd(Petal.Length),
  n = .N
), by = Species]

# Visualization
ggplot(iris_subset, aes(Species, Petal.Length, fill = Species)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.3) +
    theme_light() +
    labs(title = "Petal Length: Setosa vs Versicolor")

# Two-sample t-test
t.test(Petal.Length ~ Species, data = iris_subset)
```

---  

### 6.3. Paired Samples t-test  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

4. **Comparing Means: t-tests**  
   - One-sample t-test  
   - Two-sample t-test  
   → Paired samples t-test  
   - Visualizing group differences  

</div>  

<div class="callout callout-concept">  
<div class="callout-title">When to Use Paired Samples t-test</div>  

**Purpose:** Compare means from **same subjects measured twice**  

**Question:** "Did weight change after a diet program?"  

**Requirements:**  
- Two measurements on same subjects (before/after, matched pairs)  
- Continuous variable  
- Differences approximately normally distributed  

**Hypotheses:**  
- H₀: μ_diff = 0 (no change)  
- H₁: μ_diff ≠ 0 (there is change)  

**Key:** We test the **differences**, not the groups!  

</div>  

First, calculate the differences:  

```{r paired-prep, exercise=TRUE, exercise.setup="shared-data"}
# Add difference column
weight_data[, difference := after - before]

# Summary statistics
weight_data[, .(
  mean_before = mean(before),
  mean_after = mean(after),
  mean_diff = mean(difference),
  sd_diff = sd(difference)
)]
```

Visualize before/after:  

```{r paired-viz, exercise=TRUE, exercise.setup="shared-data"}
# Reshape for visualization
weight_long <- melt(weight_data, 
                    id.vars = "participant",
                    measure.vars = c("before", "after"),
                    variable.name = "time",
                    value.name = "weight")

# Boxplot comparison
ggplot(weight_long, aes(time, weight, fill = time)) +
    geom_boxplot(alpha = 0.7) +
    geom_line(aes(group = participant), alpha = 0.3) +
    theme_light() +
    labs(title = "Weight Before and After Diet Program", y = "Weight (kg)")

# Histogram of differences
ggplot(weight_data, aes(difference)) +
    geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "black") +
    geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
    theme_light() +
    labs(title = "Weight Change Distribution", x = "Change (After - Before)")
```

Perform the paired t-test:  

```{r paired-test, exercise=TRUE, exercise.setup="shared-data"}
# Paired t-test
t.test(weight_data[,after], weight_data[,before], paired = TRUE)

# Check: same as one-sample test on differences
t.test(weight_data[,difference], mu = 0)
```

---  

### 6.4. Visualizing Group Differences  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

4. **Comparing Means: t-tests**  
   - One-sample t-test  
   - Two-sample t-test  
   - Paired samples t-test  
   → Visualizing group differences  

</div>  

Multiple ways to visualize group comparisons:  

```{r viz-comparison, exercise=TRUE, exercise.setup="shared-data"}
# 1. Boxplot with individual points
ggplot(tidy, aes(Gender, Height, fill = Gender)) +
    geom_boxplot(alpha = 0.5) +
    geom_jitter(width = 0.1, alpha = 0.3) +
    theme_light()

# 2. Violin plot
ggplot(tidy, aes(Gender, Height, fill = Gender)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.1, alpha = 0.5) +
    theme_light()

# 3. Density plot with means
mean_height <- tidy[, .(mean_height = mean(Height)), by = Gender]

ggplot(tidy, aes(Height, fill = Gender)) +
    geom_density(alpha = 0.5) +
    geom_vline(data = mean_height, aes(xintercept = mean_height, color = Gender),
               linetype = "dashed", size = 1) +
    theme_light()
```

---  

## 7. Comparing Multiple Groups  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

5. **Comparing Multiple Groups**  
   → ANOVA (Analysis of Variance)  
   - Visualizing multiple groups  
   - Post-hoc analysis  

</div>  

### 7.1. ANOVA (Analysis of Variance)  

<div class="callout callout-concept">  
<div class="callout-title">When to Use ANOVA</div>  

**Purpose:** Compare means across **three or more groups**  

**Question:** "Do different fertilizers produce different plant growth?"  

**Requirements:**  
- Continuous outcome variable  
- Categorical grouping variable (≥ 3 groups)  
- Independent observations  
- Approximately normal distributions within groups  
- Equal variances across groups (homogeneity)  

**Hypotheses:**  
- H₀: All group means are equal  
- H₁: At least one mean differs  

**Important:** ANOVA tells you if *any* groups differ, not *which* ones!  

</div>  

**About the `growth_data` dataset:**  
This dataset contains plant growth measurements (in cm) from an experiment with 60 plants. Plants were randomly assigned to three fertilizer groups: Control (no fertilizer), TypeA, and TypeB (20 plants per group).  

**Step 1: Explore the data**  

```{r anova-explore, exercise=TRUE, exercise.setup="shared-data"}
# Summary statistics by fertilizer type
growth_data[, .(
  mean_growth = mean(growth_cm),
  sd_growth = sd(growth_cm),
  n = .N
), by = fertilizer]
```

**Step 2: Visualize the groups**  

Create visualizations to compare the three groups.  

**Visualization 1: Boxplot with individual points**  

```{r anova-viz1, exercise=TRUE, exercise.setup="shared-data"}
# Boxplot showing distribution for each fertilizer type
ggplot(growth_data, aes(x = fertilizer, y = growth_cm, fill = fertilizer)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.3) +
    theme_light() +
    labs(title = "Plant Growth by Fertilizer Type",
         x = "Fertilizer", y = "Growth (cm)")
```

**Visualization 2: Density plot**  

```{r anova-viz2, exercise=TRUE, exercise.setup="shared-data"}
# Density plot showing overlapping distributions
ggplot(growth_data, aes(x = growth_cm, fill = fertilizer)) +
    geom_density(alpha = 0.5) +
    theme_light() +
    labs(title = "Growth Distribution by Fertilizer Type",
         x = "Growth (cm)", y = "Density")
```

**Step 3: Perform ANOVA**  

```{r anova-test, exercise=TRUE, exercise.setup="shared-data"}
# ANOVA test
anova_result <- aov(growth_cm ~ fertilizer, data = growth_data)
summary(anova_result)
```

---  

### 7.2. Post-hoc Analysis  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

5. **Comparing Multiple Groups**  
   - ANOVA (Analysis of Variance)  
   - Visualizing multiple groups  
   → Post-hoc analysis  

</div>  

<div class="callout callout-tip">  
<div class="callout-title">After ANOVA: Which Groups Differ?</div>  

If ANOVA is significant (p < 0.05), we know groups differ. But which ones?  

**Post-hoc tests** compare all pairs of groups, with correction for multiple testing.  

</div>  

Common post-hoc test: Tukey's HSD (Honest Significant Difference)  

```{r posthoc-test, exercise=TRUE, exercise.setup="shared-data"}
# Tukey's HSD test
anova_result <- aov(growth_cm ~ fertilizer, data = growth_data)
tukey_result <- TukeyHSD(anova_result)
tukey_result

# Visualize Tukey results
plot(tukey_result)
```

**Your Turn:** Perform ANOVA comparing `Sepal.Length` across all three iris species:  

```{r anova-exercise, exercise=TRUE, exercise.setup="shared-data"}
# 1. Calculate summary statistics by species
# 2. Create a boxplot
# 3. Perform ANOVA
# 4. If significant, perform Tukey's HSD


```

```{r anova-exercise-hint}
# Summary
iris_dt[, .(mean_sepal = mean(Sepal.Length)), by = Species]

# ANOVA
anova_result <- aov(Sepal.Length ~ Species, data = iris_dt)
summary(anova_result)
```

```{r anova-exercise-solution}
# Summary statistics
iris_dt[, .(
  mean_sepal = mean(Sepal.Length),
  sd_sepal = sd(Sepal.Length),
  n = .N
), by = Species]

# Visualization
ggplot(iris_dt, aes(Species, Sepal.Length, fill = Species)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.3) +
    theme_light() +
    labs(title = "Sepal Length by Species")

# ANOVA
anova_result <- aov(Sepal.Length ~ Species, data = iris_dt)
summary(anova_result)

# Post-hoc
TukeyHSD(anova_result)
```

---  

### 7.3. Visualizing Multiple Groups  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

5. **Comparing Multiple Groups**  
   - ANOVA (Analysis of Variance)  
   → Visualizing multiple groups  
   - Post-hoc analysis  

</div>  

Advanced visualizations for multiple groups:  

```{r multi-viz, exercise=TRUE, exercise.setup="shared-data"}
# Boxplot with summary statistics
ggplot(growth_data, aes(fertilizer, growth_cm, fill = fertilizer)) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.4) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
    theme_light() +
    labs(title = "Plant Growth by Fertilizer (Diamond = Mean)",
         x = "Fertilizer Type", y = "Growth (cm)") +
    theme(legend.position = "none")

# Violin plot with quartiles
ggplot(growth_data, aes(fertilizer, growth_cm, fill = fertilizer)) +
    geom_violin(alpha = 0.7) +
    geom_boxplot(width = 0.1, alpha = 0.8) +
    theme_light() +
    labs(title = "Growth Distribution by Fertilizer")

# Mean plot with error bars (95% CI)
growth_summary <- growth_data[, .(
  mean_growth = mean(growth_cm),
  se_growth = sd(growth_cm) / sqrt(.N),
  n = .N
), by = fertilizer]

growth_summary[, ci_lower := mean_growth - 1.96 * se_growth]
growth_summary[, ci_upper := mean_growth + 1.96 * se_growth]

ggplot(growth_summary, aes(fertilizer, mean_growth, fill = fertilizer)) +
    geom_col(alpha = 0.7) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    theme_light() +
    labs(title = "Mean Growth by Fertilizer (with 95% CI)",
         x = "Fertilizer", y = "Mean Growth (cm)")
```

---  

## 8. Integrated Workflow  

<div class="callout callout-success">  
<div class="callout-title">📍 In This Section</div>  

6. **Integrated Workflow**  
   → Complete analysis examples  
   - From data manipulation to visualization to testing  

</div>  

### 8.1. Complete Analysis Example 1: Height by Continent  

Let's perform a complete analysis from start to finish with explicit step-by-step instructions.  

**Research Question:** "Do people from different continents have different average heights?"  

**About the analysis:**  
We'll use ANOVA since we're comparing means across MORE than 2 groups (continents).  

---  

**STEP 1: Explore the Data**  

First, count how many observations we have per continent:  

```{r workflow1-step1, exercise=TRUE, exercise.setup="shared-data"}
# Count observations per continent
tidy[, .N, by = Continent]
```

---  

**STEP 2: Calculate Summary Statistics**  

Calculate mean, standard deviation, and count for each continent:  

```{r workflow1-step2, exercise=TRUE, exercise.setup="shared-data"}
# Summary statistics by continent
height_summary <- tidy[, .(
  mean_height = mean(Height),
  sd_height = sd(Height),
  n = .N
), by = Continent]
height_summary
```

---  

**STEP 3: Create Visualization**  

Create a boxplot to visualize height distributions across continents.  

**What to plot:**  
- **x-axis:** Continent  
- **y-axis:** Height  
- **geom:** Use `geom_boxplot()` for the boxes  
- **Add:** Use `geom_jitter()` to show individual points  
- **Color:** Fill boxes by Continent  

```{r workflow1-step3, exercise=TRUE, exercise.setup="shared-data"}
# Boxplot with individual points
ggplot(tidy, aes(x = Continent, y = Height, fill = Continent)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.3) +
    theme_light() +
    labs(title = "Height by Continent", y = "Height (cm)") +
    theme(legend.position = "none")
```

---  

**STEP 4: Perform ANOVA**  

Test if there are significant differences between continents:  

```{r workflow1-step4, exercise=TRUE, exercise.setup="shared-data"}
# Perform ANOVA
anova_result <- aov(Height ~ Continent, data = tidy)
summary(anova_result)
```

---  

**STEP 5: Post-hoc Test (If ANOVA is Significant)**  

If p < 0.05, find out WHICH continents differ from each other:  

```{r workflow1-step5, exercise=TRUE, exercise.setup="shared-data"}
# Tukey's HSD post-hoc test
anova_result <- aov(Height ~ Continent, data = tidy)
TukeyHSD(anova_result)
```

---  

### 8.2. Complete Analysis Example 2: Weight Loss Study  

**Research Question:** "Is the weight loss program effective?"  

**About the analysis:**  
We'll use a PAIRED t-test since we have the same people measured twice (before and after).  

**About the dataset:**  
`weight_data` contains weight measurements for 30 participants before and after a diet program.  

---  

**STEP 1: Calculate the Differences**  

Add a new column showing weight change for each participant:  

```{r workflow2-step1, exercise=TRUE, exercise.setup="shared-data"}
# Calculate weight loss (before - after)
# Note: Positive values mean weight loss
weight_data[, weight_loss := before - after]

# View first few rows
weight_data
```

---  

**STEP 2: Summary Statistics**  

Calculate means and standard deviations:  

```{r workflow2-step2, exercise=TRUE, exercise.setup="shared-data"}
# Calculate weight loss (before - after)
weight_data[, weight_loss := before - after]

# Summary statistics
weight_data[, .(
  mean_before = mean(before),
  mean_after = mean(after),
  mean_loss = mean(weight_loss),
  sd_loss = sd(weight_loss)
)]
```

---  

**STEP 3a: Visualize Before vs After**  

Create a boxplot comparing weights before and after.  

**What to plot:**  
- **x-axis:** time (before vs after)  
- **y-axis:** weight  
- **geom:** Use `geom_boxplot()` to compare distributions  
- **Add:** Use `geom_line()` with `group = participant` to show individual changes  

```{r workflow2-step3a, exercise=TRUE, exercise.setup="shared-data"}
# Note: weight_long is already prepared in shared-data

# Boxplot showing before/after with individual trajectories
ggplot(weight_long, aes(x = time, y = weight, fill = time)) +
    geom_boxplot(alpha = 0.7) +
    geom_line(aes(group = participant), alpha = 0.2) +
    geom_point(alpha = 0.3) +
    theme_light() +
    labs(title = "Weight Before and After Diet Program",
         x = "Time", y = "Weight (kg)")
```

---  

**STEP 3b: Visualize the Weight Changes**  

Create a histogram of the weight changes (differences).  

**What to plot:**  
- **x-axis:** difference (weight change)  
- **geom:** Use `geom_histogram()` to show distribution of changes  
- **Add:** Use `geom_vline()` at x = 0 to show the "no change" line  
- **bins:** Set to 10 for appropriate binning  

```{r workflow2-step3b, exercise=TRUE, exercise.setup="shared-data"}
# Histogram of weight changes
ggplot(weight_data, aes(x = difference)) +
    geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7, color = "black") +
    geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
    theme_light() +
    labs(title = "Weight Change Distribution", 
         x = "Change (After - Before in kg)",
         subtitle = "Negative values = weight loss, Positive values = weight gain")
```

---  

**STEP 4: Perform Paired t-test**  

Test if the weight change is significantly different from zero:  

```{r workflow2-step4, exercise=TRUE, exercise.setup="shared-data"}
# Paired t-test
t.test(weight_data[,before], weight_data[,after], paired = TRUE)

# Or equivalently, test if difference differs from 0
# t.test(weight_data[,difference], mu = 0)
```

---  

**STEP 5: Calculate Effect Size (Optional)**  

Cohen's d shows the magnitude of the effect (how big is the change?):  

```{r workflow2-step5, exercise=TRUE, exercise.setup="shared-data"}
# Calculate weight loss (before - after)
weight_data[, weight_loss := before - after]

# Calculate Cohen's d
weight_data[, .(
  cohens_d = mean(weight_loss) / sd(weight_loss)
)]

# Interpretation:
# |d| < 0.2: small effect
# |d| = 0.5: medium effect
# |d| > 0.8: large effect
```

---  

### 8.3. Final Challenge: Your Complete Analysis  

**Research Question:** "Is there a relationship between Gender and Continent, and do Heights differ by Gender and Continent?"  

This challenge combines THREE different analyses. Complete each part step-by-step.  

---  

**TASK 1: Chi-squared Test - Gender and Continent Relationship**  

Test if Gender distribution is independent of Continent.  

**Step 1a: Create a stacked bar chart**  
- Use `ggplot()` with `tidy` dataset  
- x-axis: `Continent`  
- fill: `Gender` (to show gender within each continent)  
- geom: `geom_bar(position = "fill")` (fill to show proportions)  
- Add labels with `labs()`  

**Step 1b: Create a contingency table**  
- Use `tidy[,.N,.(Continent,Gender)]`  

**Step 1c: Perform chi-squared test**  
- Use `chisq.test(tidy[,Continent], tidy[,Gender])`  

```{r final-part1, exercise=TRUE, exercise.setup="shared-data"}
# TASK 1: Your code here




```

---  

**TASK 2: Two-Sample t-test - Height by Gender**  

Compare average heights between males and females.  

**Step 2a: Calculate summary statistics**  
- Use data.table: `tidy[, .(mean_height = ..., sd_height = ..., n = .N), by = Gender]`  

**Step 2b: Create a boxplot**  
- Use `ggplot()` with `tidy` dataset  
- x-axis: `Gender`  
- y-axis: `Height`  
- fill: `Gender`  
- geom: `geom_boxplot(alpha = 0.7)`   
- Add individual points with `geom_jitter(width = 0.1, alpha = 0.2)`  

**Step 2c: Perform two-sample t-test**  
- Use `t.test(Height ~ Gender, data = tidy)`  

```{r final-part2, exercise=TRUE, exercise.setup="shared-data"}
# TASK 2: Your code here




```

---  

**TASK 3: ANOVA - Height by Continent**  

Compare average heights across all continents.  

**Step 3a: Calculate summary statistics**  
- Use data.table: `tidy[, .(mean_height = ..., sd_height = ..., n = .N), by = Continent]`  

**Step 3b: Create a boxplot**  
- Use `ggplot()` with `tidy` dataset  
- x-axis: `Continent`  
- y-axis: `Height`  
- fill: `Continent`  
- geom: `geom_boxplot(alpha = 0.7)`  
- Add points with `geom_jitter(width = 0.2, alpha = 0.3)`  
- Remove legend with `theme(legend.position = "none")`  

**Step 3c: Perform ANOVA**  
- Create ANOVA model: `anova_result <- aov(Height ~ Continent, data = tidy)`  
- View results: `summary(anova_result)`  

**Step 3d: Perform post-hoc test (if significant)**  
- Use `TukeyHSD(anova_result)`  

```{r final-part3, exercise=TRUE, exercise.setup="shared-data"}
# TASK 3: Your code here




```

```{r final-part1-hint}
# TASK 1 HINT: Chi-squared test
# Step 1a: Visualization
ggplot(tidy, aes(x = Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(title = "Gender Distribution by Continent", y = "Proportion")

# Step 1b: Contingency table with .N
tidy[,.N,.(Continent,Gender)]

# Step 1c: Test
chisq.test(tidy[,Continent], tidy[,Gender])
```

```{r final-part2-hint}
# TASK 2 HINT: Two-sample t-test
# Step 2a: Summary
tidy[, .(mean_height = mean(Height), sd_height = sd(Height), n = .N), by = Gender]

# Step 2b: Visualization
ggplot(tidy, aes(x = Gender, y = Height, fill = Gender)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    theme_light() +
    labs(title = "Height by Gender")

# Step 2c: Test
t.test(Height ~ Gender, data = tidy)
```

```{r final-part3-hint}
# TASK 3 HINT: ANOVA
# Step 3a: Summary
tidy[, .(mean_height = mean(Height), sd_height = sd(Height), n = .N), by = Continent]

# Step 3b: Visualization
ggplot(tidy, aes(x = Continent, y = Height, fill = Continent)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.3) +
    theme_light() +
    labs(title = "Height by Continent", y = "Height (cm)") +
    theme(legend.position = "none")

# Step 3c: ANOVA
anova_result <- aov(Height ~ Continent, data = tidy)
summary(anova_result)

# Step 3d: Post-hoc
TukeyHSD(anova_result)
```

```{r final-part3-solution}
# ==================== TASK 1: Categorical Analysis ====================
# Question: Are Gender and Continent independent?

# Step 1a: Visualization - Stacked bar chart
ggplot(tidy, aes(x = Continent, fill = Gender)) +
    geom_bar(position = "fill") +
    theme_light() +
    labs(title = "Gender Distribution by Continent", y = "Proportion")

# Step 1b: Contingency table with .N
tidy[,.N,.(Continent,Gender)]

# Step 1c: Chi-squared test
chisq.test(tidy[,Continent], tidy[,Gender])


# ==================== TASK 2: Height by Gender ====================
# Question: Do males and females have different heights?

# Step 2a: Summary statistics
tidy[, .(mean_height = mean(Height), sd_height = sd(Height), n = .N), by = Gender]

# Step 2b: Visualization - Boxplot
ggplot(tidy, aes(x = Gender, y = Height, fill = Gender)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.1, alpha = 0.2) +
    theme_light() +
    labs(title = "Height by Gender", y = "Height (cm)")

# Step 2c: Two-sample t-test
t.test(Height ~ Gender, data = tidy)


# ==================== TASK 3: Height by Continent ====================
# Question: Do heights differ across continents?

# Step 3a: Summary statistics
tidy[, .(mean_height = mean(Height), sd_height = sd(Height), n = .N), by = Continent]

# Step 3b: Visualization - Boxplot
ggplot(tidy, aes(x = Continent, y = Height, fill = Continent)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.3) +
    theme_light() +
    labs(title = "Height by Continent", y = "Height (cm)") +
    theme(legend.position = "none")

# Step 3c: ANOVA
anova_result <- aov(Height ~ Continent, data = tidy)
summary(anova_result)

# Step 3d: Post-hoc test
TukeyHSD(anova_result)


# ==================== BONUS: Height by Gender AND Continent ====================
# Explore the interaction between Gender and Continent

# Visualization: Two-way comparison
ggplot(tidy, aes(x = Continent, y = Height, fill = Gender)) +
    geom_boxplot(alpha = 0.7) +
    theme_light() +
    labs(title = "Height by Continent and Gender", y = "Height (cm)")

# Two-way ANOVA (tests both main effects and interaction)
anova_2way <- aov(Height ~ Gender * Continent, data = tidy)
summary(anova_2way)
```

---  

## 9. Summary and Quick Reference  

### 9.1. Congratulations!  

You've learned how to combine data manipulation, visualization, and statistical testing!  

### 9.2. What You've Learned  

**Data Manipulation (data.table):**  
- `DT[i, j, by]` syntax for filtering, selecting, computing  
- Grouping with `by`  
- Calculating summaries with `.N`, `mean()`, `sd()`  

**Statistical Tests:**  

| Test | Purpose | When to Use | R Function |  
|------|---------|-------------|------------|  
| **Chi-squared** | Test independence of categorical variables | 2+ categorical variables, n ≥ 5 per cell | `chisq.test()` |  
| **Fisher's exact** | Test independence (small samples) | 2 categorical variables, small n | `fisher.test()` |  
| **One-sample t-test** | Compare mean to known value | 1 continuous variable vs. reference | `t.test(x, mu = value)` |  
| **Two-sample t-test** | Compare 2 independent group means | 2 groups, continuous outcome | `t.test(y ~ group)` |  
| **Paired t-test** | Compare 2 related measurements | Before/after, matched pairs | `t.test(x, y, paired = TRUE)` |  
| **ANOVA** | Compare 3+ group means | 3+ groups, continuous outcome | `aov(y ~ group)` |  
| **Tukey HSD** | Post-hoc after ANOVA | After significant ANOVA | `TukeyHSD()` |  

**Visualization (ggplot2):**  
- Bar charts for categorical data  
- Histograms and density plots for distributions  
- Boxplots and violin plots for group comparisons  
- Always visualize before testing!  

### 9.3. Key Workflow  

1. **Explore**: Use data.table to understand your data  
2. **Visualize**: Create plots with ggplot2  
3. **Test**: Choose and perform appropriate statistical test  
4. **Interpret**: Look at p-values and effect sizes  
5. **Communicate**: Report findings clearly  

### 9.4. Best Practices  

<div class="callout callout-tip">  
<div class="callout-title">Analysis Checklist</div>  

✓ Always visualize your data first  
✓ Check assumptions before testing  
✓ Use appropriate test for your data type  
✓ Report effect sizes, not just p-values  
✓ Consider practical significance, not just statistical significance  
✓ Document your analysis workflow  

</div>  

---  

### 9.5. Download Your Work  

<div class="callout callout-tip">  
<div class="callout-title">Save Your Work</div>  

Save this completed tutorial as an HTML file for future reference.  
</div>  

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---  

### 9.6. Please Give Feedback!  

<div class="callout callout-success">  
<div class="callout-title">💬 Your Feedback Matters</div>  

Help us improve this tutorial! Share your experience and suggestions.  
</div>  

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/F3xueJLoTrVyPPLM7" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```

---  

### 9.7. Additional Resources  

**R Packages:**  
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)  
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)  

**Statistics Resources:**  
- [Quick-R: Basic Statistics](https://www.statmethods.net/stats/index.html)  
- [R for Data Science](https://r4ds.had.co.nz/)  

**Statistical Tests:**  
- [Choosing the Right Test](https://www.scribbr.com/statistics/statistical-tests/)  
- [Statistical Test Decision Tree](https://stats.oarc.ucla.edu/other/mult-pkg/whatstat/)  
