---
title: "Data Visualization with ggplot2"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: "../assets/kuzman_theme.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()

# ggplot2 datasets
tidy <- fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
setnames(works, "GradeNumerical", "Mark")
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))
```

## Welcome

Welcome to **Data Visualization with ggplot2**

### Author

**dr.sc. Maja Kuzman** - Kuzman Consulting d.o.o.

<div class="callout callout-info">
<div class="callout-title">How to Use This Tutorial</div>

- Click **"Run Code"** to execute R code
    - ctrl + enter (windows) or command + enter (macos)
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

</div>

<div class="callout callout-success">
<div class="callout-title">What You Will Learn</div>

In this tutorial, you will master data visualization with ggplot2:

1. **Installation & Setup**  
   - Installing and loading packages  
   - Sample data  
   
2. **ggplot2 vs Base R**  
   - Why ggplot2 is powerful  
   
3. **Basic ggplot2 Logic**  
   - **ggplot() + aes()**: Initialize plots and set scales  
   - **geom_*()**: Add visual layers (organized by variable types)  
   - **facet_*()**: Create subplots  
   - **annotate()**: Add text and labels  
   
4. **Complete Example**  
   - Building a plot from scratch  
   
5. **Customization**  
   - Global customization (entire plot)  
   - Layer-specific customization  
   - Fixed aesthetics  
   
6. **Final Challenge**  
   - End-to-end plot with customization  

7. **Advanced Customizations**  
   - Customizing scales and colors  
   - Theme customization  
   
8. **Saving Plots**  
   - Using ggsave()  
   
9. **Interactive Plots**  
   - Making plots interactive with plotly  

</div>

---

## Installation & Setup

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Installation & Setup**  
   → Installing and loading packages  
   - Sample data  

</div>

### Installing Packages

You only need to install packages ONCE. Use `install.packages()`:

```{r install-packages, echo=TRUE, eval=FALSE}
# Install the required packages
install.packages("data.table")
install.packages("ggplot2")
```

### Loading Libraries

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Installation & Setup**  
   - Installing and loading packages  
   → Sample data  

</div>

Load them every session with `library()`:

```{r load-libs, echo=TRUE, eval=FALSE}
# Load the libraries
library(data.table)
library(ggplot2)
```

### Sample Data

We'll use the built-in `iris` dataset. Preview it:

```{r data-preview, exercise=TRUE}
# Display the iris dataset
iris
```

---

## ggplot2 vs Base R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **ggplot2 vs Base R**  
   → Why ggplot2 is powerful  

</div>

### Base R Approach

Create a scatterplot using base R:

```{r base-plot, exercise=TRUE}
# Base R scatterplot
plot(iris$Sepal.Length, iris$Petal.Length, col = iris$Species)
```

### ggplot2 Approach

The same plot with ggplot2:

```{r ggplot-simple, exercise=TRUE}
# ggplot2 scatterplot
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) + 
    geom_point() +
    theme_light()
```

### The Power of ggplot2

<div class="callout callout-success">
<div class="callout-title">Grammar of Graphics</div>

ggplot2 is based on the **Grammar of Graphics** - a consistent system for building visualizations by combining independent components. Once you learn the logic, you can create almost any plot!  

</div>

Watch how easy it is to enhance the plot. Try running this demo:

```{r ggplot-power, exercise=TRUE}
# Start simple
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) + 
    geom_point() +
    theme_light()

# Now enhance it step by step:
# - Larger points with different shapes
# - Facet by species
# - Add linear regression lines
```

```{r ggplot-power-solution}
# Enhanced version with many features
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species, shape = Species)) + 
    geom_point(size = 3) +
    facet_wrap(~Species) + 
    geom_smooth(method = "lm", linetype = "dashed") +
    theme_bw() +
    labs(title = "Sepal vs Petal Length by Species",
         subtitle = "With linear regression lines",
         x = "Sepal Length (cm)", 
         y = "Petal Length (cm)") +
    annotate("text", x = 6, y = 2, label = "Beautiful!", color = "purple", size = 5)
```

---

## Basic ggplot2 Logic

<div class="callout callout-info">
<div class="callout-title">The ggplot2 Formula</div>

**Core pattern:** 

```r
ggplot(data, aes(x, y, ...)) +    # 1. Initialize + set scales  
    geom_*(...) +                  # 2. Add visual layers  
    facet_*(...) +                 # 3. Create subplots (optional)  
    annotate(...)                  # 4. Add labels (optional)  
```

</div>

---

### 1. ggplot() + aes(): Initialize & Set Scales

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Basic ggplot2 Logic**  
   → **ggplot() + aes()**: Initialize and set scales  

</div>

<div class="callout callout-concept">
<div class="callout-title">What is ggplot()?</div>

`ggplot()` initializes a plot object and sets up the data and aesthetic mappings.  

**Aesthetics (aes)** map your data to visual properties:  
- **x, y**: Position on axes  
- **color**: Color by variable  
- **fill**: Fill color  
- **size**: Size of points/lines  
- **shape**: Point shapes  
- **alpha**: Transparency  

</div>

#### Three Ways to Initialize

You can initialize a ggplot in three ways:

```{r init-demo, exercise=TRUE}
# 1. Empty - no data
p1 <- ggplot()
p1

# 2. With data only
p2 <- ggplot(iris)
p2

# 3. With data + aesthetics
p3 <- ggplot(iris, aes(Sepal.Length, Petal.Length))
p3
```

<div class="callout callout-danger">
<div class="callout-title">Common Mistake!</div>

Calling `ggplot()` without adding a `geom_*()` layer results in a blank plot! You must add at least one geom to see your data.  
</div>

#### Your Turn: Initialize a Plot

Create a ggplot with the `iris` dataset, mapping `Species` to x and `Petal.Length` to y:

```{r init-exercise, exercise=TRUE}
# Initialize with data and aesthetics


```

```{r init-exercise-solution}
p <- ggplot(iris, aes(Species, Petal.Length))
p
```

---

### 2. geom_*(): Add Visual Layers

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Basic ggplot2 Logic**  
   - ggplot() + aes(): Initialize and set scales  
   → **geom_*()**: Add visual layers  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Geoms?</div>

**Geometric objects (geoms)** define how your data is displayed. The choice of geom depends on:  
- How many variables you're plotting (1 or 2)  
- The type of variables (categorical or continuous)  

</div>

<div class="callout callout-info">
<div class="callout-title">Common Geoms Reference</div>

**Geoms (Visual Layers):**

| Function | Use Case | Example |
|----------|----------|--------|
| `geom_point()` | Scatterplot | `geom_point()` |
| `geom_line()` | Line plot | `geom_line()` |
| `geom_bar()` | Bar chart (count) | `geom_bar()` |
| `geom_col()` | Bar chart (values) | `geom_col()` |
| `geom_histogram()` | Histogram | `geom_histogram(bins = 30)` |
| `geom_density()` | Density plot | `geom_density(alpha = 0.5)` |
| `geom_boxplot()` | Box plot | `geom_boxplot()` |
| `geom_violin()` | Violin plot | `geom_violin()` |
| `geom_smooth()` | Smoothed line/trend | `geom_smooth(method = "lm")` |
| `geom_jitter()` | Jittered points | `geom_jitter(width = 0.2)` |

</div>

#### Geoms for ONE Variable

**Categorical variable → Bar chart**

```{r geom-bar-demo, exercise=TRUE}
# Bar chart of Species
ggplot(iris, aes(Species)) +
    geom_bar() +
    theme_light()
```

**Continuous variable → Histogram**

```{r geom-hist-demo, exercise=TRUE, exercise.setup="shared-data"}
# Histogram of Height
ggplot(tidy, aes(Height)) +
    geom_histogram(bins = 30, alpha = 0.7, color = "black") +
    theme_light()
```

**Continuous variable → Density plot**

```{r geom-density-demo, exercise=TRUE, exercise.setup="shared-data"}
# Density plot of Height by Gender
ggplot(tidy, aes(Height, fill = Gender)) +
    geom_density(alpha = 0.5) +
    theme_light()
```

#### Your Turn: One Variable

Create a bar chart of `Continent` from the `tidy` dataset:

```{r one-var-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Create a bar chart


```

```{r one-var-exercise-solution}
ggplot(tidy, aes(Continent)) +
    geom_bar() +
    theme_light()
```

---

#### Geoms for TWO Variables

The choice depends on variable types:

**Continuous X, Continuous Y → Scatterplot**

```{r geom-point-demo, exercise=TRUE}
# Scatterplot
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point() +
    theme_light()
```

**Continuous X, Continuous Y → Scatterplot + Smooth line**

```{r geom-smooth-demo, exercise=TRUE}
# Scatterplot with trend line
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_light()
```

**Discrete X, Continuous Y → Boxplot**

```{r geom-box-demo, exercise=TRUE}
# Boxplot by Species
ggplot(iris, aes(Species, Petal.Length)) +
    geom_boxplot() +
    theme_light()
```

**Discrete X, Continuous Y → Violin plot**

```{r geom-violin-demo, exercise=TRUE}
# Violin plot by Species
ggplot(iris, aes(Species, Petal.Length)) +
    geom_violin() +
    theme_light()
```

**Discrete X, Discrete Y → Bar chart (with fill)**

```{r geom-bar-fill-demo, exercise=TRUE, exercise.setup="shared-data"}
# Stacked bar chart
ggplot(tidy, aes(Gender, fill = Continent)) +
    geom_bar() +
    theme_light()
```

#### Your Turn: Two Variables

Create a scatterplot showing the relationship between `Height` and `Weight` from the `tidy` dataset:

```{r two-var-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Create a scatterplot


```

```{r two-var-exercise-solution}
ggplot(tidy, aes(Height, Weight)) +
    geom_point() +
    theme_light()
```

---

### 3. facet_*(): Create Subplots

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Basic ggplot2 Logic**  
   - ggplot() + aes(): Initialize and set scales  
   - geom_*(): Add visual layers  
   → **facet_*()**: Create subplots  

</div>

<div class="callout callout-concept">
<div class="callout-title">What is Faceting?</div>

**Faceting** creates multiple plots (subplots) based on a categorical variable. Perfect for comparing patterns across groups!

- `facet_wrap(~variable)` - wraps panels into a grid  
- `facet_grid(row~col)` - arranges panels in rows and columns  

</div>

#### facet_wrap() Example

```{r facet-wrap-demo, exercise=TRUE}
# Separate plots for each species
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point(aes(color = Species)) +
    facet_wrap(~Species) +
    theme_light()
```

#### facet_grid() Example

```{r facet-grid-demo, exercise=TRUE, exercise.setup="shared-data"}
# Grid of plots by Gender and Continent
ggplot(tidy, aes(Height, Weight)) +
    geom_point(alpha = 0.3) +
    facet_grid(Gender ~ Continent) +
    theme_light()
```

#### Your Turn: Faceting

Create a histogram of `Petal.Length` faceted by `Species`:

```{r facet-exercise, exercise=TRUE}
# Create faceted histogram


```

```{r facet-exercise-solution}
ggplot(iris, aes(Petal.Length)) +
    geom_histogram(bins = 20, alpha = 0.7) +
    facet_wrap(~Species) +
    theme_light()
```

---

### 4. annotate(): Add Text and Labels

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Basic ggplot2 Logic**  
   - ggplot() + aes(): Initialize and set scales    
   - geom_*(): Add visual layers  
   - facet_*(): Create subplots  
   → **annotate()**: Add text and labels  

</div>

<div class="callout callout-concept">
<div class="callout-title">What is annotate()?</div>

`annotate()` adds custom text, shapes, or lines to your plot. Useful for:  
- Highlighting specific points or regions  
- Adding explanatory text  
- Drawing attention to patterns  

</div>

#### Adding Text Annotations

```{r annotate-demo, exercise=TRUE}
# Scatterplot with annotation
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point() +
    annotate("text", x = 7, y = 2, 
             label = "Interesting pattern!", 
             color = "red", size = 5) +
    theme_light()
```

#### Your Turn: Annotate

Add an annotation to this plot at x=5.5, y=6 that says "High values":

```{r annotate-exercise, exercise=TRUE}
# Add annotation
ggplot(iris, aes(Petal.Length, Sepal.Length)) +
    geom_point() +
    theme_light()
```

```{r annotate-exercise-solution}
ggplot(iris, aes(Petal.Length, Sepal.Length)) +
    geom_point() +
    annotate("text", x = 5.5, y = 6, label = "High values", color = "blue") +
    theme_light()
```

---

## Complete Example: Build from Scratch

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Complete Example**  
   → Building a plot from scratch    

</div>

Now put it all together! Create a plot with the `iris` dataset that:

1. Maps `Petal.Length` to x-axis and `Sepal.Length` to y-axis  
2. Colors points by `Species`  
3. Adds points with `geom_point()`  
4. Adds a smooth line with `geom_smooth(method = "lm")`  
5. Facets by `Species`    
6. Adds an annotation of your choice  
7. Uses `theme_bw()`  

```{r complete-example, exercise=TRUE}
# Build your complete plot here


```

```{r complete-example-hint}
# Start with ggplot and aes
ggplot(iris, aes(Petal.Length, Sepal.Length, color = Species)) +
    # Add your layers here
```

```{r complete-example-solution}
ggplot(iris, aes(Petal.Length, Sepal.Length, color = Species)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~Species) +
    annotate("text", x = 5, y = 7, label = "Linear trend", size = 3) +
    theme_bw() +
    labs(title = "Petal vs Sepal Length",
         x = "Petal Length (cm)",
         y = "Sepal Length (cm)")
```

---

## Customization

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Customization**  
   → Global, layer-specific, and fixed aesthetics  

</div>

<div class="callout callout-concept">
<div class="callout-title">Three Ways to Customize</div>

**1. Global (in ggplot aes)** - applies to all layers    
**2. Layer-specific (in geom aes)** - applies to one geom only    
**3. Fixed aesthetics (outside aes)** - constant value for all points    

</div>

### 1. Global Customization

Set aesthetics in `ggplot(aes(...))` - applies to all layers:

```{r custom-global, exercise=TRUE}
# Color applies to both geom_point and geom_smooth
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_light()
```

### 2. Layer-Specific Customization

Set aesthetics in `geom_*(aes(...))` - applies only to that layer:

```{r custom-layer, exercise=TRUE}
# Color only applies to points, not smooth line
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point(aes(color = Species)) +
    geom_smooth(method = "lm", color = "black") +
    theme_light()
```

### 3. Fixed Aesthetics (Outside aes)

Set a constant value outside `aes()` - same for all points:

```{r custom-fixed, exercise=TRUE}
# All points are blue, size 3, 50% transparent
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point(color = "blue", size = 3, alpha = 0.5) +
    theme_light()
```

### Combining All Three

```{r custom-combined, exercise=TRUE}
# Global: x and y mapping
# Layer 1 (points): color by Species, fixed size and alpha
# Layer 2 (smooth): fixed color black
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point(aes(color = Species), size = 3, alpha = 0.6) +
    geom_smooth(method = "lm", color = "black", linetype = "dashed") +
    theme_light()
```

### Your Turn: Customize

Create a plot with `iris` where:v
- x = `Petal.Length`, y = `Petal.Width`  
- Points colored by `Species` (layer-specific)  
- Points have fixed size = 4 and alpha = 0.7  
- Add a black smooth line with `se = FALSE`  
- Use `theme_minimal()`  

```{r custom-exercise, exercise=TRUE}
# Create your customized plot


```

```{r custom-exercise-solution}
ggplot(iris, aes(Petal.Length, Petal.Width)) +
    geom_point(aes(color = Species), size = 4, alpha = 0.7) +
    geom_smooth(method = "lm", color = "black", se = FALSE) +
    theme_minimal()
```

---

## Final Challenge: End-to-End

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

6. **Final Challenge**  
   → End-to-end plot with customization    

</div>

### Challenge: Complete Visualization

Using the `works` dataset, create a complete visualization that shows the relationship between `Grade` and `Mark`.  

**Requirements:**

1. Map `Grade` to x-axis and `Mark` to y-axis  
2. Use `geom_jitter(width = 0.2)` to avoid overplotting  
3. Color points by `Grade` (layer-specific)  
4. Make points semi-transparent (alpha = 0.7)  
5. Add a boxplot layer with `alpha = 0.3`  
6. Use `theme_light()`  
7. Remove the legend with `theme(legend.position = "none")`  
8. Add appropriate labels with `labs()`  

First, preview the data:

```{r challenge-data, exercise=TRUE, exercise.setup="shared-data"}
# View the works dataset
works
```

Now create your plot:

```{r challenge-final, exercise=TRUE, exercise.setup="shared-data"}
# Create your complete visualization


```

```{r challenge-final-hint}
ggplot(works, aes(Grade, Mark)) +
    # Add jittered points with color
    # Add boxplot layer
    # Customize theme and labels
```

```{r challenge-final-solution}
ggplot(works, aes(Grade, Mark)) +
    geom_jitter(aes(color = Grade), width = 0.2, alpha = 0.7) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA) +
    theme_light() +
    theme(legend.position = "none") +
    labs(
        title = "Grade vs Mark Distribution",
        x = "Grade",
        y = "Mark"
    )
```

---

## Advanced Customizations

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

7. **Advanced Customizations**  
   → Customizing scales and colors  
   - Theme customization  

</div>

### Customizing Scales and Colors

<div class="callout callout-concept">
<div class="callout-title">What are Scales?</div>

**Scales** control how data values are mapped to visual properties. You can customize:
- **Color scales**: `scale_color_*()`, `scale_fill_*()`
- **Axis scales**: `scale_x_*()`, `scale_y_*()`
- **Size scales**: `scale_size_*()`

</div>

#### Manual Color Scales

```{r scale-manual-demo, exercise=TRUE}
# Custom colors for each species
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point(size = 3) +
    scale_color_manual(values = c("red", "blue", "green")) +
    theme_light()
```

#### Predefined Color Palettes

```{r scale-brewer-demo, exercise=TRUE}
# Using ColorBrewer palettes
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point(size = 3) +
    scale_color_brewer(palette = "Set1") +
    theme_light()
```

#### Continuous Color Scales

```{r scale-gradient-demo, exercise=TRUE}
# Gradient for continuous variables
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Petal.Width)) +
    geom_point(size = 3) +
    scale_color_gradient(low = "yellow", high = "red") +
    theme_light()
```

#### Customizing Axes

```{r scale-axes-demo, exercise=TRUE}
# Custom axis limits and breaks
ggplot(iris, aes(Sepal.Length, Petal.Length)) +
    geom_point() +
    scale_x_continuous(limits = c(4, 8), breaks = seq(4, 8, by = 0.5)) +
    scale_y_continuous(limits = c(0, 7), breaks = seq(0, 7, by = 1)) +
    theme_light()
```

#### Your Turn: Custom Colors

Create a boxplot of `Petal.Length` by `Species` with custom fill colors (your choice of colors):

```{r scale-exercise, exercise=TRUE}
# Create a boxplot with custom colors


```

```{r scale-exercise-solution}
ggplot(iris, aes(Species, Petal.Length, fill = Species)) +
    geom_boxplot(alpha = 0.7) +
    scale_fill_manual(values = c("purple", "orange", "cyan")) +
    theme_light() +
    theme(legend.position = "none")
```

---

### Theme Customization

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

7. **Advanced Customizations**  
   - Customizing scales and colors  
   → Theme customization  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Themes?</div>

**Themes** control the non-data elements of your plot:
- Background colors
- Grid lines
- Text size and fonts
- Legend position and appearance
- Axis text and titles

</div>

#### Built-in Themes

```{r theme-builtin-demo, exercise=TRUE}
# Try different built-in themes
p <- ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point(size = 2)

# theme_minimal()
p + theme_minimal()

# theme_classic()
# p + theme_classic()

# theme_dark()
# p + theme_dark()
```

#### Customizing Theme Elements

```{r theme-custom-demo, exercise=TRUE}
# Customize specific theme elements
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point(size = 3) +
    theme_light() +
    theme(
        legend.position = "bottom",
        legend.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 16, hjust = 0.5),
        panel.grid.minor = element_blank()
    ) +
    labs(title = "Iris Dataset Analysis")
```

#### Complete Customization Example

```{r theme-complete-demo, exercise=TRUE}
# Highly customized plot
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point(size = 3, alpha = 0.7) +
    scale_color_brewer(palette = "Dark2") +
    labs(
        title = "Sepal vs Petal Length",
        subtitle = "Across three Iris species",
        x = "Sepal Length (cm)",
        y = "Petal Length (cm)",
        color = "Species",
        caption = "Source: iris dataset"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 12, hjust = 0.5),
        legend.position = "top",
        panel.grid.minor = element_blank()
    )
```

#### Your Turn: Complete Theme

Create a customized plot with the `tidy` dataset showing `Height` vs `Weight`, colored by `Gender`:
- Use custom colors
- Add appropriate labels
- Customize the theme
- Place legend at bottom

```{r theme-exercise, exercise=TRUE, exercise.setup="shared-data"}
# Create your fully customized plot


```

```{r theme-exercise-solution}
ggplot(tidy, aes(Height, Weight, color = Gender)) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c("steelblue", "coral")) +
    labs(
        title = "Height vs Weight by Gender",
        x = "Height (cm)",
        y = "Weight (kg)"
    ) +
    theme_bw() +
    theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", hjust = 0.5)
    )
```

---

## Saving Plots

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

8. **Saving Plots**  
   → Using ggsave()  

</div>

<div class="callout callout-concept">
<div class="callout-title">Saving with ggsave()</div>

`ggsave()` saves the last plot you created. You can:
- Change the file format by changing the extension (.pdf, .png, .jpg, .svg)
- Control dimensions with `width` and `height`
- Adjust resolution with `dpi`

</div>

### Example: Saving Plots

```{r ggsave-demo, echo=TRUE, eval=FALSE}
# Create a plot
plot2 <- ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point() +
    theme_light()

# Display it
plot2

# Save as PDF
ggsave("plot2.pdf", width = 8, height = 6)

# Save as PNG with high resolution
ggsave("plot2.png", width = 8, height = 6, dpi = 300)

# Save specific plot (not just the last one)
ggsave("plot2_custom.pdf", plot = plot2, width = 10, height = 7)
```

**Common formats:**
- `.pdf` - Vector format, scalable
- `.png` - Raster format, good for web
- `.jpg` - Raster format, compressed
- `.svg` - Vector format, web-friendly

---

## Interactive Plots

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

9. **Interactive Plots**  
   → Making plots interactive with plotly  

</div>

<div class="callout callout-concept">
<div class="callout-title">What is plotly?</div>

**plotly** converts your ggplot2 plots into interactive visualizations! Users can:
- Hover over points to see values
- Zoom in and out
- Pan around the plot
- Toggle legend items on/off

Simply wrap your ggplot with `ggplotly()`!

</div>

### Make Plots Interactive

```{r plotly-demo, exercise=TRUE, eval=FALSE}
# Load plotly
library(plotly)

# Create a ggplot
plot2 <- ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point() +
    theme_light()

# Make it interactive
ggplotly(plot2)
```

Try hovering over points to see the values!

### Your Turn: Interactive Plot

Create an interactive scatterplot with the `tidy` dataset showing `Height` vs `Weight`:

```{r plotly-exercise, exercise=TRUE, exercise.setup="shared-data", eval=FALSE}
# Create an interactive plot


```

```{r plotly-exercise-solution, eval=FALSE}
library(plotly)

p <- ggplot(tidy, aes(Height, Weight, color = Gender)) +
    geom_point(alpha = 0.5) +
    theme_light()

ggplotly(p)
```

---

## Quick Reference: All Functions

<div class="callout callout-info">
<div class="callout-title">Functions Used in This Tutorial</div>

**Core Functions:**

| Function | Description | Example |
|----------|-------------|--------|
| `ggplot()` | Initialize a plot | `ggplot(data, aes(x, y))` |
| `aes()` | Map aesthetics | `aes(x = col1, y = col2, color = col3)` |

**Geoms (Visual Layers):**

| Function | Use Case | Example |
|----------|----------|--------|
| `geom_point()` | Scatterplot | `geom_point()` |
| `geom_line()` | Line plot | `geom_line()` |
| `geom_bar()` | Bar chart (count) | `geom_bar()` |
| `geom_col()` | Bar chart (values) | `geom_col()` |
| `geom_histogram()` | Histogram | `geom_histogram(bins = 30)` |
| `geom_density()` | Density plot | `geom_density(alpha = 0.5)` |
| `geom_boxplot()` | Box plot | `geom_boxplot()` |
| `geom_violin()` | Violin plot | `geom_violin()` |
| `geom_smooth()` | Smoothed line/trend | `geom_smooth(method = "lm")` |
| `geom_jitter()` | Jittered points | `geom_jitter(width = 0.2)` |

**Faceting:**

| Function | Description | Example |
|----------|-------------|--------|
| `facet_wrap()` | Wrap facets into grid | `facet_wrap(~Species)` |
| `facet_grid()` | Arrange in rows/columns | `facet_grid(row~col)` |

**Scales:**

| Function | Description | Example |
|----------|-------------|--------|
| `scale_color_manual()` | Manual colors | `scale_color_manual(values = c("red", "blue"))` |
| `scale_fill_manual()` | Manual fill colors | `scale_fill_manual(values = c("#FF0000", "#0000FF"))` |
| `scale_color_brewer()` | ColorBrewer palettes | `scale_color_brewer(palette = "Set1")` |
| `scale_color_gradient()` | Color gradient | `scale_color_gradient(low = "white", high = "red")` |
| `scale_x_continuous()` | Continuous x-axis | `scale_x_continuous(limits = c(0, 10))` |
| `scale_y_continuous()` | Continuous y-axis | `scale_y_continuous(breaks = seq(0, 100, 10))` |

**Themes:**

| Function | Description | Example |
|----------|-------------|--------|
| `theme_light()` | Light theme | `theme_light()` |
| `theme_minimal()` | Minimal theme | `theme_minimal()` |
| `theme_classic()` | Classic theme | `theme_classic()` |
| `theme_bw()` | Black and white | `theme_bw()` |
| `theme_dark()` | Dark theme | `theme_dark()` |
| `theme()` | Customize theme | `theme(legend.position = "bottom")` |

**Labels:**

| Function | Description | Example |
|----------|-------------|--------|
| `labs()` | Add labels | `labs(title = "My Plot", x = "X axis", y = "Y axis")` |
| `xlab()` | X-axis label | `xlab("X variable")` |
| `ylab()` | Y-axis label | `ylab("Y variable")` |
| `ggtitle()` | Plot title | `ggtitle("My Title")` |

**Annotations:**

| Function | Description | Example |
|----------|-------------|--------|
| `annotate()` | Add text/shapes | `annotate("text", x = 5, y = 10, label = "Note")` |

**Other Functions:**

| Function | Description | Example |
|----------|-------------|--------|
| `ggsave()` | Save plot to file | `ggsave("plot.png", width = 8, height = 6)` |
| `ggplotly()` | Make plot interactive | `ggplotly(plot)` (from plotly package) |
| `coord_flip()` | Flip coordinates | `coord_flip()` |

**Common Aesthetic Mappings:**

| Aesthetic | Description | Example |
|-----------|-------------|---------|
| `x` | X-axis position | `aes(x = height)` |
| `y` | Y-axis position | `aes(y = weight)` |
| `color` | Color of points/lines | `aes(color = species)` |
| `fill` | Fill color | `aes(fill = category)` |
| `size` | Size of points/lines | `aes(size = population)` |
| `alpha` | Transparency | `aes(alpha = 0.5)` |
| `shape` | Point shape | `aes(shape = type)` |
| `linetype` | Line type | `aes(linetype = group)` |

</div>

---

## Congratulations!

You've completed the ggplot2 tutorial!

### What You've Learned:

- Why ggplot2 is powerful (Grammar of Graphics)
- **ggplot() + aes()**: Initializing plots and setting scales
- **geom_*()**: Visual layers for different variable types
  - One variable: bar, histogram, density
  - Two variables: point, smooth, boxplot, violin, bar
- **facet_*()**: Creating subplots
- **annotate()**: Adding text and labels
- **Customization**: Global, layer-specific, and fixed aesthetics
- **Advanced customizations**: Scales, colors, and themes
- **Saving plots**: Using ggsave()
- **Interactive plots**: Using plotly

### Next Steps:

In the next tutorial we will get deeper into R - with more practical examples and using tables (data.table) !


---

### Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

### Please Give Feedback!

<div class="callout callout-success">
<div class="callout-title">💬 Your Feedback Matters</div>

Help us improve this tutorial! Share your experience and suggestions.
</div>

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/B7Wy7G29wQTuEKjU9" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```

---

### Quick Reference

<div class="callout callout-info">
<div class="callout-title">ggplot2 Cheat Sheet</div>

**Basic Structure:**
```r
ggplot(data, aes(x, y, ...)) + 
    geom_*(...) + 
    facet_*(...) +
    labs(...) +
    theme_*()
```

**Common Geoms:**
- `geom_point()` - scatter plot
- `geom_line()` - line plot
- `geom_bar()` - bar chart
- `geom_histogram()` - histogram
- `geom_boxplot()` - box plot
- `geom_violin()` - violin plot
- `geom_smooth()` - smooth line

**Common Aesthetics:**
- Position: `x`, `y`
- Color: `color`, `fill`
- Shape: `shape`, `linetype`
- Size: `size`, `alpha`

</div>

### Additional Resources

- [ggplot2 Documentation](https://ggplot2.tidyverse.org/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [ggplot2 Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)
