---
title: "Data Visualization with ggplot2"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Server setup (empty - download handled by JavaScript)
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
tidy <- fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))
```

## Welcome

Welcome to **Data Visualization with ggplot2**

### Author

**dr.sc. Maja Kuzman** - Kuzman Consulting d.o.o.

<div class="callout callout-info">
<div class="callout-title">How to Use This Tutorial</div>

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises
</div>

<div class="callout callout-success">
<div class="callout-title">What You'll Learn</div>

- Creating beautiful visualizations with the Grammar of Graphics  
- Mastering ggplot2's layered approach  
- Customizing plots with themes, colors, and facets  
- Making publication-ready graphics
</div>

---

## Installation

### Installing Packages

You only need to install packages ONCE:

```{r install-packages, exercise=TRUE, eval=FALSE}
install.packages("data.table")
install.packages("ggplot2")
```

### Loading Libraries

Load them every session:

```{r load-libs, exercise=TRUE}
library(data.table)
library(ggplot2)
```

### Sample Data

```{r data-preview, exercise=TRUE, exercise.setup="shared-data"}
tidy
```

---

## Variable Types

<div class="callout callout-concept">
<div class="callout-title">Choosing the Right Visualization</div>

The type of visualization you choose depends on your **variable types**. Categorical variables work best with bar charts or pie charts, while continuous variables are better suited for histograms, scatter plots, or line graphs.
</div>

### Categorical Variables

- **Nominal** - no order (e.g., colors, continents)
- **Ordinal** - ordered (e.g., grades, ratings)

### Quantitative Variables

- **Discrete** - countable (e.g., number of children)
- **Continuous** - measurable (e.g., height, weight)

---

## Graphs for One Variable

### Categorical Variable: Bar Chart

```{r bar-demo, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Continent)) +
    geom_bar() +
    theme_light()
```

### Quantitative Variable: Density Plot

```{r density-demo, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Height, fill = Gender)) +
    geom_density(alpha = 0.3) +
    theme_light()
```

---

## Graphs with Two Variables

### Continuous X, Continuous Y: Scatterplot

```{r scatter-demo, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Height, Weight)) +
    geom_point() +
    theme_light()
```

---

## Discrete X, Continuous Y

### Boxplot

```{r boxplot-demo, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Gender, Height, fill = Gender)) +
    geom_boxplot(alpha = 0.3) +
    theme_light()
```

### Violin Plot

```{r violin-demo, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Gender, Height, fill = Gender)) +
    geom_violin(alpha = 0.3) +
    theme_light()
```

---

## Discrete X, Discrete Y: Bar Chart

### Simple Bar Chart

```{r bar-simple, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Gender)) +
    geom_bar() +
    theme_light()
```

### Stacked Bar Chart

```{r bar-stacked, exercise=TRUE, exercise.setup="shared-data"}
ggplot(tidy, aes(Gender, Height, fill = Continent)) +
    geom_bar(stat = "identity") +
    theme_light()
```

---

## ggplot2 vs Base R

### Base R Plot

```{r base-plot, exercise=TRUE}
plot(iris$Sepal.Length, iris$Petal.Length, col = iris$Species)
```

### ggplot2 Plot

```{r ggplot-plot, exercise=TRUE}
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) + 
    geom_point() +
    theme_light()
```

---

## Why ggplot2?

<div class="callout callout-success">
<div class="callout-title">The Power of ggplot2</div>

ggplot2 is based on the **Grammar of Graphics** - a consistent system for building any visualization by combining independent components. Once you learn the logic, you can create almost any plot!
</div>

### Advantages

- Elegant  
- Highly customizable  
- Uniform syntax  
- Natural and expressive  
- Very popular

### Disadvantages

- Steep learning curve  
- Can be slow with large data  
- Evolving rapidly

---

## Basic ggplot Logic

<div class="callout callout-info">
<div class="callout-title">The ggplot2 Formula</div>

**Core pattern:** `ggplot(data, aes(x, y)) + geom_*()`

- **ggplot()** = initialize the plot with data
- **aes()** = aesthetic mappings (which variables go where)
- **geom_*** = geometric objects (points, bars, lines, etc.)
- **+** = layer components together
</div>

### ggplot(data, aes(x, y))

Three ways to initialize:

```{r ggplot-init, exercise=TRUE, eval=FALSE}
p <- ggplot()
p <- ggplot(data_frame)
p <- ggplot(data_frame, aes(x, y))
```

<div class="callout callout-danger">
<div class="callout-title">Common Mistake!</div>

Calling `ggplot()` without adding a geom layer results in a **blank plot or error**. You must add at least one `geom_*()` to see anything!
</div>

**Note:** You always need at least one layer (geom) for a plot to display!

---

## Mapping Aesthetics

<div class="callout callout-tip">
<div class="callout-title">Pro Tip: Flexible Aesthetics</div>

Aesthetics can be defined:
- **Globally** in `ggplot(aes(...))` - applies to all layers
- **Per layer** in `geom_*(aes(...))` - applies only to that layer

This flexibility lets you create complex, multi-layered visualizations!
</div>

### The aes() Function

- **HOW** data is stored
- **HOW** data is split

```{r aes-demo, exercise=TRUE}
p <- ggplot(iris, aes(Species, Petal.Length))
p  # This shows only the empty plot structure
```

---

## Layers: The Building Blocks

<div class="callout callout-success">
<div class="callout-title">The Power of Layers</div>

ggplot2 uses **`+`** to add layers, building plots like a cake:
1. Start with data and aesthetics (the base)
2. Add geoms (the filling)
3. Add scales, themes, facets (the frosting)

Each layer adds something new without replacing what's there!
</div>

### Adding Layers with +

We use `+` to add layers:

- **geoms** - geometric objects (what the data looks like)
- **scales** - control axes and colors
- **facets** - create subplots
- **stats** - statistical transformations

---

## Geometric Objects (geoms)

<div class="callout callout-info">
<div class="callout-title">Common Geoms Reference</div>

**Continuous data:**
- `geom_point()` - scatter plots
- `geom_line()` - line graphs
- `geom_smooth()` - trend lines

**Categorical data:**
- `geom_bar()` / `geom_col()` - bar charts
- `geom_boxplot()` - box plots
- `geom_violin()` - violin plots

**Distributions:**
- `geom_histogram()` - histograms
- `geom_density()` - density curves
</div>

Common geoms:

- `geom_point()` - scatter plot
- `geom_line()` - line plot
- `geom_bar()` - bar chart
- `geom_boxplot()` - box plot
- `geom_smooth()` - smoothed line

---

## Example 1: Step-by-Step

### Step 1: Set Basic Layout

X axis is categorical, Y axis is numerical.

Fill the empty spaces to use `iris` dataset with Species on x-axis and Petal.Length on y-axis:

```{r example1-step1, exercise=TRUE}
p <- ggplot(iris, aes(Species, Petal.Length))
p
```

What does `p` look like? (Just the axes!)

---

## Example 1: Add Points

### Step 2: Add Scatter Plot Layer

```{r example1-step2, exercise=TRUE}
p <- ggplot(iris, aes(Species, Petal.Length))
p + geom_point()
```

---

## Example 1: Color and Boxplot

### Step 3: Color Points by Petal.Length

```{r example1-step3, exercise=TRUE}
p <- ggplot(iris, aes(Species, Petal.Length))
p + geom_point(aes(color = Petal.Length))
```

### Step 4: Add Boxplot Layer

```{r example1-step4, exercise=TRUE}
p <- ggplot(iris, aes(Species, Petal.Length))
p + geom_point(aes(color = Petal.Length)) +
    geom_boxplot(alpha = 0.3)
```

---

## Exercise 1: Build Your Own Plot

Create a plot showing the relationship between Species (x) and Petal.Length (y):

1. Set the basic layout
2. Add points
3. Color points by Petal.Length
4. Add boxplot

```{r exercise1, exercise=TRUE}
# Your code here


```

```{r exercise1-solution}
p <- ggplot(iris, aes(Species, Petal.Length))
p + geom_point(aes(color = Petal.Length)) +
    geom_boxplot(alpha = 0.3)
```

---

## Example 2: Numerical vs Numerical

### Step 1: Set Layout

Both axes are numerical. Looking for connection between Petal.Length and Sepal.Length:

```{r example2-step1, exercise=TRUE}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p2
```

### Step 2: Add Points

```{r example2-step2, exercise=TRUE}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p2 + geom_point()
```

---

## Example 2: Color by Species

### Step 3: Override Aesthetics

Color points by Species (set inside `geom_point`):

```{r example2-step3, exercise=TRUE}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p3 <- p2 + geom_point(aes(color = Species))
p3
```

---

## Example 2: Add Smoothing Lines

### Step 4: Add Linear Regression

Add `geom_smooth()` grouped by Species:

```{r example2-step4, exercise=TRUE}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p3 <- p2 + geom_point(aes(color = Species))
p4 <- p3 + geom_smooth(aes(group = Species), method = "lm")
p4
```

---

## Example 2: Faceting

<div class="callout callout-tip">
<div class="callout-title">Creating Subplots with Faceting</div>

**Faceting** creates multiple plots based on a categorical variable. Use `facet_wrap(~variable)` or `facet_grid(row~col)` to split your visualization - perfect for comparing patterns across groups!
</div>

### Step 5: Separate by Species

Use `facet_wrap()` to create separate plots:

```{r example2-step5, exercise=TRUE}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p3 <- p2 + geom_point(aes(color = Species))
p4 <- p3 + geom_smooth(aes(group = Species), method = "lm")
p4 + facet_wrap(~Species)
```

---

## Exercise 2: Complete Analysis

Build the complete plot step by step:

1. Set layout with Petal.Length (x) and Sepal.Length (y)
2. Add colored points by Species
3. Add smooth lines grouped by Species
4. Facet by Species

```{r exercise2, exercise=TRUE}
# Your code here


```

```{r exercise2-solution}
p2 <- ggplot(iris, aes(Petal.Length, Sepal.Length))
p2 + geom_point(aes(color = Species)) +
    geom_smooth(aes(group = Species), method = "lm") +
    facet_wrap(~Species)
```

---

## Challenge 1: Recreate Plot

### Works Dataset

```{r challenge1-data, exercise=TRUE, exercise.setup="shared-data"}
works
```

**Your Task:** Create a plot showing the relationship between Grade and Mark.

**Hint:** Use `geom_jitter()` to add random noise to points!

```{r challenge1, exercise=TRUE, exercise.setup="shared-data"}
# Your code here


```

```{r challenge1-hint}
# Use geom_jitter() instead of geom_point()
# Try different colors by Grade
ggplot(works, aes(Grade, Mark, color = Grade)) +
    geom_jitter(width = 0.2) +
    theme_light()
```

```{r challenge1-solution}
ggplot(works, aes(Grade, Mark, color = Grade)) +
    geom_jitter(width = 0.2, alpha =0.7) +
    theme_light() +
    theme(legend.position = "none")
```

---

## Challenge 2: Advanced Bar Plot

### Recommendation Dataset

```{r challenge2-data, exercise=TRUE, exercise.setup="shared-data"}
df
table(df$recommend)
```

**Your Task:** Create a horizontal bar plot with custom colors.

**Colors:** `c("#d7191c", "#fdae61", "#ffffbf", "#a6d96a", "#1a9641")`

**Hints:**
- Use `scale_x_discrete()` to set x-axis limits
- Use `scale_fill_manual()` for custom colors
- Try `coord_flip()` for horizontal bars

```{r challenge2, exercise=TRUE, exercise.setup="shared-data"}
# Your code here


```

```{r challenge2-hint}
# Start with a basic bar plot
ggplot(df, aes(recommend, fill = recommend)) +
    geom_bar() +
    scale_fill_manual(
        values = c("#d7191c", "#fdae61", "#ffffbf", "#a6d96a", "#1a9641")
    ) +
    coord_flip() +
    theme_light()
```

```{r challenge2-solution}
ggplot(df, aes(recommend, fill = recommend)) +
    geom_bar() +
    scale_fill_manual(
        breaks = c("Strongly no", "No", "Maybe", "Yes", "Strongly yes"),
        values = c("#d7191c", "#fdae61", "#ffffbf", "#a6d96a", "#1a9641")
    ) +
    coord_flip() +
    theme_light() +
    labs(x = "Recommendation", y = "Count") +
    theme(legend.position = "none")
```

---

## Interactive Plots with plotly

### Make Plots Interactive

```{r plotly-demo, exercise=TRUE, eval=FALSE}
library(plotly)

plot2 <- ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
    geom_point() +
    theme_light()

ggplotly(plot2)
```

Try hovering over points to see the values!

---

## Congratulations!

You've completed the ggplot2 tutorial!

### What You've Learned:

- Variable types and appropriate visualizations  
- Basic ggplot2 syntax: `ggplot(data, aes(x, y)) + geom_*()`  
- Common geoms: point, bar, boxplot, violin, smooth  
- Layering multiple geoms  
- Faceting for subplots  
- Customizing colors and themes  
- Creating interactive plots with plotly

### Next Steps:

Practice with your own datasets and explore more ggplot2 features!

---

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

## Additional Resources

- [ggplot2 Documentation](https://ggplot2.tidyverse.org/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [ggplot2 Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)
- [plotly for R](https://plotly.com/r/)
