---
title: "Introduction to R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
gradethis::gradethis_setup()
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Server setup (empty - download handled by JavaScript)
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Shared data for exercises
myFirstNumber <- 0.1
myFirstVector <- c("some words","p","awesome", "last one")
myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
```

## Welcome

Welcome to **Introduction to R**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## Getting Started with R

<div class="callout callout-info">
<div class="callout-title">R as a Calculator</div>

R can perform mathematical operations just like a calculator. This is your first step in learning R programming!
</div>

### How can we use R and RStudio?

- Console
- Script
- Notebook (add new chunk: Ctrl + Alt + I)

```{r calculator-demo, exercise=TRUE}
1 + 1

```

### Try it!

<div class="callout callout-tip">
<div class="callout-title">Understanding the %% Operator</div>

The `%%` operator performs modulo division, which returns the **remainder** after division. For example, `13 %% 3` returns `1` because 13 divided by 3 is 4 with a remainder of 1.
</div>

Add your calculation here. Calculate the result of 13%%3, 12%%3, 15%%3.  
Can you guess what %% does?

```{r modulo-exercise, exercise=TRUE}
# Calculate modulo operations
13 %% 3

```

```{r modulo-exercise-hint}
# Try calculating:
# 12 %% 3
# 15 %% 3
# The %% operator gives you the remainder after division
```

```{r modulo-exercise-solution}
13 %% 3  # Result: 1
12 %% 3  # Result: 0
15 %% 3  # Result: 0
# %% gives the remainder (modulo operation)
```

```{r modulo-exercise-check}
grade_this_code()
```

### What have we learned so far?

**We can use R as a calculator!**

*We can even do better than that...*

---

## Working with Vectors

<div class="callout callout-concept">
<div class="callout-title">What are Vectors?</div>

Vectors are the fundamental data structure in R. They store multiple values of the **same type** (all numbers, all text, etc.). Think of them as a ordered list of values.
</div>

### Variables and Data Structures

Variables store values that you can reuse throughout your code. In R, we use the assignment operator `<-` to assign values to variables.

```{r variables-demo, exercise=TRUE, exercise.setup="shared-data"}
myFirstNumber <- 0.1
myFirstVector <- c(2, 3, 7, 8)

# Call the variable by name
print(myFirstNumber)
myFirstNumber
```

### Data Structures in R

- **Vectors** - one-dimensional arrays of data
- **Matrix** - two-dimensional arrays
- **Data.frame** - tables with rows and columns
- **List** - collections of different objects
- **Factors** - categorical variables

### Creating Vectors with c()

You can make a vector using the function `c()` (concatenate).

```{r vector-creation, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4

# Print the vectors
myFirstVector
myFirstSequence
```

### Subsetting with [ ]

<div class="callout callout-concept">
<div class="callout-title">Subsetting Vectors</div>

Use square brackets `[ ]` to extract specific elements from a vector. You can subset by **position** (numeric index) or by **logical condition** (TRUE/FALSE values).
</div>

**Print the whole vector:**

```{r subsetting-demo, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector
myFirstSequence
```

### Print Third Element

```{r subsetting-single, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector[3]
myFirstSequence[3]
```

### Access by Position

<div class="callout callout-danger">
<div class="callout-title">Common Mistake: Wrong Subsetting Syntax</div>

**This won't work:**
```r
myFirstVector[1,3]  # Wrong! This is matrix/data.frame syntax
```

Vectors use single brackets with a vector of positions, not comma-separated values.

**Correct way** - provide a VECTOR of positions:
```r
myFirstVector[c(1, 3)]  # Right! Use c() to combine positions
```
</div>

```{r subsetting-multiple, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector[c(1, 3)]

# Or create a variable with positions
somePositions <- c(1, 3)
myFirstVector[somePositions]
```

### Exercise 1: Subsetting by Position

1. Create a vector named `myvector` that contains numbers 15, 16, 17, 18, and 20
2. Get the 2nd and 4th numbers by subsetting

```{r exercise-subsetting, exercise=TRUE}
# Create your vector here
myvector <- c(15, 16, 17, 18, 20)

# Subset to get 2nd and 4th elements

```

```{r exercise-subsetting-hint}
# Use c() to specify positions: c(2, 4)
myvector[c(2, 4)]
```

```{r exercise-subsetting-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(2, 4)]  # Returns 16 and 18
```

```{r exercise-subsetting-check}
grade_this_code()
```

### Access by Logical Vector

Provide a LOGICAL vector:

```{r logical-subsetting, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector
myFirstVector[c(TRUE, FALSE, TRUE, FALSE)]

# Or use a logical variable
someLogicalVector <- c(TRUE, TRUE, FALSE, FALSE)
myFirstVector[someLogicalVector]
```

### Exercise 2: Subsetting by Logical Vector

Get the 2nd and 4th numbers in `myvector` by providing a logical vector.

```{r exercise-logical, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Create logical vector and subset

```

```{r exercise-logical-hint}
# Think: which positions should be TRUE?
# Position: 1     2     3      4     5
# Logical:  FALSE TRUE  FALSE TRUE  FALSE
```

```{r exercise-logical-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE, FALSE, TRUE, FALSE)]
```

```{r exercise-logical-check}
grade_this_code()
```

### Understanding Type Conversion

```{r conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector <- c(1, 0, 1, 0, 1)
myFirstVector

# Numbers are converted to logical!
myFirstVector[someothervector]

# Explicit conversion
myFirstVector[as.logical(someothervector)]
```

---

## Vector Operations

<div class="callout callout-success">
<div class="callout-title">Vectorized Operations</div>

One of R's most powerful features is **vectorized operations**. Operations apply to each element automatically without needing loops!
</div>

### Basic Arithmetic

Same as on numbers: `+`, `-`, `/`, `*`

**Multiplication by Constant:**

```{r mult-constant, exercise=TRUE, exercise.setup="shared-data"}
someothervector
someothervector * 0.5
```

### Multiplication by Other Vector

**I) SAME size**

```{r mult-same, exercise=TRUE, exercise.setup="shared-data"}
someothervector
1:5
someothervector * 1:5
```

**II) DIFFERENT size: Recycling**

<div class="callout callout-warning">
<div class="callout-title">Vector Recycling</div>

When vectors of different lengths are used together, R **recycles** (repeats) the shorter vector. This can be useful but also a source of bugs if unexpected!
</div>

```{r mult-recycle, exercise=TRUE, exercise.setup="shared-data"}
someothervector
c(0.3, 0.1)
someothervector * c(0.3, 0.1)

# Vector recycling in action
c(1, 2, 3, 4, 5, 6) * 1:2
```

### Exercise 3: Multiplication & Recycling

1. Create `myvector` with values 15, 16, 17, 18, 20
2. Multiply `myvector` by `c(0.1, 0.2)` - what do you expect to get?
3. Check if you get what you expected by comparing

```{r exercise-mult, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Multiply by c(0.1, 0.2)


# Compare to your expected result

```

```{r exercise-mult-solution}
myvector <- c(15, 16, 17, 18, 20)
result <- myvector * c(0.1, 0.2)
result

# Expected: c(1.5, 3.2, 1.7, 3.6, 2.0)
expected <- c(1.5, 3.2, 1.7, 3.6, 2.0)
result == expected
```

```{r exercise-mult-check}
grade_this_code()
```

### Comparisons and Subsetting

<div class="callout callout-info">
<div class="callout-title">Logical Subsetting</div>

Comparison operators (`==`, `!=`, `>`, `<`, `>=`, `<=`) return TRUE/FALSE vectors. These can be used to filter data based on conditions - a fundamental skill in data analysis!
</div>

Comparisons return a logical vector:

```{r comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector
someothervector == 1

someothervector == c(1, 0, 1, 0, 1)

someothervector > 0
```

### What Happened Here?

```{r subset-by-comparison, exercise=TRUE, exercise.setup="shared-data"}
someothervector[someothervector > 0]
```

**Remember:** Subsetting by logical indices returns elements at positions which are TRUE!

### Exercise 4: Subsetting with Comparison

Return only elements in `myvector` that are smaller than 17.

```{r exercise-comparison, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Use subsetting and comparison
myvector[???]
```

```{r exercise-comparison-hint}
# Use the < operator to create a logical vector
myvector[myvector < 17]
```

```{r exercise-comparison-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector < 17]  # Returns 15 and 16
```

```{r exercise-comparison-check}
grade_this_code()
```

### Logical Operators

- **AND:** `&`
- **OR:** `|`
- **NOT:** `!`

```{r logical-ops, exercise=TRUE}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)

firstLogical & secondLogical  # AND
firstLogical | secondLogical  # OR
!firstLogical                 # NOT
```

### Exercise 5: Logical Operators

**Remember:**
- `x == 0` - where is x equal to 0
- `x %% 5` - gives you the modulo
- `x != 0` - not equal to

Return all elements from `myvector` that are divisible by 3.

```{r exercise-logical-ops, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Find elements divisible by 3

```

```{r exercise-logical-ops-hint}
# A number is divisible by 3 if modulo 3 equals 0
myvector[myvector %% 3 == 0]
```

```{r exercise-logical-ops-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector %% 3 == 0]  # Returns 15 and 18
```

```{r exercise-logical-ops-check}
grade_this_code()
```

### Extra Challenge!

Return all elements from `myvector` which are on EVEN positions (2, 4, 6, ...).

```{r exercise-even-positions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Get elements at even positions

```

```{r exercise-even-positions-solution}
myvector <- c(15, 16, 17, 18, 20)
# Create a logical vector for even positions
even_positions <- seq(1, length(myvector)) %% 2 == 0
myvector[even_positions]

# Or more directly:
myvector[c(FALSE, TRUE)]  # Uses recycling
```

```{r exercise-even-positions-check}
grade_this_code()
```

---

## Functions and Help

<div class="callout callout-tip">
<div class="callout-title">Essential Functions</div>

R has thousands of built-in functions. Learning the most common ones will make your work much more efficient.
</div>

### Useful Vector Functions

```{r func-length, exercise=TRUE, exercise.setup="shared-data"}
length(someothervector)
```

**unique() - Unique Elements:**

```{r func-unique, exercise=TRUE, exercise.setup="shared-data"}
unique(someothervector)
```

**table() - Count Elements:**

```{r func-table, exercise=TRUE, exercise.setup="shared-data"}
table(someothervector)
```

### Math Functions

```{r math-functions, exercise=TRUE, exercise.setup="shared-data"}
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```

### Try These Functions!

```{r exercise-functions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Calculate sum, mean, and standard deviation



```

```{r exercise-functions-solution}
myvector <- c(15, 16, 17, 18, 20)
sum(myvector)
mean(myvector)
sd(myvector)
```

```{r exercise-functions-check}
grade_this_code()
```

### Getting Help

```{r help-demo, exercise=TRUE, eval=FALSE}
?runif
example(runif)
```

You can also use:
- `?function_name` - opens help documentation
- `example(function_name)` - shows examples
- `help.search("keyword")` - searches help files

---

## Congratulations!

You've completed the Introduction to R tutorial!

### What You've Learned:

- Using R as a calculator  
- Creating and working with variables  
- Understanding vectors and data structures  
- Subsetting vectors by position and logical conditions  
- Vector operations and recycling  
- Comparison and logical operators  
- Useful R functions  
- How to get help in R

### Next Steps:

Continue exploring R programming and data analysis!

### Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/)
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)
- [R Documentation](https://www.rdocumentation.org/)

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```
