---
title: "Introduction to R - Part 1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()
# IntroToR variables
myFirstNumber <- 0.1
myFirstVector <- c(3,5,7,9)
x <- 5
name <- ""
mySecondVector <- c("some words", "p", "awesome", "last one")

myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
myvector <- c(15, 16, 17, 18, 20)

# data.table datasets
ddf <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
ddf_dt <- as.data.table(ddf)
iris_dt <- as.data.table(iris)

# ggplot2 datasets
tidy <- ddf
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))
```

## Welcome

Welcome to **Introduction to R - Part 1**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## What You Will Learn

<div class="callout callout-success">
<div class="callout-title">Tutorial Contents - Part 1</div>

In this tutorial, you will learn the fundamentals of R programming:

1. **Getting Started with R**  
   - Using R as a calculator  
   - Manually calculate mean in R  
   
2. **Variables and Data Structures**  
   - Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   - Creating variables and vectors  
   - Descriptive statistics (sd, var, median, range, IQR, quantiles)  
   
3. **Functions and Help**  
   - Vector structure functions  
   - Mathematical functions  
   - Z-scores, skewness, and kurtosis
   - Getting help in R  
   
4. **Working with Vectors** (Introduction)  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  

**Note:** Part 2 continues with logical subsetting, type conversion, comparisons, and logical operators.

</div>

---

## Getting Started with R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started with R**  
   → Using R as a calculator  
   - Manually calculate mean in R  

</div>

<div class="callout callout-info">
<div class="callout-title">R as a Calculator</div>

R can perform mathematical operations just like a calculator. This is your first step in learning R programming!
</div>

### How can we use R and RStudio?

- Console
- Script
- Notebook (add new chunk: Ctrl + Alt + I)

Try using R as a calculator. Type `1 + 1` and run the code:

```{r calculator-demo, exercise=TRUE}
# Use R as a calculator


```

```{r calculator-demo-hint}
1 + 1
```

```{r calculator-demo-solution}
1 + 1
```

### Try it!

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started with R**  
   - Using R as a calculator  
   → Manually calculate mean in R  

</div>

<div class="callout callout-tip">
<div class="callout-title">Calculating the Mean</div>

The **mean** (average) is calculated by adding all numbers together and dividing by how many numbers there are. For example, the mean of 2, 4, and 6 is `(2 + 4 + 6) / 3 = 4`.
</div>

Calculate the mean value of the numbers 3, 5, 7, and 9 using the operators `+`, `/`, and `()`.

```{r mean-exercise, exercise=TRUE}
# Calculate the mean of 3, 5, 7, 9


```

```{r mean-exercise-hint}
# Add all numbers together: (3 + 5 + 7 + 9)
# Then divide by how many numbers there are: / 4
# Put it together: (3 + 5 + 7 + 9) / 4
```

```{r mean-exercise-solution}
# Calculate the mean
(3 + 5 + 7 + 9) / 4
# Result: 6
```

### What have we learned so far?

**We can use R as a calculator!**

*We can even do better than that...*

---


## Variables and Data Structures

### Data Structures in R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Variables and Data Structures**  
   → Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   - Creating variables and vectors  

</div>

<div class="callout callout-concept">
<div class="callout-title">R Data Structures Overview</div>

R has several fundamental data structures for organizing information:

**1. Vector** - One-dimensional array (all same type)
```
[1, 2, 3, 4, 5]
```

**2. Matrix** - Two-dimensional array (all same type)
```
     [,1] [,2] [,3]
[1,]    1    4    7
[2,]    2    5    8
[3,]    3    6    9
```

**3. Data Frame** - Table with named columns (different types allowed)
```
  name    age  score
1 "Anna"   25   95.5
2 "Bob"    30   87.3
3 "Chris"  22   91.0
```

**4. List** - Collection of any objects (can mix types)
```
[[1]] 10
[[2]] c("a", "b") 
[[3]] data.frame(x=1:3)
```

**5. Factor** - Categorical variable with levels
```
[1] low  high medium high
Levels: low medium high
```

</div>

Here is an examples of how basic R data structure are created: vector, matrix, data frame, list, and factor:

```{r data-structures-visual, exercise=TRUE, exercise.setup="shared-data"}
my_vector <- c(1, 2, 3, 4, 5)
my_vector
```

```{r data-structures-visual2, exercise=TRUE, exercise.setup="shared-data"}
my_matrix <- matrix(1:9, nrow=3, ncol=3)
my_matrix
```

```{r data-structures-visual3, exercise=TRUE, exercise.setup="shared-data"}

my_df <- data.frame(
  name = c("Anna", "Bob", "Chris"),
  age = c(25, 30, 22),
  score = c(95.5, 87.3, 91.0)
)
my_df
```
```{r data-structures-visual4, exercise=TRUE, exercise.setup="shared-data"}

my_list <- list(
  numbers = 1:5,
  text = "hello",
  data = data.frame(x=1:3, y=4:6)
)
my_list
```
```{r data-structures-visual5, exercise=TRUE, exercise.setup="shared-data"}

my_factor <- factor(c("low", "high", "medium", "high"),
                    levels = c("low", "medium", "high"))
my_factor
```

**Key Differences:** 
 
- **Vector & Matrix**: All elements must be the same type (numeric, character, etc.)  
- **Data Frame**: Each column can be a different type (like a spreadsheet)  
- **List**: Most flexible - can contain any mix of objects  
- **Factor**: Stores categorical data efficiently with predefined levels  

---

### Variables

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Variables and Data Structures**  
   - Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   → Creating variables and vectors  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Variables?</div>

A **variable** is a named container that stores a value in memory. Think of it as a labeled box — you put something inside (a number, text, a whole dataset), and later you can retrieve it by its name.

**Why use variables?**  
- Avoid repeating the same value over and over  
- Make code readable — `height` is clearer than `173.5`  
- Store results of calculations for later use  

**Naming rules:**  
- Must start with a letter (not a number)  
- Can contain letters, numbers, `.` and `_`  
- Case-sensitive: `myVar` and `myvar` are different  

</div>

<div class="callout callout-info">
<div class="callout-title">Working with Variables</div>

In R, we create variables using the **assignment operator** `<-`:

```r
x <- 5         # x now holds the value 5
name <- "Maja"  # name now holds the text "Maja"
```

- To see the value stored in a variable, simply type its name to print its content  
- To store multiple values in a variable, use `c()` — this creates a **vector**  

</div>

- Create a variable `myFirstNumber` with value `0.1` and a vector `myFirstVector` with `c(3, 5, 7, 9)`.   
- Print `myFirstNumber`.  
- now calculate the mean of the four numbers 3,5,7,9 by writing mean(myFirstVector).  

```{r variables-demo, exercise=TRUE, exercise.setup="shared-data"}
# Create variables and print them


```

```{r variables-demo-hint}
myFirstNumber <- 0.1
myFirstVector <- c(3, 5, 7, 9)
print(myFirstNumber)
myFirstNumber
```

```{r variables-demo-solution}
myFirstNumber <- 0.1
myFirstVector <- c(3, 5, 7, 9)
print(myFirstNumber)
mean(myFirstVector)
```

Now use `myFirstVector` to calculate:

1. The **standard deviation** using `sd()`
2. The **variance** using `var()`
3. The **median** using `median()`
4. The **range** (min and max) using `range()`

```{r variables-stats, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c(3, 5, 7, 9)

# Calculate standard deviation


# Calculate variance


# Calculate median


# Calculate range (min and max)


```

```{r variables-stats-hint}
# Standard deviation
sd(myFirstVector)
# Variance
var(myFirstVector)
```

```{r variables-stats-hint-2}
# Median
median(myFirstVector)
# Range
range(myFirstVector)
```

```{r variables-stats-solution}
myFirstVector <- c(3, 5, 7, 9)
sd(myFirstVector)      # 2.581989
var(myFirstVector)      # 6.666667
median(myFirstVector)   # 6
range(myFirstVector)    # 3 9
```

Now calculate:

1. The **interquartile range (IQR)** using `IQR()`
2. The **5th percentile** using `quantile()` with `probs = 0.05`
3. The **95th percentile** using `quantile()` with `probs = 0.95`

```{r variables-quantiles, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c(3, 5, 7, 9)

# Calculate IQR


# Calculate 5th percentile


# Calculate 95th percentile


```

```{r variables-quantiles-hint}
# IQR = difference between 75th and 25th percentile
IQR(myFirstVector)
# quantile() with probs argument
quantile(myFirstVector, probs = 0.05)
```

```{r variables-quantiles-hint-2}
# 95th percentile
quantile(myFirstVector, probs = 0.95)
# You can also get multiple quantiles at once:
quantile(myFirstVector, probs = c(0.05, 0.95))
```

```{r variables-quantiles-solution}
myFirstVector <- c(3, 5, 7, 9)
IQR(myFirstVector)                          # 3
quantile(myFirstVector, probs = 0.05)       # 5th percentile
quantile(myFirstVector, probs = 0.95)       # 95th percentile
quantile(myFirstVector, probs = c(0.05, 0.95))  # both at once
```

---

<!-- NEW CONCEPT: FUNCTIONS -->
## Functions and Help

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   → Vector structure functions  
   - Mathematical functions  
   - Getting help in R

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Built-in Functions</div>

R has thousands of **built-in functions** that perform common tasks:
- **Structure functions**: `length()`, `unique()`, `table()`
- **Math functions**: `sum()`, `mean()`, `sd()`, `median()`
- **Statistical functions**: `summary()`, `min()`, `max()`

Functions take input (arguments) and return output.
</div>

---

<!-- STEP 2: DEMONSTRATE - Structure Functions -->
### Demonstration: Vector Structure Functions

Use `length()`, `unique()`, and `table()` on `someothervector`:

```{r func-structure-demo, exercise=TRUE, exercise.setup="shared-data"}
# Explore vector structure with functions


```

```{r func-structure-demo-hint}
someothervector
length(someothervector)
unique(someothervector)
table(someothervector)
```

```{r func-structure-demo-solution}
someothervector
length(someothervector)
unique(someothervector)
table(someothervector)
```

**What's happening here?**
- `length()` returns the number of elements (5)
- `unique()` removes duplicates (0 and 1)
- `table()` counts how many times each value appears

---

<!-- STEP 3: DEMONSTRATE - Math Functions -->
### Demonstration: Mathematical Functions

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   - Vector structure functions  
   → Mathematical functions  
   - Getting help in R

</div>

Calculate `sum()`, `mean()`, `sd()`, and `summary()` of `someothervector`:

```{r math-functions-demo, exercise=TRUE, exercise.setup="shared-data"}
# Calculate statistics on someothervector


```

```{r math-functions-demo-hint}
someothervector
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```

```{r math-functions-demo-solution}
someothervector
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```

**What's happening here?**
- `sum()` adds all elements: 1+0+1+0+1 = 3
- `mean()` calculates average: 3/5 = 0.6
- `sd()` measures variability
- `summary()` gives min, max, median, mean, quartiles

---

### Z-Score (Standardization)

<div class="callout callout-concept">
<div class="callout-title">What is a Z-Score?</div>

A **z-score** tells you how many standard deviations a value is from the mean. It is calculated as:

$$z = \frac{x - \bar{x}}{s}$$

where $\bar{x}$ is the mean and $s$ is the standard deviation.

**In R you can:**. 
- Calculate manually: `(x - mean(x)) / sd(x)`  
- Use the built-in function: `scale(x)`. 

</div>

**Task:**
Create a vector `scores <- c(70, 85, 90, 60, 95)` and calculate the z-score for each value.

```{r zscore-exercise, exercise=TRUE}
scores <- c(70, 85, 90, 60, 95)

# Calculate z-scores manually


# Or use the built-in scale() function


```

```{r zscore-exercise-hint}
# Manual z-score: subtract the mean, then divide by sd
z_manual <- (scores - mean(scores)) / sd(scores)
z_manual
```

```{r zscore-exercise-hint-2}
# Built-in function (returns a matrix, use as.numeric() to get a vector)
z_builtin <- scale(scores)
z_builtin
```

```{r zscore-exercise-solution}
scores <- c(70, 85, 90, 60, 95)

# Manual calculation
z_manual <- (scores - mean(scores)) / sd(scores)
z_manual

# Built-in function
z_builtin <- scale(scores)
z_builtin
```

---

### Skewness and Kurtosis

<div class="callout callout-concept">
<div class="callout-title">Skewness & Kurtosis</div>

**Skewness** measures the **asymmetry** of a distribution:
- **Skewness = 0**: Symmetric (like a normal distribution)
- **Skewness > 0**: Right-skewed (long tail to the right)
- **Skewness < 0**: Left-skewed (long tail to the left)

**Kurtosis** measures how heavy the **tails** are:
- **Kurtosis ≈ 3** (or excess kurtosis ≈ 0): Normal distribution
- **Kurtosis > 3**: Heavier tails (more outliers)
- **Kurtosis < 3**: Lighter tails (fewer outliers)

R does not have built-in skewness/kurtosis functions, but you can calculate them manually or use a package like `moments`.

**Manual formulas:**

$$\text{skewness} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^3$$

$$\text{kurtosis} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^4$$

</div>

**Task:**
Using the vector `scores <- c(70, 85, 90, 60, 95)`, calculate skewness and kurtosis manually.

```{r skewness-exercise, exercise=TRUE}
scores <- c(70, 85, 90, 60, 95)

# Step 1: Standardize (z-scores)
z <- (scores - mean(scores)) / sd(scores)

# Step 2: Calculate skewness (mean of z^3)


# Step 3: Calculate kurtosis (mean of z^4)


```

```{r skewness-exercise-hint}
# Skewness: raise z-scores to the 3rd power, then take the mean
z <- (scores - mean(scores)) / sd(scores)
skewness <- mean(z^3)
skewness
# Negative value → left-skewed (tail extends to the left)
```

```{r skewness-exercise-hint-2}
# Kurtosis: raise z-scores to the 4th power, then take the mean
z <- (scores - mean(scores)) / sd(scores)
kurtosis <- mean(z^4)
kurtosis
# Excess kurtosis (compared to normal distribution):
excess_kurtosis <- kurtosis - 3
excess_kurtosis
```

```{r skewness-exercise-solution}
scores <- c(70, 85, 90, 60, 95)

# Standardize
z <- (scores - mean(scores)) / sd(scores)

# Skewness (mean of z^3)
skewness <- mean(z^3)
skewness    # negative → left-skewed

# Kurtosis (mean of z^4)
kurtosis <- mean(z^4)
kurtosis

# Excess kurtosis (relative to normal distribution = 3)
excess_kurtosis <- kurtosis - 3
excess_kurtosis
```

---

<!-- STEP 5: GETTING HELP -->
### Getting Help in R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   - Vector structure functions  
   - Mathematical functions  
   → Getting help in R

</div>

<div class="callout callout-info">
<div class="callout-title">Help Resources</div>

When you need to learn about a function:
- **`?function_name`** - Opens help documentation
- **`example(function_name)`** - Shows working examples
- **`help.search("keyword")`** - Searches help files

```r
?mean
example(mean)
help.search("average")
```
</div>

Try `?mean` and `example(runif)` to explore R's help system:

```{r help-demo, exercise=TRUE, eval=FALSE}
# Try getting help on functions


```

```{r help-demo-hint, eval=FALSE}
?mean
example(runif)
```

```{r help-demo-solution, eval=FALSE}
?mean
example(runif)
```

---

## Working with Vectors

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   → Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Vectors?</div>

Vectors are the fundamental data structure in R. They store multiple values of the **same type** (all numbers, all text, etc.). Think of them as a ordered list of values.
</div>

### Creating Vectors with c()

You can make a vector using the function `c()` (concatenate).

Create `mySecondVector` with `c("some words", "p", "awesome", "last one")` and `myFirstSequence` with `1:4`, then print them:

```{r vector-creation, exercise=TRUE, exercise.setup="shared-data"}
# Create vectors and display them


```

```{r vector-creation-hint}
mySecondVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4
mySecondVector
myFirstSequence
```

```{r vector-creation-solution}
mySecondVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4
mySecondVector
myFirstSequence
```

### Subsetting with [ ]

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   → Subsetting with `[ ]`  
   - Accessing single and multiple elements  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Subsetting Vectors</div>

**Subsetting** means extracting specific elements from a vector. Think of it like selecting items from a list. In R, we use square brackets `[ ]` to do this.   

**Two main ways to subset:**.  
- **By position**: Get elements at specific locations (1st, 3rd, etc.).  
- **By condition**: Get elements that meet certain criteria (> 5, == "text", etc.).  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Accessing Single Elements

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`   
   - Subsetting with `[ ]`  
   → Accessing single and multiple elements  

</div>

Let's start simple - getting just one element by its position:  

Display `mySecondVector` and `myFirstSequence`, then get the 3rd element from each using `[3]`:  

```{r subsetting-single, exercise=TRUE, exercise.setup="shared-data"}
# Get the 3rd element from each vector


```

```{r subsetting-single-hint}
mySecondVector
mySecondVector[3]
```

```{r subsetting-single-solution}
myFirstVector
myFirstSequence
myFirstVector[3]
myFirstSequence[3]
```

**What's happening here?**
- `mySecondVector[3]` says "give me the element at position 3".  
- R uses **1-based indexing** (first element is position 1, not 0).  
- Result: The third item from each vector.  

---

<!-- STEP 3: DEMONSTRATE - Multiple Elements -->
### Demonstration: Accessing Multiple Elements

Now let's get multiple elements at once:   

Get elements at positions 1 and 3 from `mySecondVector` using `c()`, then try storing positions in a variable:   

```{r subsetting-multiple-demo, exercise=TRUE, exercise.setup="shared-data"}
# Get elements at positions 1 and 3


```

```{r subsetting-multiple-demo-hint}
mySecondVector[c(1, 3)]

```

```{r subsetting-multiple-demo-hint-2}
positions <- c(1, 2, 4)
mySecondVector[positions]
```


```{r subsetting-multiple-demo-solution}
mySecondVector[c(1, 3)]
positions <- c(1, 2, 4)
mySecondVector[positions]
```

**What's happening here?**
- `c(1, 3)` creates a vector of positions we want.  
- `mySecondVector[c(1, 3)]` returns elements at those positions.  
- You MUST use `c()` to combine multiple positions.  

---

<!-- STEP 4: COMMON MISTAKE WARNING -->
<div class="callout callout-danger">
<div class="callout-title">Common Mistake: Wrong Subsetting Syntax</div>

**This won't work:**
```r
mySecondVector[1,3]  # ❌ WRONG! This is matrix/data.frame syntax
```

**Why it fails:** Vectors are one-dimensional. The comma `,` is for 2D structures (matrices, data frames).

**Correct way:**
```r
mySecondVector[c(1, 3)]  # ✅ RIGHT! Use c() to combine positions  
```
</div>

---

<!-- STEP 5: EXERCISE - Now Student Tries -->
### Your Turn: Practice Subsetting by Position  

**Task:**   
1. Create a vector named `myvector` with numbers: 15, 16, 17, 18, and 20.  
2. Use subsetting to get the 2nd and 4th elements.  

```{r exercise-subsetting, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Subset to get 2nd and 4th elements (should return 16 and 18)

```

```{r exercise-subsetting-hint}
# Use c() to create a vector
myvector <- c(15, 16, 17, 18, 20)
```

```{r exercise-subsetting-hint-2}
# Use c() to specify positions: c(2, 4)
myvector[c(2, 4)]
```

```{r exercise-subsetting-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(2, 4)]  # Returns 16 and 18
```

---

## Quick Reference: Functions Used in Part 1

<div class="callout callout-info">
<div class="callout-title">Functions Used in This Tutorial</div>

**Creating Objects:**

| Function | Description | Example |
|----------|-------------|--------|
| `c()` | Combine values into a vector | `c(1, 2, 3)` |
| `matrix()` | Create a matrix | `matrix(1:9, nrow=3)` |
| `data.frame()` | Create a data frame | `data.frame(x=1:3, y=4:6)` |
| `list()` | Create a list | `list(a=1, b="text")` |
| `factor()` | Create a factor (categorical) | `factor(c("low", "high"))` |
| `:` | Sequence operator | `1:10` creates 1,2,3...10 |

**Descriptive Statistics:**

| Function | Description | Example |
|----------|-------------|--------|
| `mean()` | Arithmetic mean | `mean(x)` |
| `median()` | Median value | `median(x)` |
| `sd()` | Standard deviation | `sd(x)` |
| `var()` | Variance | `var(x)` |
| `sum()` | Sum of all elements | `sum(x)` |
| `min()` | Minimum value | `min(x)` |
| `max()` | Maximum value | `max(x)` |
| `range()` | Min and max | `range(x)` |
| `IQR()` | Interquartile range | `IQR(x)` |
| `quantile()` | Quantiles / percentiles | `quantile(x, probs=0.95)` |
| `summary()` | Summary statistics | `summary(x)` |
| `scale()` | Standardize (z-scores) | `scale(x)` |

**Vector Structure:**

| Function | Description | Example |
|----------|-------------|--------|
| `length()` | Number of elements | `length(x)` |
| `unique()` | Remove duplicates | `unique(x)` |
| `table()` | Frequency table | `table(x)` |

**Subsetting:**

| Operator | Description | Example |
|----------|-------------|--------|
| `[ ]` | Extract elements by position | `x[1]`, `x[c(1,3)]` |

**Output & Help:**

| Function | Description | Example |
|----------|-------------|--------|
| `print()` | Print a value | `print(x)` |
| `?` | Open help page | `?mean` |
| `example()` | Run examples from help | `example(mean)` |
| `help.search()` | Search help files | `help.search("average")` |

</div>

---

## Congratulations!

You've completed Part 1 of the Introduction to R tutorial!

### What You've Learned:

- Using R as a calculator  
- Creating and working with variables  
- Understanding R data structures (vectors, matrices, data frames, lists, factors)  
- Descriptive statistics (sd, var, median, range, IQR, quantiles)  
- Calculating z-scores, skewness, and kurtosis
- Using R functions and getting help
- Creating vectors with `c()`  
- Subsetting vectors by position using `[ ]`

### Next Steps:

**Continue with Part 2** to learn about:
- Logical subsetting (using TRUE/FALSE)
- Type conversion  
- Comparison operators and conditional filtering  
- Logical operators (AND, OR, NOT)  
- Vector operations and recycling  

### Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/)  
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)  
- [R Documentation](https://www.rdocumentation.org/)  

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

## Please Give Feedback!

<div class="callout callout-success">
<div class="callout-title">💬 Your Feedback Matters</div>

Help us improve this tutorial! Share your experience and suggestions.
</div>

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/B7Wy7G29wQTuEKjU9" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```
