---
title: "Introduction to R - Part 2"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()
# IntroToR variables
myFirstNumber <- 0.1
myFirstVector <- c(3,5,7,9)
x <- 5
name <- ""
mySecondVector <- c("some words", "p", "awesome", "last one")

myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
myvector <- c(15, 16, 17, 18, 20)

# data.table datasets
ddf <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
ddf_dt <- as.data.table(ddf)
iris_dt <- as.data.table(iris)

# ggplot2 datasets
tidy <- ddf
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))
```

## Welcome

Welcome to **Introduction to R - Part 2**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

### About This Tutorial

This is the second part of the Introduction to R tutorial series. In Part 1, you learned about R as a calculator, variables, data structures, functions, and basic vector subsetting by position.

In Part 2, you'll continue learning about vectors and discover powerful techniques for filtering and manipulating data.

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## What You Will Learn

<div class="callout callout-success">
<div class="callout-title">Tutorial Contents - Part 2</div>

This tutorial continues from Part 1 with advanced vector operations:

4. **Working with Vectors** (Continued)  
   - Using logical vectors for subsetting (TRUE/FALSE)  
   - Understanding type conversion (`as.logical()`, `as.numeric()`)  
   
5. **Comparisons and Subsetting**  
   - Using comparison operators (`==`, `!=`, `>`, `<`, `>=`, `<=`)  
   - Conditional subsetting and filtering  
   
6. **Logical Operators**  
   - Understanding AND (`&`), OR (`|`), NOT (`!`)  
   - Multi-condition filtering  

**Additional Content:**  
- Vector Operations (multiplication, element-wise operations, recycling)

**Prerequisites:** Complete Part 1 or be familiar with R basics, variables, data structures, functions, and positional subsetting.

</div>

---

## Working with Vectors (Continued)

### Access by Logical Vector

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   → Using logical vectors for subsetting  
   - Understanding type conversion  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Logical Subsetting</div>

Instead of positions (1, 2, 3...), you can use TRUE/FALSE values to select elements:  
- **TRUE** = keep this element  
- **FALSE** = skip this element  

The logical vector must be the same length as your data vector.  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Using TRUE/FALSE to Select Elements

Use `TRUE/FALSE` values to select elements from `mySecondVector`. Select the 1st and 3rd elements:

```{r logical-subsetting-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use TRUE/FALSE to select elements


```

```{r logical-subsetting-demo-hint}
mySecondVector
mySecondVector[c(TRUE, FALSE, TRUE, FALSE)]
```

```{r logical-subsetting-demo-solution}
mySecondVector
mySecondVector[c(TRUE, FALSE, TRUE, FALSE)]
```

**What's happening here?**
- Position 1: TRUE → include "some words"  
- Position 2: FALSE → skip "p"  
- Position 3: TRUE → include "awesome"  
- Position 4: FALSE → skip "last one"  

---

<!-- STEP 3: DEMONSTRATE - Using Variables -->
### Demonstration: Storing Logical Vectors

You can save logical vectors in variables:

Create a logical vector `my_selection` and use it to subset `mySecondVector`:

```{r logical-variable-demo, exercise=TRUE, exercise.setup="shared-data"}
# Store a logical vector and use it for subsetting


```

```{r logical-variable-demo-hint}
my_selection <- c(TRUE, TRUE, FALSE, FALSE)
mySecondVector[my_selection]
```

```{r logical-variable-demo-solution}
my_selection <- c(TRUE, TRUE, FALSE, FALSE)
mySecondVector[my_selection]
```

**Why is this useful?** 
- Logical vectors can be created automatically from conditions (coming next!)  
- Makes code more readable and reusable  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Pro Tip: Logical Vector Length</div>

The logical vector should match the length of your data:
```r
mySecondVector[c(TRUE, FALSE)]  # Only 2 values for 4-element vector?
```

R will **recycle** the pattern: TRUE, FALSE, TRUE, FALSE  

This can be useful but also cause errors if unintended!  
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Logical Subsetting

**Task:** 
Use a logical vector to get the 2nd and 4th elements from `myvector`.

Think about which positions should be TRUE vs FALSE.

```{r exercise-logical, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Create a logical vector to select 2nd and 4th elements


# Use it to subset myvector

```

```{r exercise-logical-hint}
# Think: which positions should be TRUE?
# Position: 1     2     3      4     5
# Logical:  FALSE TRUE  FALSE TRUE  FALSE
```

```{r exercise-logical-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE, FALSE, TRUE, FALSE)]
```

---

<!-- NEW CONCEPT: TYPE CONVERSION -->
### Understanding Type Conversion

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   → Understanding type conversion  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Automatic Type Conversion</div>

R sometimes converts data types automatically:
- **Numbers to Logical:** 0 = FALSE, any other number = TRUE
- When you use numbers where logical values are expected, R converts them

You can also convert explicitly using `as.logical()`, `as.numeric()`, etc.
</div>

---

<!-- STEP 2: DEMONSTRATE - Automatic Conversion -->
### Demonstration: Numbers as Logical Vectors

Use `someothervector` (which contains 1s and 0s) to subset `mySecondVector`. Observe how R behaves:

```{r conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use a numeric vector with 0s and 1s to subset


```

```{r conversion-demo-hint}
someothervector <- c(1, 0, 1, 0, 1)
mySecondVector[someothervector]
```

```{r conversion-demo-solution}
someothervector <- c(1, 0, 1, 0, 1)
mySecondVector[someothervector]
```

**What's happening here?**
- R sees `someothervector` contains numbers  
- Context requires logical (TRUE/FALSE) for subsetting  
- R does not automatically convert: 1 → TRUE, 0 → FALSE
- Result: Elements at positions 1, 1, 1

---

<!-- STEP 3: DEMONSTRATE - Explicit Conversion -->
### Demonstration: Explicit Conversion with as.logical()

Explicitly convert `someothervector` to logical using `as.logical()` and use it to subset `myFirstVector`:  

```{r explicit-conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
# Explicitly convert to logical and subset


```

```{r explicit-conversion-demo-hint}
someothervector <- c(1, 0, 1, 0, 1)
as.logical(someothervector)
mySecondVector[as.logical(someothervector)]
```

```{r explicit-conversion-demo-solution}
someothervector <- c(1, 0, 1, 0, 1)
as.logical(someothervector)
mySecondVector[as.logical(someothervector)]
```

**Why use explicit conversion?**
- Makes your code clearer and more readable  
- Prevents surprises if R's automatic conversion isn't what you expect  

---

<!-- STEP 4: INFO CALLOUT -->
<div class="callout callout-info">
<div class="callout-title">Conversion Rules</div>

**Numeric to Logical:**
- 0 → FALSE  
- Any other number (1, 2, -5, 0.5, etc.) → TRUE  

**Logical to Numeric:**
- FALSE → 0  
- TRUE → 1  
</div>

---

<!-- NEW CONCEPT: COMPARISONS AND FILTERING -->
## Comparisons and Subsetting

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Comparisons and Subsetting**  
   → Using comparison operators  
   - Conditional subsetting and filtering  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Filtering Data with Comparisons</div>

**Comparison operators** create TRUE/FALSE vectors automatically:  
- `==` equals  
- `!=` not equals  
- `>` greater than  
- `<` less than  
- `>=` greater or equal  
- `<=` less or equal  

These logical vectors can be used for **conditional subsetting** - one of the most powerful data analysis tools!  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Comparisons -->
### Demonstration: Comparison Operators

Check which elements of `someothervector` equal 1 (`== 1`) and which are greater than 0 (`> 0`):

```{r comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use comparison operators


```

```{r comparison-demo-hint}
someothervector
someothervector == 1
someothervector > 0
```

```{r comparison-demo-solution}
someothervector
someothervector == 1
someothervector > 0
```

**What's happening here?**
- `someothervector == 1` checks each element: is it 1?  
- Returns: `TRUE FALSE TRUE FALSE TRUE`  
- `someothervector > 0` checks: is it greater than 0?  
- Returns logical vector showing which positions meet the condition  

---

<!-- STEP 3: DEMONSTRATE - Using Comparisons for Subsetting -->
### Demonstration: Conditional Subsetting

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Comparisons and Subsetting**  
   - Using comparison operators  
   → Conditional subsetting and filtering  

</div>

Subset `someothervector` to get only elements greater than 0:

```{r subset-by-comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
# Filter to keep only elements > 0


```

```{r subset-by-comparison-demo-hint}
someothervector
someothervector[someothervector > 0]
```

```{r subset-by-comparison-demo-solution}
someothervector
someothervector[someothervector > 0]
```

**What's happening here?**
1. `someothervector > 0` creates: `TRUE FALSE TRUE FALSE TRUE`  
2. Use this as subsetting index  
3. Result: only elements where condition is TRUE (values: 1, 1, 1)  

**This is filtering!** You're selecting data based on conditions.  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Breaking Down Complex Subsetting</div>

You can split complex subsetting into steps:

```r
# Step by step  
condition <- someothervector > 0  
condition  # See the TRUE/FALSE pattern  
someothervector[condition]  # Apply it  
  
# Or in one line  
someothervector[someothervector > 0]  
```

Both give the same result!
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Conditional Subsetting

**Task:** 
Get only the elements from `myvector` that are **smaller than 17**.  

```{r exercise-comparison, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)  

# Your code here: filter for values < 17

```

```{r exercise-comparison-hint}
# Use the < operator to create a logical vector
myvector[myvector < 17]
```

```{r exercise-comparison-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector < 17]  # Returns 15 and 16
```

---

<!-- NEW CONCEPT: LOGICAL OPERATORS -->
## Logical Operators

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

6. **Logical Operators**  
   → Understanding AND, OR, NOT    
   - Multi-condition filtering  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Combining Conditions</div>

Logical operators let you combine multiple TRUE/FALSE conditions:  
- **AND (`&`)**: Both must be TRUE    
- **OR (`|`)**: At least one must be TRUE    
- **NOT (`!`)**: Flips TRUE ↔ FALSE  

These let you create complex filters: "values greater than 10 AND less than 20"
</div>

---

<!-- STEP 2: DEMONSTRATE - AND, OR, NOT -->
### Demonstration: Basic Logical Operations

Create `firstLogical` and `secondLogical` vectors. Then test AND (`&`), OR (`|`), and NOT (`!`) operators:

```{r logical-ops-demo, exercise=TRUE}
# Test logical operators: &, |, !


```

```{r logical-ops-demo-hint}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
!firstLogical
```

```{r logical-ops-demo-solution}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
!firstLogical
```

**What's happening here?**
- `&` (AND): TRUE only when BOTH are TRUE  
  - Position 1: TRUE & TRUE = **TRUE**  
  - Position 2: TRUE & FALSE = **FALSE**  
- `|` (OR): TRUE when AT LEAST ONE is TRUE  
  - Position 2: TRUE | FALSE = **TRUE**  
  - Position 3: FALSE | TRUE = **TRUE**  
- `!` (NOT): Flips each value  
  - TRUE becomes FALSE, FALSE becomes TRUE  

---

<!-- STEP 3: DEMONSTRATE - Combining Conditions for Filtering -->
### Demonstration: Multi-Condition Filtering

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

6. **Logical Operators**  
   - Understanding AND, OR, NOT    
   → Multi-condition filtering  

</div>

Create a `numbers` vector `c(5, 10, 15, 20, 25)`. Use `&` to find numbers between 10 and 20 (inclusive), then use the condition for subsetting:

```{r multi-condition-demo, exercise=TRUE}
# Filter numbers between 10 and 20


```

```{r multi-condition-demo-hint}
numbers <- c(5, 10, 15, 20, 25)
numbers >= 10 & numbers <= 20
numbers[numbers >= 10 & numbers <= 20]
```

```{r multi-condition-demo-solution}
numbers <- c(5, 10, 15, 20, 25)
numbers >= 10 & numbers <= 20
numbers[numbers >= 10 & numbers <= 20]
```

**What's happening here?**
1. `numbers >= 10` → `FALSE TRUE TRUE TRUE TRUE`  
2. `numbers <= 20` → `TRUE TRUE TRUE TRUE FALSE`  
3. Combine with `&` → `FALSE TRUE TRUE TRUE FALSE`  
4. Result: 10, 15, 20 (values meeting BOTH conditions)  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Truth Table Quick Reference</div>

| A     | B     | A & B | A \| B | !A    |
|-------|-------|-------|--------|-------|
| TRUE  | TRUE  | TRUE  | TRUE   | FALSE |
| TRUE  | FALSE | FALSE | TRUE   | FALSE |
| FALSE | TRUE  | FALSE | TRUE   | TRUE  |
| FALSE | FALSE | FALSE | FALSE  | TRUE  |
</div>

---

<!-- STEP 5: EXERCISE -->
### Extra Challenge!  

Your Turn: Practice Logical Operators

**Task:** 
Find all elements from `myvector` that are **divisible by 3**.

**Hint:** A number is divisible by 3 if `number %% 3 == 0`

```{r exercise-logical-ops, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Find elements divisible by 3

```

```{r exercise-logical-ops-hint}
# A number is divisible by 3 if modulo 3 equals 0
myvector[myvector %% 3 == 0]
```

```{r exercise-logical-ops-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector %% 3 == 0]  # Returns 15 and 18
```

---

### Extra Challenge!

Return all elements from `myvector` which are on EVEN positions (2, 4, 6, ...).

```{r exercise-even-positions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Get elements at even positions

```

```{r exercise-even-positions-hint}
# You can use recycling: c(FALSE, TRUE) will create FALSE, TRUE, FALSE, TRUE, FALSE
myvector[c(FALSE, TRUE)]
```

```{r exercise-even-positions-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE)]  # Uses recycling: positions 2 and 4
```

---

## Additional Content: Vector Operations

<div class="callout callout-info">
<div class="callout-title">📍 Additional Content</div>

**Vector Operations**  
- Multiplication by constants  
- Element-wise operations  
- Vector recycling  

</div>

<div class="callout callout-concept">
<div class="callout-title">Vectorized Operations</div>

One of R's most powerful features is **vectorized operations**:
- Operations apply to **each element automatically**
- No need for loops!
- Works with: `+`, `-`, `*`, `/`, and more

**Example:** `c(1,2,3) * 2` multiplies each element by 2
</div>

---

### Demonstration: Multiplication by a Constant

Multiply `someothervector` by `0.5` and observe the element-wise result:

```{r mult-constant-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply vector by a constant


```

```{r mult-constant-demo-hint}
someothervector
someothervector * 0.5
```

```{r mult-constant-demo-solution}
someothervector
someothervector * 0.5
```

**What's happening here?**
- Each element is multiplied by 0.5
- No loop needed - R does it automatically
- Result: `[1] 0.5 0.0 0.5 0.0 0.5`

---

### Demonstration: Multiplying Two Vectors (Same Length)

Multiply `someothervector` by `1:5` (element by element) and observe the result:

```{r mult-same-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply two vectors of same length


```

```{r mult-same-demo-hint}
someothervector
1:5
someothervector * 1:5
```

```{r mult-same-demo-solution}
someothervector
1:5
someothervector * 1:5
```

**What's happening here?**
- Element 1 × Element 1: `1 × 1 = 1`  
- Element 2 × Element 2: `0 × 2 = 0`  
- Element 3 × Element 3: `1 × 3 = 3`  
- And so on...  

---

### Demonstration: Different Size Vectors (Recycling)

Multiply `someothervector` (length 5) by `c(0.3, 0.1)` (length 2) and observe how R recycles the shorter vector:  

```{r mult-recycle-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply vectors of different lengths (recycling)


```

```{r mult-recycle-demo-hint}
someothervector
c(0.3, 0.1)
someothervector * c(0.3, 0.1)
```

```{r mult-recycle-demo-solution}
someothervector
c(0.3, 0.1)
someothervector * c(0.3, 0.1)
```

**What's happening here?**
R repeats `c(0.3, 0.1)` to match the length:  
- Position 1: `1 × 0.3 = 0.3`  
- Position 2: `0 × 0.1 = 0.0`  
- Position 3: `1 × 0.3 = 0.3` (recycled!)  
- Position 4: `0 × 0.1 = 0.0` (recycled!)  
- Position 5: `1 × 0.3 = 0.3` (recycled again!)  

---

<div class="callout callout-warning">
<div class="callout-title">Vector Recycling: Power & Peril</div>

**Recycling** repeats the shorter vector to match the longer one.

**Useful for:**
```r
c(1, 2, 3, 4) * 2  # Same as c(1, 2, 3, 4) * c(2, 2, 2, 2)
```

**Dangerous if unexpected:**
```r
c(1, 2, 3, 4, 5) * c(1, 2)  # Did you mean this?
# Result: 1×1, 2×2, 3×1, 4×2, 5×1
```

Always check vector lengths match if you don't intend recycling!
</div>

---

### Your Turn: Practice Vector Operations

**Task:** 
1. Create `myvector` with values 15, 16, 17, 18, 20  
2. Multiply it by `c(0.1, 0.2)`  
3. Think about what recycling will do before running!  

```{r exercise-mult, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Multiply by c(0.1, 0.2)


# Step 3: What did you get? Was it what you expected?


```

```{r exercise-mult-hint}
# Expected pattern:
# 15 × 0.1, 16 × 0.2, 17 × 0.1, 18 × 0.2, 20 × 0.1
myvector * c(0.1, 0.2)
```

```{r exercise-mult-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector * c(0.1, 0.2)
# Result: 1.5, 3.2, 1.7, 3.6, 2.0
```

---

## Quick Reference: Functions and Operators Used in Part 2

<div class="callout callout-info">
<div class="callout-title">Functions and Operators Used in This Tutorial</div>

**Vector Creation:**

| Function/Operator | Description | Example |
|----------|-------------|--------|
| `c()` | Combine values into a vector | `c(1, 2, 3)` |
| `:` | Sequence operator | `1:5` creates 1,2,3,4,5 |

**Type Conversion:**

| Function | Description | Example |
|----------|-------------|--------|
| `as.logical()` | Convert to TRUE/FALSE | `as.logical(c(1,0,1))` |
| `as.numeric()` | Convert to numbers | `as.numeric(c(TRUE, FALSE))` |

**Subsetting:**

| Operator | Description | Example |
|----------|-------------|--------|
| `[ ]` | Extract elements | `x[c(TRUE, FALSE, TRUE)]` |

**Comparison Operators:**

| Operator | Description | Example |
|----------|-------------|--------|
| `==` | Equals | `x == 5` |
| `!=` | Not equals | `x != 5` |
| `>` | Greater than | `x > 10` |
| `<` | Less than | `x < 20` |
| `>=` | Greater or equal | `x >= 10` |
| `<=` | Less or equal | `x <= 20` |

**Logical Operators:**

| Operator | Description | Example |
|----------|-------------|--------|
| `&` | AND (both must be TRUE) | `x > 10 & x < 20` |
| `|` | OR (at least one TRUE) | `x < 10 | x > 20` |
| `!` | NOT (flip TRUE/FALSE) | `!c(TRUE, FALSE)` |

**Arithmetic Operators:**

| Operator | Description | Example |
|----------|-------------|--------|
| `+` | Addition | `x + 5` or `x + y` |
| `-` | Subtraction | `x - 5` or `x - y` |
| `*` | Multiplication | `x * 2` or `x * y` |
| `/` | Division | `x / 2` or `x / y` |
| `%%` | Modulo (remainder) | `x %% 3` |

**Note:** All arithmetic operators work element-wise on vectors and support recycling.

</div>

---

## Congratulations!

You've completed Part 2 of the Introduction to R tutorial!

### What You've Learned:

- Using logical vectors (TRUE/FALSE) for subsetting  
- Type conversion between numeric and logical values  
- Using comparison operators (`==`, `!=`, `>`, `<`, `>=`, `<=`)  
- Conditional subsetting and filtering data  
- Combining conditions with logical operators (AND, OR, NOT)  
- Vector operations (multiplication, element-wise operations)  
- Understanding vector recycling  

### Review: Complete Series

**Part 1 covered:**
- R as a calculator and manual mean calculation  
- Variables and data structures (vectors, matrices, data frames, lists, factors)  
- Descriptive statistics (sd, var, median, range, IQR, quantiles)  
- Z-scores, skewness, and kurtosis  
- R functions and help system  
- Creating vectors and positional subsetting  

**Part 2 covered:**
- Logical subsetting and type conversion  
- Comparison operators and conditional filtering  
- Logical operators for complex conditions  
- Vector operations and recycling  

### Next Steps:

Continue exploring R programming and data manipulation with:
- **data.table** for efficient data manipulation  
- **ggplot2** for data visualization  
- More advanced R programming techniques  

### Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/)  
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)  
- [R Documentation](https://www.rdocumentation.org/)  

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

## Please Give Feedback!

<div class="callout callout-success">
<div class="callout-title">💬 Your Feedback Matters</div>

Help us improve this tutorial! Share your experience and suggestions.
</div>

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/B7Wy7G29wQTuEKjU9" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```
