---
title: "Introduction to R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Load shared data for exercises
library(data.table)
# Reinitialize data.table options (stripped by learnr exercise environment)
if (is.null(getOption("datatable.alloccol"))) data.table:::.onLoad()
# IntroToR variables
myFirstNumber <- 0.1
myFirstVector <- c(3,5,7,9)
x <- 5
name <- ""
mySecondVector <- c("some words", "p", "awesome", "last one")

myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
myvector <- c(15, 16, 17, 18, 20)

# data.table datasets
ddf <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/tidyData.txt", header=T)
ddf_dt <- as.data.table(ddf)
iris_dt <- as.data.table(iris)

# ggplot2 datasets
tidy <- ddf
works <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/works.txt", header = T)
works$Grade <- factor(works$Grade, levels=c("Extremely bad", "Below average", "Average", "Good", "Excellent"))
df <- data.table::fread("http://hex.bioinfo.hr/~mfabijanic/df.txt", header=TRUE, fill=T)
colnames(df) <- "recommend"
df$recommend <- factor(df$recommend, levels=c("Strongly no","No", "Maybe", "Yes", "Strongly yes"))
```

## Welcome

Welcome to **Introduction to R**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## What You Will Learn

<div class="callout callout-success">
<div class="callout-title">Tutorial Contents</div>

In this tutorial, you will master the fundamentals of R programming:

1. **Getting Started with R**  
   - Using R as a calculator  
   - Manually calculate mean in R  
   
2. **Variables and Data Structures**  
   - Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   - Creating variables and vectors  
   - Descriptive statistics (sd, var, median, range, IQR, quantiles)  
   
3. **Functions and Help**  
   - Vector structure functions  
   - Mathematical functions  
   - Getting help in R  
   
4. **Working with Vectors**  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   - Understanding type conversion  
   
5. **Comparisons and Subsetting**  
   - Using comparison operators  
   - Conditional subsetting and filtering  
   
6. **Logical Operators**  
   - Understanding AND, OR, NOT  
   - Multi-condition filtering  

**Additional Content:**  
- Vector Operations (multiplication, element-wise ops, recycling)  
- Quick Reference: All Functions Used in This Tutorial

</div>

---

## Getting Started with R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started with R**  
   → Using R as a calculator  
   - Manually calculate mean in R  

</div>

<div class="callout callout-info">
<div class="callout-title">R as a Calculator</div>

R can perform mathematical operations just like a calculator. This is your first step in learning R programming!
</div>

### How can we use R and RStudio?

- Console
- Script
- Notebook (add new chunk: Ctrl + Alt + I)

Try using R as a calculator. Type `1 + 1` and run the code:

```{r calculator-demo, exercise=TRUE}
# Use R as a calculator


```

```{r calculator-demo-hint}
1 + 1
```

```{r calculator-demo-solution}
1 + 1
```

### Try it!

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

1. **Getting Started with R**  
   - Using R as a calculator  
   → Manually calculate mean in R  

</div>

<div class="callout callout-tip">
<div class="callout-title">Calculating the Mean</div>

The **mean** (average) is calculated by adding all numbers together and dividing by how many numbers there are. For example, the mean of 2, 4, and 6 is `(2 + 4 + 6) / 3 = 4`.
</div>

Calculate the mean value of the numbers 3, 5, 7, and 9 using the operators `+`, `/`, and `()`.

```{r mean-exercise, exercise=TRUE}
# Calculate the mean of 3, 5, 7, 9


```

```{r mean-exercise-hint}
# Add all numbers together: (3 + 5 + 7 + 9)
# Then divide by how many numbers there are: / 4
# Put it together: (3 + 5 + 7 + 9) / 4
```

```{r mean-exercise-solution}
# Calculate the mean
(3 + 5 + 7 + 9) / 4
# Result: 6
```

### What have we learned so far?

**We can use R as a calculator!**

*We can even do better than that...*

---


## Variables and Data Structures

### Data Structures in R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Variables and Data Structures**  
   → Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   - Creating variables and vectors  

</div>

<div class="callout callout-concept">
<div class="callout-title">R Data Structures Overview</div>

R has several fundamental data structures for organizing information:

**1. Vector** - One-dimensional array (all same type)
```
[1, 2, 3, 4, 5]
```

**2. Matrix** - Two-dimensional array (all same type)
```
     [,1] [,2] [,3]
[1,]    1    4    7
[2,]    2    5    8
[3,]    3    6    9
```

**3. Data Frame** - Table with named columns (different types allowed)
```
  name    age  score
1 "Anna"   25   95.5
2 "Bob"    30   87.3
3 "Chris"  22   91.0
```

**4. List** - Collection of any objects (can mix types)
```
[[1]] 10
[[2]] c("a", "b") 
[[3]] data.frame(x=1:3)
```

**5. Factor** - Categorical variable with levels
```
[1] low  high medium high
Levels: low medium high
```

</div>

Here is an examples of how basic R data structure are created: vector, matrix, data frame, list, and factor:

```{r data-structures-visual, exercise=TRUE, exercise.setup="shared-data"}
my_vector <- c(1, 2, 3, 4, 5)
my_vector
```

```{r data-structures-visual2, exercise=TRUE, exercise.setup="shared-data"}
my_matrix <- matrix(1:9, nrow=3, ncol=3)
my_matrix
```

```{r data-structures-visual3, exercise=TRUE, exercise.setup="shared-data"}

my_df <- data.frame(
  name = c("Anna", "Bob", "Chris"),
  age = c(25, 30, 22),
  score = c(95.5, 87.3, 91.0)
)
my_df
```
```{r data-structures-visual4, exercise=TRUE, exercise.setup="shared-data"}

my_list <- list(
  numbers = 1:5,
  text = "hello",
  data = data.frame(x=1:3, y=4:6)
)
my_list
```
```{r data-structures-visual5, exercise=TRUE, exercise.setup="shared-data"}

my_factor <- factor(c("low", "high", "medium", "high"),
                    levels = c("low", "medium", "high"))
my_factor
```

**Key Differences:** 
 
- **Vector & Matrix**: All elements must be the same type (numeric, character, etc.)  
- **Data Frame**: Each column can be a different type (like a spreadsheet)  
- **List**: Most flexible - can contain any mix of objects  
- **Factor**: Stores categorical data efficiently with predefined levels  

---

### Variables

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

2. **Variables and Data Structures**  
   - Understanding R data structures (vectors, matrices, data frames, lists, factors)  
   → Creating variables and vectors  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Variables?</div>

A **variable** is a named container that stores a value in memory. Think of it as a labeled box — you put something inside (a number, text, a whole dataset), and later you can retrieve it by its name.

**Why use variables?**  
- Avoid repeating the same value over and over  
- Make code readable — `height` is clearer than `173.5`  
- Store results of calculations for later use  

**Naming rules:**  
- Must start with a letter (not a number)  
- Can contain letters, numbers, `.` and `_`  
- Case-sensitive: `myVar` and `myvar` are different  

</div>

<div class="callout callout-info">
<div class="callout-title">Working with Variables</div>

In R, we create variables using the **assignment operator** `<-`:

```r
x <- 5         # x now holds the value 5
name <- "Maja"  # name now holds the text "Maja"
```

- To see the value stored in a variable, simply type its name to print its content  
- To store multiple values in a variable, use `c()` — this creates a **vector**  

</div>

- Create a variable `myFirstNumber` with value `0.1` and a vector `myFirstVector` with `c(3, 5, 7, 9)`.   
- Print `myFirstNumber`.  
- now calculate the mean of the four numbers 3,5,7,9 by writing mean(myFirstVector).  

```{r variables-demo, exercise=TRUE, exercise.setup="shared-data"}
# Create variables and print them


```

```{r variables-demo-hint}
myFirstNumber <- 0.1
myFirstVector <- c(3, 5, 7, 9)
print(myFirstNumber)
myFirstNumber
```

```{r variables-demo-solution}
myFirstNumber <- 0.1
myFirstVector <- c(3, 5, 7, 9)
print(myFirstNumber)
mean(myFirstVector)
```

Now use `myFirstVector` to calculate:

1. The **standard deviation** using `sd()`
2. The **variance** using `var()`
3. The **median** using `median()`
4. The **range** (min and max) using `range()`

```{r variables-stats, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c(3, 5, 7, 9)

# Calculate standard deviation


# Calculate variance


# Calculate median


# Calculate range (min and max)


```

```{r variables-stats-hint}
# Standard deviation
sd(myFirstVector)
# Variance
var(myFirstVector)
```

```{r variables-stats-hint-2}
# Median
median(myFirstVector)
# Range
range(myFirstVector)
```

```{r variables-stats-solution}
myFirstVector <- c(3, 5, 7, 9)
sd(myFirstVector)      # 2.581989
var(myFirstVector)      # 6.666667
median(myFirstVector)   # 6
range(myFirstVector)    # 3 9
```

Now calculate:

1. The **interquartile range (IQR)** using `IQR()`
2. The **5th percentile** using `quantile()` with `probs = 0.05`
3. The **95th percentile** using `quantile()` with `probs = 0.95`

```{r variables-quantiles, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c(3, 5, 7, 9)

# Calculate IQR


# Calculate 5th percentile


# Calculate 95th percentile


```

```{r variables-quantiles-hint}
# IQR = difference between 75th and 25th percentile
IQR(myFirstVector)
# quantile() with probs argument
quantile(myFirstVector, probs = 0.05)
```

```{r variables-quantiles-hint-2}
# 95th percentile
quantile(myFirstVector, probs = 0.95)
# You can also get multiple quantiles at once:
quantile(myFirstVector, probs = c(0.05, 0.95))
```

```{r variables-quantiles-solution}
myFirstVector <- c(3, 5, 7, 9)
IQR(myFirstVector)                          # 3
quantile(myFirstVector, probs = 0.05)       # 5th percentile
quantile(myFirstVector, probs = 0.95)       # 95th percentile
quantile(myFirstVector, probs = c(0.05, 0.95))  # both at once
```

---

<!-- NEW CONCEPT: FUNCTIONS -->
## Functions and Help

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   → Vector structure functions  
   - Mathematical functions  
   - Getting help in R

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Built-in Functions</div>

R has thousands of **built-in functions** that perform common tasks:
- **Structure functions**: `length()`, `unique()`, `table()`
- **Math functions**: `sum()`, `mean()`, `sd()`, `median()`
- **Statistical functions**: `summary()`, `min()`, `max()`

Functions take input (arguments) and return output.
</div>

---

<!-- STEP 2: DEMONSTRATE - Structure Functions -->
### Demonstration: Vector Structure Functions

Use `length()`, `unique()`, and `table()` on `someothervector`:

```{r func-structure-demo, exercise=TRUE, exercise.setup="shared-data"}
# Explore vector structure with functions


```

```{r func-structure-demo-hint}
someothervector
length(someothervector)
unique(someothervector)
table(someothervector)
```

```{r func-structure-demo-solution}
someothervector
length(someothervector)
unique(someothervector)
table(someothervector)
```

**What's happening here?**
- `length()` returns the number of elements (5)
- `unique()` removes duplicates (0 and 1)
- `table()` counts how many times each value appears

---

<!-- STEP 3: DEMONSTRATE - Math Functions -->
### Demonstration: Mathematical Functions

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   - Vector structure functions  
   → Mathematical functions  
   - Getting help in R

</div>

Calculate `sum()`, `mean()`, `sd()`, and `summary()` of `someothervector`:

```{r math-functions-demo, exercise=TRUE, exercise.setup="shared-data"}
# Calculate statistics on someothervector


```

```{r math-functions-demo-hint}
someothervector
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```

```{r math-functions-demo-solution}
someothervector
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```

**What's happening here?**
- `sum()` adds all elements: 1+0+1+0+1 = 3
- `mean()` calculates average: 3/5 = 0.6
- `sd()` measures variability
- `summary()` gives min, max, median, mean, quartiles

---

### Z-Score (Standardization)

<div class="callout callout-concept">
<div class="callout-title">What is a Z-Score?</div>

A **z-score** tells you how many standard deviations a value is from the mean. It is calculated as:

$$z = \frac{x - \bar{x}}{s}$$

where $\bar{x}$ is the mean and $s$ is the standard deviation.

**In R you can:**. 
- Calculate manually: `(x - mean(x)) / sd(x)`  
- Use the built-in function: `scale(x)`. 

</div>

**Task:**
Create a vector `scores <- c(70, 85, 90, 60, 95)` and calculate the z-score for each value.

```{r zscore-exercise, exercise=TRUE}
scores <- c(70, 85, 90, 60, 95)

# Calculate z-scores manually


# Or use the built-in scale() function


```

```{r zscore-exercise-hint}
# Manual z-score: subtract the mean, then divide by sd
z_manual <- (scores - mean(scores)) / sd(scores)
z_manual
```

```{r zscore-exercise-hint-2}
# Built-in function (returns a matrix, use as.numeric() to get a vector)
z_builtin <- scale(scores)
z_builtin
```

```{r zscore-exercise-solution}
scores <- c(70, 85, 90, 60, 95)

# Manual calculation
z_manual <- (scores - mean(scores)) / sd(scores)
z_manual

# Built-in function
z_builtin <- scale(scores)
z_builtin
```

---

### Skewness and Kurtosis

<div class="callout callout-concept">
<div class="callout-title">Skewness & Kurtosis</div>

**Skewness** measures the **asymmetry** of a distribution:
- **Skewness = 0**: Symmetric (like a normal distribution)
- **Skewness > 0**: Right-skewed (long tail to the right)
- **Skewness < 0**: Left-skewed (long tail to the left)

**Kurtosis** measures how heavy the **tails** are:
- **Kurtosis ≈ 3** (or excess kurtosis ≈ 0): Normal distribution
- **Kurtosis > 3**: Heavier tails (more outliers)
- **Kurtosis < 3**: Lighter tails (fewer outliers)

R does not have built-in skewness/kurtosis functions, but you can calculate them manually or use a package like `moments`.

**Manual formulas:**

$$\text{skewness} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^3$$

$$\text{kurtosis} = \frac{1}{n} \sum \left(\frac{x_i - \bar{x}}{s}\right)^4$$

</div>

**Task:**
Using the vector `scores <- c(70, 85, 90, 60, 95)`, calculate skewness and kurtosis manually.

```{r skewness-exercise, exercise=TRUE}
scores <- c(70, 85, 90, 60, 95)

# Step 1: Standardize (z-scores)
z <- (scores - mean(scores)) / sd(scores)

# Step 2: Calculate skewness (mean of z^3)


# Step 3: Calculate kurtosis (mean of z^4)


```

```{r skewness-exercise-hint}
# Skewness: raise z-scores to the 3rd power, then take the mean
z <- (scores - mean(scores)) / sd(scores)
skewness <- mean(z^3)
skewness
# Negative value → left-skewed (tail extends to the left)
```

```{r skewness-exercise-hint-2}
# Kurtosis: raise z-scores to the 4th power, then take the mean
z <- (scores - mean(scores)) / sd(scores)
kurtosis <- mean(z^4)
kurtosis
# Excess kurtosis (compared to normal distribution):
excess_kurtosis <- kurtosis - 3
excess_kurtosis
```

```{r skewness-exercise-solution}
scores <- c(70, 85, 90, 60, 95)

# Standardize
z <- (scores - mean(scores)) / sd(scores)

# Skewness (mean of z^3)
skewness <- mean(z^3)
skewness    # negative → left-skewed

# Kurtosis (mean of z^4)
kurtosis <- mean(z^4)
kurtosis

# Excess kurtosis (relative to normal distribution = 3)
excess_kurtosis <- kurtosis - 3
excess_kurtosis
```

---

<!-- STEP 5: GETTING HELP -->
### Getting Help in R

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

3. **Functions and Help**  
   - Vector structure functions  
   - Mathematical functions  
   → Getting help in R

</div>

<div class="callout callout-info">
<div class="callout-title">Help Resources</div>

When you need to learn about a function:
- **`?function_name`** - Opens help documentation
- **`example(function_name)`** - Shows working examples
- **`help.search("keyword")`** - Searches help files

```r
?mean
example(mean)
help.search("average")
```
</div>

Try `?mean` and `example(runif)` to explore R's help system:

```{r help-demo, exercise=TRUE, eval=FALSE}
# Try getting help on functions


```

```{r help-demo-hint, eval=FALSE}
?mean
example(runif)
```

```{r help-demo-solution, eval=FALSE}
?mean
example(runif)
```

---

## Working with Vectors

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   → Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   - Understanding type conversion  

</div>

<div class="callout callout-concept">
<div class="callout-title">What are Vectors?</div>

Vectors are the fundamental data structure in R. They store multiple values of the **same type** (all numbers, all text, etc.). Think of them as a ordered list of values.
</div>

### Creating Vectors with c()

You can make a vector using the function `c()` (concatenate).

Create `mySecondVector` with `c("some words", "p", "awesome", "last one")` and `myFirstSequence` with `1:4`, then print them:

```{r vector-creation, exercise=TRUE, exercise.setup="shared-data"}
# Create vectors and display them


```

```{r vector-creation-hint}
mySecondVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4
mySecondVector
myFirstSequence
```

```{r vector-creation-solution}
mySecondVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4
mySecondVector
myFirstSequence
```

### Subsetting with [ ]

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   → Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   - Understanding type conversion  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Subsetting Vectors</div>

**Subsetting** means extracting specific elements from a vector. Think of it like selecting items from a list. In R, we use square brackets `[ ]` to do this.   

**Two main ways to subset:**.  
- **By position**: Get elements at specific locations (1st, 3rd, etc.).  
- **By condition**: Get elements that meet certain criteria (> 5, == "text", etc.).  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Accessing Single Elements

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`   
   - Subsetting with `[ ]`  
   → Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   - Understanding type conversion  

</div>

Let's start simple - getting just one element by its position:  

Display `mySecondVector` and `myFirstSequence`, then get the 3rd element from each using `[3]`:  

```{r subsetting-single, exercise=TRUE, exercise.setup="shared-data"}
# Get the 3rd element from each vector


```

```{r subsetting-single-hint}
mySecondVector
mySecondVector[3]
```

```{r subsetting-single-solution}
myFirstVector
myFirstSequence
myFirstVector[3]
myFirstSequence[3]
```

**What's happening here?**
- `mySecondVector[3]` says "give me the element at position 3".  
- R uses **1-based indexing** (first element is position 1, not 0).  
- Result: The third item from each vector.  

---

<!-- STEP 3: DEMONSTRATE - Multiple Elements -->
### Demonstration: Accessing Multiple Elements

Now let's get multiple elements at once:   

Get elements at positions 1 and 3 from `mySecondVector` using `c()`, then try storing positions in a variable:   

```{r subsetting-multiple-demo, exercise=TRUE, exercise.setup="shared-data"}
# Get elements at positions 1 and 3


```

```{r subsetting-multiple-demo-hint}
mySecondVector[c(1, 3)]

```

```{r subsetting-multiple-demo-hint-2}
positions <- c(1, 2, 4)
mySecondVector[positions]
```


```{r subsetting-multiple-demo-solution}
mySecondVector[c(1, 3)]
positions <- c(1, 2, 4)
mySecondVector[positions]
```

**What's happening here?**
- `c(1, 3)` creates a vector of positions we want.  
- `mySecondVector[c(1, 3)]` returns elements at those positions.  
- You MUST use `c()` to combine multiple positions.  

---

<!-- STEP 4: COMMON MISTAKE WARNING -->
<div class="callout callout-danger">
<div class="callout-title">Common Mistake: Wrong Subsetting Syntax</div>

**This won't work:**
```r
mySecondVector[1,3]  # ❌ WRONG! This is matrix/data.frame syntax
```

**Why it fails:** Vectors are one-dimensional. The comma `,` is for 2D structures (matrices, data frames).

**Correct way:**
```r
mySecondVector[c(1, 3)]  # ✅ RIGHT! Use c() to combine positions  
```
</div>

---

<!-- STEP 5: EXERCISE - Now Student Tries -->
### Your Turn: Practice Subsetting by Position  

**Task:**   
1. Create a vector named `myvector` with numbers: 15, 16, 17, 18, and 20.  
2. Use subsetting to get the 2nd and 4th elements.  

```{r exercise-subsetting, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Subset to get 2nd and 4th elements (should return 16 and 18)

```

```{r exercise-subsetting-hint}
# Use c() to create a vector
myvector <- c(15, 16, 17, 18, 20)
```

```{r exercise-subsetting-hint-2}
# Use c() to specify positions: c(2, 4)
myvector[c(2, 4)]
```

```{r exercise-subsetting-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(2, 4)]  # Returns 16 and 18
```

---

<!-- NEW CONCEPT: LOGICAL SUBSETTING -->
### Access by Logical Vector

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   → Using logical vectors for subsetting  
   - Understanding type conversion  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Logical Subsetting</div>

Instead of positions (1, 2, 3...), you can use TRUE/FALSE values to select elements:  
- **TRUE** = keep this element  
- **FALSE** = skip this element  

The logical vector must be the same length as your data vector.  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Using TRUE/FALSE to Select Elements

Use `TRUE/FALSE` values to select elements from `mySecondVector`. Select the 1st and 3rd elements:

```{r logical-subsetting-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use TRUE/FALSE to select elements


```

```{r logical-subsetting-demo-hint}
mySecondVector
mySecondVector[c(TRUE, FALSE, TRUE, FALSE)]
```

```{r logical-subsetting-demo-solution}
mySecondVector
mySecondVector[c(TRUE, FALSE, TRUE, FALSE)]
```

**What's happening here?**
- Position 1: TRUE → include "some words"  
- Position 2: FALSE → skip "p"  
- Position 3: TRUE → include "awesome"  
- Position 4: FALSE → skip "last one"  

---

<!-- STEP 3: DEMONSTRATE - Using Variables -->
### Demonstration: Storing Logical Vectors

You can save logical vectors in variables:

Create a logical vector `my_selection` and use it to subset `mySecondVector`:

```{r logical-variable-demo, exercise=TRUE, exercise.setup="shared-data"}
# Store a logical vector and use it for subsetting


```

```{r logical-variable-demo-hint}
my_selection <- c(TRUE, TRUE, FALSE, FALSE)
mySecondVector[my_selection]
```

```{r logical-variable-demo-solution}
my_selection <- c(TRUE, TRUE, FALSE, FALSE)
mySecondVector[my_selection]
```

**Why is this useful?** 
- Logical vectors can be created automatically from conditions (coming next!)  
- Makes code more readable and reusable  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Pro Tip: Logical Vector Length</div>

The logical vector should match the length of your data:
```r
mySecondVector[c(TRUE, FALSE)]  # Only 2 values for 4-element vector?
```

R will **recycle** the pattern: TRUE, FALSE, TRUE, FALSE  

This can be useful but also cause errors if unintended!  
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Logical Subsetting

**Task:** 
Use a logical vector to get the 2nd and 4th elements from `myvector`.

Think about which positions should be TRUE vs FALSE.

```{r exercise-logical, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Create a logical vector to select 2nd and 4th elements


# Use it to subset myvector

```

```{r exercise-logical-hint}
# Think: which positions should be TRUE?
# Position: 1     2     3      4     5
# Logical:  FALSE TRUE  FALSE TRUE  FALSE
```

```{r exercise-logical-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE, FALSE, TRUE, FALSE)]
```

---

<!-- NEW CONCEPT: TYPE CONVERSION -->
### Understanding Type Conversion

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

4. **Working with Vectors**  
   - Creating vectors with `c()`  
   - Subsetting with `[ ]`  
   - Accessing single and multiple elements  
   - Using logical vectors for subsetting  
   → Understanding type conversion  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Automatic Type Conversion</div>

R sometimes converts data types automatically:
- **Numbers to Logical:** 0 = FALSE, any other number = TRUE
- When you use numbers where logical values are expected, R converts them

You can also convert explicitly using `as.logical()`, `as.numeric()`, etc.
</div>

---

<!-- STEP 2: DEMONSTRATE - Automatic Conversion -->
### Demonstration: Numbers as Logical Vectors

Use `someothervector` (which contains 1s and 0s) to subset `mySecondVector`. Observe how R behaves:

```{r conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use a numeric vector with 0s and 1s to subset


```

```{r conversion-demo-hint}
someothervector <- c(1, 0, 1, 0, 1)
mySecondVector[someothervector]
```

```{r conversion-demo-solution}
someothervector <- c(1, 0, 1, 0, 1)
mySecondVector[someothervector]
```

**What's happening here?**
- R sees `someothervector` contains numbers  
- Context requires logical (TRUE/FALSE) for subsetting  
- R does not automatically convert: 1 → TRUE, 0 → FALSE
- Result: Elements at positions 1, 1, 1

---

<!-- STEP 3: DEMONSTRATE - Explicit Conversion -->
### Demonstration: Explicit Conversion with as.logical()

Explicitly convert `someothervector` to logical using `as.logical()` and use it to subset `myFirstVector`:  

```{r explicit-conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
# Explicitly convert to logical and subset


```

```{r explicit-conversion-demo-hint}
someothervector <- c(1, 0, 1, 0, 1)
as.logical(someothervector)
mySecondVector[as.logical(someothervector)]
```

```{r explicit-conversion-demo-solution}
someothervector <- c(1, 0, 1, 0, 1)
as.logical(someothervector)
mySecondVector[as.logical(someothervector)]
```

**Why use explicit conversion?**
- Makes your code clearer and more readable  
- Prevents surprises if R's automatic conversion isn't what you expect  

---

<!-- STEP 4: INFO CALLOUT -->
<div class="callout callout-info">
<div class="callout-title">Conversion Rules</div>

**Numeric to Logical:**
- 0 → FALSE  
- Any other number (1, 2, -5, 0.5, etc.) → TRUE  

**Logical to Numeric:**
- FALSE → 0  
- TRUE → 1  
</div>

---

<!-- NEW CONCEPT: COMPARISONS AND FILTERING -->
## Comparisons and Subsetting

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Comparisons and Subsetting**  
   → Using comparison operators  
   - Conditional subsetting and filtering  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Filtering Data with Comparisons</div>

**Comparison operators** create TRUE/FALSE vectors automatically:  
- `==` equals  
- `!=` not equals  
- `>` greater than  
- `<` less than  
- `>=` greater or equal  
- `<=` less or equal  

These logical vectors can be used for **conditional subsetting** - one of the most powerful data analysis tools!  
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Comparisons -->
### Demonstration: Comparison Operators

Check which elements of `someothervector` equal 1 (`== 1`) and which are greater than 0 (`> 0`):

```{r comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
# Use comparison operators


```

```{r comparison-demo-hint}
someothervector
someothervector == 1
someothervector > 0
```

```{r comparison-demo-solution}
someothervector
someothervector == 1
someothervector > 0
```

**What's happening here?**
- `someothervector == 1` checks each element: is it 1?  
- Returns: `TRUE FALSE TRUE FALSE TRUE`  
- `someothervector > 0` checks: is it greater than 0?  
- Returns logical vector showing which positions meet the condition  

---

<!-- STEP 3: DEMONSTRATE - Using Comparisons for Subsetting -->
### Demonstration: Conditional Subsetting

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

5. **Comparisons and Subsetting**  
   - Using comparison operators  
   → Conditional subsetting and filtering  

</div>

Subset `someothervector` to get only elements greater than 0:

```{r subset-by-comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
# Filter to keep only elements > 0


```

```{r subset-by-comparison-demo-hint}
someothervector
someothervector[someothervector > 0]
```

```{r subset-by-comparison-demo-solution}
someothervector
someothervector[someothervector > 0]
```

**What's happening here?**
1. `someothervector > 0` creates: `TRUE FALSE TRUE FALSE TRUE`  
2. Use this as subsetting index  
3. Result: only elements where condition is TRUE (values: 1, 1, 1)  

**This is filtering!** You're selecting data based on conditions.  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Breaking Down Complex Subsetting</div>

You can split complex subsetting into steps:

```r
# Step by step  
condition <- someothervector > 0  
condition  # See the TRUE/FALSE pattern  
someothervector[condition]  # Apply it  
  
# Or in one line  
someothervector[someothervector > 0]  
```

Both give the same result!
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Conditional Subsetting

**Task:** 
Get only the elements from `myvector` that are **smaller than 17**.  

```{r exercise-comparison, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)  

# Your code here: filter for values < 17

```

```{r exercise-comparison-hint}
# Use the < operator to create a logical vector
myvector[myvector < 17]
```

```{r exercise-comparison-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector < 17]  # Returns 15 and 16
```

---

<!-- NEW CONCEPT: LOGICAL OPERATORS -->
## Logical Operators

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

6. **Logical Operators**  
   → Understanding AND, OR, NOT    
   - Multi-condition filtering  

</div>

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Combining Conditions</div>

Logical operators let you combine multiple TRUE/FALSE conditions:  
- **AND (`&`)**: Both must be TRUE    
- **OR (`|`)**: At least one must be TRUE    
- **NOT (`!`)**: Flips TRUE ↔ FALSE  

These let you create complex filters: "values greater than 10 AND less than 20"
</div>

---

<!-- STEP 2: DEMONSTRATE - AND, OR, NOT -->
### Demonstration: Basic Logical Operations

Create `firstLogical` and `secondLogical` vectors. Then test AND (`&`), OR (`|`), and NOT (`!`) operators:

```{r logical-ops-demo, exercise=TRUE}
# Test logical operators: &, |, !


```

```{r logical-ops-demo-hint}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
!firstLogical
```

```{r logical-ops-demo-solution}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
!firstLogical
```

**What's happening here?**
- `&` (AND): TRUE only when BOTH are TRUE  
  - Position 1: TRUE & TRUE = **TRUE**  
  - Position 2: TRUE & FALSE = **FALSE**  
- `|` (OR): TRUE when AT LEAST ONE is TRUE  
  - Position 2: TRUE | FALSE = **TRUE**  
  - Position 3: FALSE | TRUE = **TRUE**  
- `!` (NOT): Flips each value  
  - TRUE becomes FALSE, FALSE becomes TRUE  

---

<!-- STEP 3: DEMONSTRATE - Combining Conditions for Filtering -->
### Demonstration: Multi-Condition Filtering

<div class="callout callout-success">
<div class="callout-title">📍 In This Section</div>

6. **Logical Operators**  
   - Understanding AND, OR, NOT    
   → Multi-condition filtering  

</div>

Create a `numbers` vector `c(5, 10, 15, 20, 25)`. Use `&` to find numbers between 10 and 20 (inclusive), then use the condition for subsetting:

```{r multi-condition-demo, exercise=TRUE}
# Filter numbers between 10 and 20


```

```{r multi-condition-demo-hint}
numbers <- c(5, 10, 15, 20, 25)
numbers >= 10 & numbers <= 20
numbers[numbers >= 10 & numbers <= 20]
```

```{r multi-condition-demo-solution}
numbers <- c(5, 10, 15, 20, 25)
numbers >= 10 & numbers <= 20
numbers[numbers >= 10 & numbers <= 20]
```

**What's happening here?**
1. `numbers >= 10` → `FALSE TRUE TRUE TRUE TRUE`  
2. `numbers <= 20` → `TRUE TRUE TRUE TRUE FALSE`  
3. Combine with `&` → `FALSE TRUE TRUE TRUE FALSE`  
4. Result: 10, 15, 20 (values meeting BOTH conditions)  

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Truth Table Quick Reference</div>

| A     | B     | A & B | A \| B | !A    |
|-------|-------|-------|--------|-------|
| TRUE  | TRUE  | TRUE  | TRUE   | FALSE |
| TRUE  | FALSE | FALSE | TRUE   | FALSE |
| FALSE | TRUE  | FALSE | TRUE   | TRUE  |
| FALSE | FALSE | FALSE | FALSE  | TRUE  |
</div>

---

<!-- STEP 5: EXERCISE -->
### Extra Challenge!  

Your Turn: Practice Logical Operators

**Task:** 
Find all elements from `myvector` that are **divisible by 3**.

**Hint:** A number is divisible by 3 if `number %% 3 == 0`

```{r exercise-logical-ops, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Find elements divisible by 3

```

```{r exercise-logical-ops-hint}
# A number is divisible by 3 if modulo 3 equals 0
myvector[myvector %% 3 == 0]
```

```{r exercise-logical-ops-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector %% 3 == 0]  # Returns 15 and 18
```

---

### Extra Challenge!

Return all elements from `myvector` which are on EVEN positions (2, 4, 6, ...).

```{r exercise-even-positions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Get elements at even positions

```

```{r exercise-even-positions-hint}
# You can use recycling: c(FALSE, TRUE) will create FALSE, TRUE, FALSE, TRUE, FALSE
myvector[c(FALSE, TRUE)]
```

```{r exercise-even-positions-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE)]  # Uses recycling: positions 2 and 4
```

---

## Additional Content: Vector Operations

<div class="callout callout-info">
<div class="callout-title">📍 Additional Content</div>

**Vector Operations**  
- Multiplication by constants  
- Element-wise operations  
- Vector recycling  

</div>

<div class="callout callout-concept">
<div class="callout-title">Vectorized Operations</div>

One of R's most powerful features is **vectorized operations**:
- Operations apply to **each element automatically**
- No need for loops!
- Works with: `+`, `-`, `*`, `/`, and more

**Example:** `c(1,2,3) * 2` multiplies each element by 2
</div>

---

### Demonstration: Multiplication by a Constant

Multiply `someothervector` by `0.5` and observe the element-wise result:

```{r mult-constant-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply vector by a constant


```

```{r mult-constant-demo-hint}
someothervector
someothervector * 0.5
```

```{r mult-constant-demo-solution}
someothervector
someothervector * 0.5
```

**What's happening here?**
- Each element is multiplied by 0.5
- No loop needed - R does it automatically
- Result: `[1] 0.5 0.0 0.5 0.0 0.5`

---

### Demonstration: Multiplying Two Vectors (Same Length)

Multiply `someothervector` by `1:5` (element by element) and observe the result:

```{r mult-same-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply two vectors of same length


```

```{r mult-same-demo-hint}
someothervector
1:5
someothervector * 1:5
```

```{r mult-same-demo-solution}
someothervector
1:5
someothervector * 1:5
```

**What's happening here?**
- Element 1 × Element 1: `1 × 1 = 1`  
- Element 2 × Element 2: `0 × 2 = 0`  
- Element 3 × Element 3: `1 × 3 = 3`  
- And so on...  

---

### Demonstration: Different Size Vectors (Recycling)

Multiply `someothervector` (length 5) by `c(0.3, 0.1)` (length 2) and observe how R recycles the shorter vector:  

```{r mult-recycle-demo, exercise=TRUE, exercise.setup="shared-data"}
# Multiply vectors of different lengths (recycling)


```

```{r mult-recycle-demo-hint}
someothervector
c(0.3, 0.1)
someothervector * c(0.3, 0.1)
```

```{r mult-recycle-demo-solution}
someothervector
c(0.3, 0.1)
someothervector * c(0.3, 0.1)
```

**What's happening here?**
R repeats `c(0.3, 0.1)` to match the length:  
- Position 1: `1 × 0.3 = 0.3`  
- Position 2: `0 × 0.1 = 0.0`  
- Position 3: `1 × 0.3 = 0.3` (recycled!)  
- Position 4: `0 × 0.1 = 0.0` (recycled!)  
- Position 5: `1 × 0.3 = 0.3` (recycled again!)  

---

<div class="callout callout-warning">
<div class="callout-title">Vector Recycling: Power & Peril</div>

**Recycling** repeats the shorter vector to match the longer one.

**Useful for:**
```r
c(1, 2, 3, 4) * 2  # Same as c(1, 2, 3, 4) * c(2, 2, 2, 2)
```

**Dangerous if unexpected:**
```r
c(1, 2, 3, 4, 5) * c(1, 2)  # Did you mean this?
# Result: 1×1, 2×2, 3×1, 4×2, 5×1
```

Always check vector lengths match if you don't intend recycling!
</div>

---

### Your Turn: Practice Vector Operations

**Task:** 
1. Create `myvector` with values 15, 16, 17, 18, 20  
2. Multiply it by `c(0.1, 0.2)`  
3. Think about what recycling will do before running!  

```{r exercise-mult, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Multiply by c(0.1, 0.2)


# Step 3: What did you get? Was it what you expected?


```

```{r exercise-mult-hint}
# Expected pattern:
# 15 × 0.1, 16 × 0.2, 17 × 0.1, 18 × 0.2, 20 × 0.1
myvector * c(0.1, 0.2)
```

```{r exercise-mult-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector * c(0.1, 0.2)
# Result: 1.5, 3.2, 1.7, 3.6, 2.0
```

---

### ⚠️ BE CAREFUL: Comparing Decimal Numbers with ==

<div class="callout callout-danger">
<div class="callout-title">⚠️ DANGER: Floating Point Precision Issues</div>

**Never use `==` to compare decimal numbers!**

Computers store decimal numbers with limited precision, which can cause unexpected results:

```r
0.1 + 0.2 == 0.3  # You'd expect TRUE, but it returns FALSE!
```

**Why?** Computers store numbers in binary (0s and 1s), and some decimals can't be represented exactly. This causes tiny rounding errors that make "equal" numbers appear unequal.

**The Problem in Filtering:**
```r
x <- c(0.1, 0.2, 0.3, 0.4)
x[x == 0.3]  # Might not find 0.3 due to precision errors!
```

</div>

---

### Demonstration: The Problem with == for Decimals

See the floating point precision problem in action:

```{r floating-point-problem, exercise=TRUE}
# Example 1: Simple arithmetic that "should" work
0.1 + 0.2 == 0.3

# Example 2: Why does this happen?
print(0.1 + 0.2, digits = 20)  # Shows the actual stored value
print(0.3, digits = 20)         # Compare to what we expect

# Example 3: Filtering problem
values <- c(0.1, 0.2, 0.3, 0.4, 0.5)
calculated <- c(0.05 + 0.05, 0.1 + 0.1, 0.15 + 0.15, 0.2 + 0.2, 0.25 + 0.25)

# This might not work as expected!
calculated == 0.3
```

---

### Solution: Use Tolerance or Other Comparisons

<div class="callout callout-success">
<div class="callout-title">✅ SAFE ALTERNATIVES</div>

**Instead of `==`, use one of these methods:**

**1. Use `all.equal()` for tolerance-based comparison:**
```r
all.equal(0.1 + 0.2, 0.3)  # Returns TRUE (within tolerance)
```

**2. Use comparison operators (>, <, >=, <=) when possible:**
```r
x[x > 0.29 & x < 0.31]  # Safer than x[x == 0.3]
```

**3. Round numbers first:**
```r
round(0.1 + 0.2, 10) == round(0.3, 10)  # TRUE
```

**4. Use as.numeric(as.character(x)):**
```r
as.numeric(as.character(0.1 + 0.2 ))== 0.3  # TRUE (built-in tolerance)
```

</div>

---

### Demonstration: Safe Comparison Methods

Compare the unsafe `==` with safer alternatives:

```{r safe-comparisons, exercise=TRUE}
# The problem
0.1 + 0.2 == 0.3  # FALSE (unexpected!)

# Solution 1: all.equal() with tolerance
all.equal(0.1 + 0.2, 0.3)  # TRUE (safe!)

# Solution 2: Use range instead of exact match
x <- 0.1 + 0.2
x > 0.29 & x < 0.31  # TRUE (safe!)

# Solution 3: Round first
round(0.1 + 0.2, 10) == round(0.3, 10)  # TRUE (safe!)

# Solution 4: Check absolute difference is small
abs((0.1 + 0.2) - 0.3) < 0.0000001  # TRUE (safe!)


```

<div class="callout callout-tip">
<div class="callout-title">💡 Pro Tip: When to Worry</div>

**You don't need to worry about this when:**
- Comparing integers (whole numbers): `5 == 5` is always safe  
- Using `>`, `<`, `>=`, `<=` with decimals (these work fine!)  

**You DO need to worry when:**
- Using `==` with any decimal calculations  
- Filtering datasets for specific decimal values  
- Working with financial data or precise measurements  

**Remember:** This is a fundamental limitation of how computers store numbers, not a bug in R!

</div>

---

## Quick Reference: All Functions

<div class="callout callout-info">
<div class="callout-title">Functions Used in This Tutorial</div>

**Creating Objects:**

| Function | Description | Example |
|----------|-------------|--------|
| `c()` | Combine values into a vector | `c(1, 2, 3)` |
| `matrix()` | Create a matrix | `matrix(1:9, nrow=3)` |
| `data.frame()` | Create a data frame | `data.frame(x=1:3, y=4:6)` |
| `list()` | Create a list | `list(a=1, b="text")` |
| `factor()` | Create a factor (categorical) | `factor(c("low", "high"))` |

**Descriptive Statistics:**

| Function | Description | Example |
|----------|-------------|--------|
| `mean()` | Arithmetic mean | `mean(x)` |
| `median()` | Median value | `median(x)` |
| `sd()` | Standard deviation | `sd(x)` |
| `var()` | Variance | `var(x)` |
| `sum()` | Sum of all elements | `sum(x)` |
| `min()` | Minimum value | `min(x)` |
| `max()` | Maximum value | `max(x)` |
| `range()` | Min and max | `range(x)` |
| `IQR()` | Interquartile range | `IQR(x)` |
| `quantile()` | Quantiles / percentiles | `quantile(x, probs=0.95)` |
| `summary()` | Summary statistics | `summary(x)` |
| `scale()` | Standardize (z-scores) | `scale(x)` |

**Vector Structure:**

| Function | Description | Example |
|----------|-------------|--------|
| `length()` | Number of elements | `length(x)` |
| `unique()` | Remove duplicates | `unique(x)` |
| `table()` | Frequency table | `table(x)` |

**Type Conversion:**

| Function | Description | Example |
|----------|-------------|--------|
| `as.logical()` | Convert to TRUE/FALSE | `as.logical(c(1,0,1))` |
| `as.numeric()` | Convert to numbers | `as.numeric(c(TRUE, FALSE))` |

**Output & Help:**

| Function | Description | Example |
|----------|-------------|--------|
| `print()` | Print a value | `print(x)` |
| `?` | Open help page | `?mean` |
| `example()` | Run examples from help | `example(mean)` |
| `help.search()` | Search help files | `help.search("average")` |

</div>

---

## Congratulations!

You've completed the Introduction to R tutorial!

### What You've Learned:

- Using R as a calculator  
- Creating and working with variables  
- Descriptive statistics (sd, var, median, range, IQR, quantiles)  
- Understanding vectors and data structures  
- Useful R functions and getting help in R  
- Subsetting vectors by position and logical conditions  
- Comparison and logical operators  
- Vector operations and recycling (Additional Content)

### Next Steps:

Continue exploring R programming and data visualization with ggplot2!

### Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/)  
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)  
- [R Documentation](https://www.rdocumentation.org/)  

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```

---

## Please Give Feedback!

<div class="callout callout-success">
<div class="callout-title">💬 Your Feedback Matters</div>

Help us improve this tutorial! Share your experience and suggestions.
</div>

```{r feedback-ui, echo=FALSE}
HTML('<a href="https://forms.gle/B7Wy7G29wQTuEKjU9" target="_blank" style="display: block; margin: 20px auto; padding: 15px 30px; background-color: #28a745; color: white; text-align: center; text-decoration: none; border-radius: 5px; font-weight: bold; max-width: 300px;">📝 Please Give Feedback!</a>')
```
