---
title: "Introduction to R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Shared data for exercises
myFirstNumber <- 0.1
myFirstVector <- c("some words","p","awesome", "last one")
myFirstSequence <- 1:4
someothervector <- c(1,0,1,0,1)
```

## Welcome

Welcome to **Introduction to R**

**Author:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o.

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## Getting Started with R

<div class="callout callout-info">
<div class="callout-title">R as a Calculator</div>

R can perform mathematical operations just like a calculator. This is your first step in learning R programming!
</div>

### How can we use R and RStudio?

- Console
- Script
- Notebook (add new chunk: Ctrl + Alt + I)

```{r calculator-demo, exercise=TRUE}
1 + 1

```

### Try it!

<div class="callout callout-tip">
<div class="callout-title">Understanding the %% Operator</div>

The `%%` operator performs modulo division, which returns the **remainder** after division. For example, `13 %% 3` returns `1` because 13 divided by 3 is 4 with a remainder of 1.
</div>

Add your calculation here. Calculate the result of 13%%3, 12%%3, 15%%3.  
Can you guess what %% does?

```{r modulo-exercise, exercise=TRUE}
# Calculate modulo operations
13 %% 3

```

```{r modulo-exercise-hint}
# Try calculating:
# 12 %% 3
# 15 %% 3
# The %% operator gives you the remainder after division
```

```{r modulo-exercise-solution}
13 %% 3  # Result: 1
12 %% 3  # Result: 0
15 %% 3  # Result: 0
# %% gives the remainder (modulo operation)
```

### What have we learned so far?

**We can use R as a calculator!**

*We can even do better than that...*

---

## Working with Vectors

<div class="callout callout-concept">
<div class="callout-title">What are Vectors?</div>

Vectors are the fundamental data structure in R. They store multiple values of the **same type** (all numbers, all text, etc.). Think of them as a ordered list of values.
</div>

### Variables and Data Structures

Variables store values that you can reuse throughout your code. In R, we use the assignment operator `<-` to assign values to variables.

```{r variables-demo, exercise=TRUE, exercise.setup="shared-data"}
myFirstNumber <- 0.1
myFirstVector <- c(2, 3, 7, 8)

# Call the variable by name
print(myFirstNumber)
myFirstNumber
```

### Data Structures in R

<div class="callout callout-concept">
<div class="callout-title">R Data Structures Overview</div>

R has several fundamental data structures for organizing information:

**1. Vector** - One-dimensional array (all same type)
```
[1, 2, 3, 4, 5]
```

**2. Matrix** - Two-dimensional array (all same type)
```
     [,1] [,2] [,3]
[1,]    1    4    7
[2,]    2    5    8
[3,]    3    6    9
```

**3. Data Frame** - Table with named columns (different types allowed)
```
  name    age  score
1 "Anna"   25   95.5
2 "Bob"    30   87.3
3 "Chris"  22   91.0
```

**4. List** - Collection of any objects (can mix types)
```
[[1]] 10
[[2]] c("a", "b") 
[[3]] data.frame(x=1:3)
```

**5. Factor** - Categorical variable with levels
```
[1] low  high medium high
Levels: low medium high
```

</div>

```{r data-structures-visual, exercise=TRUE}
# Examples of each data structure:

# 1. VECTOR (atomic, homogeneous)
my_vector <- c(1, 2, 3, 4, 5)
my_vector

# 2. MATRIX (2D, homogeneous)
my_matrix <- matrix(1:9, nrow=3, ncol=3)
my_matrix

# 3. DATA FRAME (2D table, heterogeneous columns)
my_df <- data.frame(
  name = c("Anna", "Bob", "Chris"),
  age = c(25, 30, 22),
  score = c(95.5, 87.3, 91.0)
)
my_df

# 4. LIST (can contain anything)
my_list <- list(
  numbers = 1:5,
  text = "hello",
  data = data.frame(x=1:3, y=4:6)
)
my_list

# 5. FACTOR (categorical)
my_factor <- factor(c("low", "high", "medium", "high"),
                    levels = c("low", "medium", "high"))
my_factor
```

**Key Differences:**
- **Vector & Matrix**: All elements must be the same type (numeric, character, etc.)
- **Data Frame**: Each column can be a different type (like a spreadsheet)
- **List**: Most flexible - can contain any mix of objects
- **Factor**: Stores categorical data efficiently with predefined levels

### Creating Vectors with c()

You can make a vector using the function `c()` (concatenate).

```{r vector-creation, exercise=TRUE, exercise.setup="shared-data"}
myFirstVector <- c("some words", "p", "awesome", "last one")
myFirstSequence <- 1:4

# Print the vectors
myFirstVector
myFirstSequence
```

### Subsetting with [ ]

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Subsetting Vectors</div>

**Subsetting** means extracting specific elements from a vector. Think of it like selecting items from a list. In R, we use square brackets `[ ]` to do this.

**Two main ways to subset:**
- **By position**: Get elements at specific locations (1st, 3rd, etc.)
- **By condition**: Get elements that meet certain criteria (> 5, == "text", etc.)
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Accessing Single Elements

Let's start simple - getting just one element by its position:

```{r subsetting-single, exercise=TRUE, exercise.setup="shared-data"}
# Look at our vectors first
myFirstVector
myFirstSequence

# Get the 3rd element from each
myFirstVector[3]
myFirstSequence[3]
```

**What's happening here?**
- `myFirstVector[3]` says "give me the element at position 3"
- R uses **1-based indexing** (first element is position 1, not 0)
- Result: The third item from each vector

---

<!-- STEP 3: DEMONSTRATE - Multiple Elements -->
### Demonstration: Accessing Multiple Elements

Now let's get multiple elements at once:

```{r subsetting-multiple-demo, exercise=TRUE, exercise.setup="shared-data"}
# Get elements at positions 1 and 3
myFirstVector[c(1, 3)]

# We can also use a variable to store positions
positions <- c(1, 2, 4)
myFirstVector[positions]
```

**What's happening here?**
- `c(1, 3)` creates a vector of positions we want
- `myFirstVector[c(1, 3)]` returns elements at those positions
- You MUST use `c()` to combine multiple positions

---

<!-- STEP 4: COMMON MISTAKE WARNING -->
<div class="callout callout-danger">
<div class="callout-title">Common Mistake: Wrong Subsetting Syntax</div>

**This won't work:**
```r
myFirstVector[1,3]  # ❌ WRONG! This is matrix/data.frame syntax
```

**Why it fails:** Vectors are one-dimensional. The comma `,` is for 2D structures (matrices, data frames).

**Correct way:**
```r
myFirstVector[c(1, 3)]  # ✅ RIGHT! Use c() to combine positions
```
</div>

---

<!-- STEP 5: EXERCISE - Now Student Tries -->
### Your Turn: Practice Subsetting by Position

**Task:** 
1. Create a vector named `myvector` with numbers: 15, 16, 17, 18, and 20
2. Use subsetting to get the 2nd and 4th elements

```{r exercise-subsetting, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Subset to get 2nd and 4th elements (should return 16 and 18)

```

```{r exercise-subsetting-hint}
# Use c() to specify positions: c(2, 4)
myvector[c(2, 4)]
```

```{r exercise-subsetting-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(2, 4)]  # Returns 16 and 18
```

---

<!-- NEW CONCEPT: LOGICAL SUBSETTING -->
### Access by Logical Vector

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Logical Subsetting</div>

Instead of positions (1, 2, 3...), you can use TRUE/FALSE values to select elements:
- **TRUE** = keep this element
- **FALSE** = skip this element

The logical vector must be the same length as your data vector.
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Example -->
### Demonstration: Using TRUE/FALSE to Select Elements

```{r logical-subsetting-demo, exercise=TRUE, exercise.setup="shared-data"}
# Our vector
myFirstVector

# Use TRUE/FALSE to select
myFirstVector[c(TRUE, FALSE, TRUE, FALSE)]
# Result: 1st and 3rd elements (where TRUE)
```

**What's happening here?**
- Position 1: TRUE → include "some words"
- Position 2: FALSE → skip "p"
- Position 3: TRUE → include "awesome"
- Position 4: FALSE → skip "last one"

---

<!-- STEP 3: DEMONSTRATE - Using Variables -->
### Demonstration: Storing Logical Vectors

You can save logical vectors in variables:

```{r logical-variable-demo, exercise=TRUE, exercise.setup="shared-data"}
# Create a logical vector
my_selection <- c(TRUE, TRUE, FALSE, FALSE)

# Use it for subsetting
myFirstVector[my_selection]
```

**Why is this useful?** 
- Logical vectors can be created automatically from conditions (coming next!)
- Makes code more readable and reusable

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Pro Tip: Logical Vector Length</div>

The logical vector should match the length of your data:
```r
myFirstVector[c(TRUE, FALSE)]  # Only 2 values for 4-element vector?
```

R will **recycle** the pattern: TRUE, FALSE, TRUE, FALSE

This can be useful but also cause errors if unintended!
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Logical Subsetting

**Task:** 
Use a logical vector to get the 2nd and 4th elements from `myvector`.

Think about which positions should be TRUE vs FALSE.

```{r exercise-logical, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Create a logical vector to select 2nd and 4th elements


# Use it to subset myvector

```

```{r exercise-logical-hint}
# Think: which positions should be TRUE?
# Position: 1     2     3      4     5
# Logical:  FALSE TRUE  FALSE TRUE  FALSE
```

```{r exercise-logical-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE, FALSE, TRUE, FALSE)]
```

---

<!-- NEW CONCEPT: TYPE CONVERSION -->
### Understanding Type Conversion

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Automatic Type Conversion</div>

R sometimes converts data types automatically:
- **Numbers to Logical:** 0 = FALSE, any other number = TRUE
- When you use numbers where logical values are expected, R converts them

You can also convert explicitly using `as.logical()`, `as.numeric()`, etc.
</div>

---

<!-- STEP 2: DEMONSTRATE - Automatic Conversion -->
### Demonstration: Numbers as Logical Vectors

```{r conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
# A numeric vector with 0s and 1s
someothervector <- c(1, 0, 1, 0, 1)

# Use it to subset (R converts to logical!)
myFirstVector[someothervector]
```

**What's happening here?**
- R sees `someothervector` contains numbers
- Context requires logical (TRUE/FALSE) for subsetting
- R automatically converts: 1 → TRUE, 0 → FALSE
- Result: Elements at positions 1, 3, 5

---

<!-- STEP 3: DEMONSTRATE - Explicit Conversion -->
### Demonstration: Explicit Conversion with as.logical()

```{r explicit-conversion-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector <- c(1, 0, 1, 0, 1)

# Explicitly convert to logical
as.logical(someothervector)

# Use for subsetting
myFirstVector[as.logical(someothervector)]
```

**Why use explicit conversion?**
- Makes your code clearer and more readable
- Prevents surprises if R's automatic conversion isn't what you expect

---

<!-- STEP 4: INFO CALLOUT -->
<div class="callout callout-info">
<div class="callout-title">Conversion Rules</div>

**Numeric to Logical:**
- 0 → FALSE
- Any other number (1, 2, -5, 0.5, etc.) → TRUE

**Logical to Numeric:**
- FALSE → 0
- TRUE → 1
</div>

---

<!-- NEW MAJOR SECTION: VECTOR OPERATIONS -->
## Vector Operations

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Vectorized Operations</div>

One of R's most powerful features is **vectorized operations**:
- Operations apply to **each element automatically**
- No need for loops!
- Works with: `+`, `-`, `*`, `/`, and more

**Example:** `c(1,2,3) * 2` multiplies each element by 2
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Operations -->
### Demonstration: Multiplication by a Constant

```{r mult-constant-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector
someothervector * 0.5
```

**What's happening here?**
- Each element is multiplied by 0.5
- No loop needed - R does it automatically
- Result: `[1] 0.5 0.0 0.5 0.0 0.5`

---

<!-- STEP 3: DEMONSTRATE - Vector by Vector (Same Size) -->
### Demonstration: Multiplying Two Vectors (Same Length)

```{r mult-same-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector
1:5

# Element-by-element multiplication
someothervector * 1:5
```

**What's happening here?**
- Element 1 × Element 1: `1 × 1 = 1`
- Element 2 × Element 2: `0 × 2 = 0`
- Element 3 × Element 3: `1 × 3 = 3`
- And so on...

---

<!-- STEP 4: DEMONSTRATE - Vector Recycling -->
### Demonstration: Different Size Vectors (Recycling)

```{r mult-recycle-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector  # Length 5
c(0.3, 0.1)      # Length 2

# R recycles the shorter vector
someothervector * c(0.3, 0.1)
```

**What's happening here?**
R repeats `c(0.3, 0.1)` to match the length:
- Position 1: `1 × 0.3 = 0.3`
- Position 2: `0 × 0.1 = 0.0`
- Position 3: `1 × 0.3 = 0.3` (recycled!)
- Position 4: `0 × 0.1 = 0.0` (recycled!)
- Position 5: `1 × 0.3 = 0.3` (recycled again!)

---

<!-- STEP 5: WARNING ABOUT RECYCLING -->
<div class="callout callout-warning">
<div class="callout-title">Vector Recycling: Power & Peril</div>

**Recycling** repeats the shorter vector to match the longer one.

**Useful for:**
```r
c(1, 2, 3, 4) * 2  # Same as c(1, 2, 3, 4) * c(2, 2, 2, 2)
```

**Dangerous if unexpected:**
```r
c(1, 2, 3, 4, 5) * c(1, 2)  # Did you mean this?
# Result: 1×1, 2×2, 3×1, 4×2, 5×1
```

Always check vector lengths match if you don't intend recycling!
</div>

---

<!-- STEP 6: EXERCISE -->
### Your Turn: Practice Vector Operations

**Task:** 
1. Create `myvector` with values 15, 16, 17, 18, 20
2. Multiply it by `c(0.1, 0.2)`
3. Think about what recycling will do before running!

```{r exercise-mult, exercise=TRUE}
# Step 1: Create your vector


# Step 2: Multiply by c(0.1, 0.2)


# Step 3: What did you get? Was it what you expected?


```

```{r exercise-mult-hint}
# Expected pattern:
# 15 × 0.1, 16 × 0.2, 17 × 0.1, 18 × 0.2, 20 × 0.1
myvector * c(0.1, 0.2)
```

```{r exercise-mult-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector * c(0.1, 0.2)
# Result: 1.5, 3.2, 1.7, 3.6, 2.0
```

---

<!-- NEW CONCEPT: COMPARISONS AND FILTERING -->
## Comparisons and Subsetting

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Filtering Data with Comparisons</div>

**Comparison operators** create TRUE/FALSE vectors automatically:
- `==` equals
- `!=` not equals
- `>` greater than
- `<` less than
- `>=` greater or equal
- `<=` less or equal

These logical vectors can be used for **conditional subsetting** - one of the most powerful data analysis tools!
</div>

---

<!-- STEP 2: DEMONSTRATE - Simple Comparisons -->
### Demonstration: Comparison Operators

```{r comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector

# Which elements equal 1?
someothervector == 1

# Which elements are greater than 0?
someothervector > 0
```

**What's happening here?**
- `someothervector == 1` checks each element: is it 1?
- Returns: `TRUE FALSE TRUE FALSE TRUE`
- `someothervector > 0` checks: is it greater than 0?
- Returns logical vector showing which positions meet the condition

---

<!-- STEP 3: DEMONSTRATE - Using Comparisons for Subsetting -->
### Demonstration: Conditional Subsetting

```{r subset-by-comparison-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector

# Get only elements greater than 0
someothervector[someothervector > 0]
```

**What's happening here?**
1. `someothervector > 0` creates: `TRUE FALSE TRUE FALSE TRUE`
2. Use this as subsetting index
3. Result: only elements where condition is TRUE (values: 1, 1, 1)

**This is filtering!** You're selecting data based on conditions.

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Breaking Down Complex Subsetting</div>

You can split complex subsetting into steps:

```r
# Step by step
condition <- someothervector > 0
condition  # See the TRUE/FALSE pattern
someothervector[condition]  # Apply it

# Or in one line
someothervector[someothervector > 0]
```

Both give the same result!
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Conditional Subsetting

**Task:** 
Get only the elements from `myvector` that are **smaller than 17**.

```{r exercise-comparison, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Your code here: filter for values < 17

```

```{r exercise-comparison-hint}
# Use the < operator to create a logical vector
myvector[myvector < 17]
```

```{r exercise-comparison-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector < 17]  # Returns 15 and 16
```

---

<!-- NEW CONCEPT: LOGICAL OPERATORS -->
## Logical Operators

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Combining Conditions</div>

Logical operators let you combine multiple TRUE/FALSE conditions:
- **AND (`&`)**: Both must be TRUE
- **OR (`|`)**: At least one must be TRUE  
- **NOT (`!`)**: Flips TRUE ↔ FALSE

These let you create complex filters: "values greater than 10 AND less than 20"
</div>

---

<!-- STEP 2: DEMONSTRATE - AND, OR, NOT -->
### Demonstration: Basic Logical Operations

```{r logical-ops-demo, exercise=TRUE}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)

# AND: both must be TRUE
firstLogical & secondLogical

# OR: at least one must be TRUE
firstLogical | secondLogical

# NOT: flip the values
!firstLogical
```

**What's happening here?**
- `&` (AND): TRUE only when BOTH are TRUE
  - Position 1: TRUE & TRUE = **TRUE**
  - Position 2: TRUE & FALSE = **FALSE**
- `|` (OR): TRUE when AT LEAST ONE is TRUE
  - Position 2: TRUE | FALSE = **TRUE**
  - Position 3: FALSE | TRUE = **TRUE**
- `!` (NOT): Flips each value
  - TRUE becomes FALSE, FALSE becomes TRUE

---

<!-- STEP 3: DEMONSTRATE - Combining Conditions for Filtering -->
### Demonstration: Multi-Condition Filtering

```{r multi-condition-demo, exercise=TRUE}
numbers <- c(5, 10, 15, 20, 25)

# Get numbers between 10 and 20 (inclusive)
numbers >= 10 & numbers <= 20

# Use for subsetting
numbers[numbers >= 10 & numbers <= 20]
```

**What's happening here?**
1. `numbers >= 10` → `FALSE TRUE TRUE TRUE TRUE`
2. `numbers <= 20` → `TRUE TRUE TRUE TRUE FALSE`
3. Combine with `&` → `FALSE TRUE TRUE TRUE FALSE`
4. Result: 10, 15, 20 (values meeting BOTH conditions)

---

<!-- STEP 4: TIP -->
<div class="callout callout-tip">
<div class="callout-title">Truth Table Quick Reference</div>

| A     | B     | A & B | A \| B | !A    |
|-------|-------|-------|--------|-------|
| TRUE  | TRUE  | TRUE  | TRUE   | FALSE |
| TRUE  | FALSE | FALSE | TRUE   | FALSE |
| FALSE | TRUE  | FALSE | TRUE   | TRUE  |
| FALSE | FALSE | FALSE | FALSE  | TRUE  |
</div>

---

<!-- STEP 5: EXERCISE -->
### Your Turn: Practice Logical Operators

**Task:** 
Find all elements from `myvector` that are **divisible by 3**.

**Hint:** A number is divisible by 3 if `number %% 3 == 0`

```{r exercise-logical-ops, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Find elements divisible by 3

```

```{r exercise-logical-ops-hint}
# A number is divisible by 3 if modulo 3 equals 0
myvector[myvector %% 3 == 0]
```

```{r exercise-logical-ops-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[myvector %% 3 == 0]  # Returns 15 and 18
```

---

### Extra Challenge!

Return all elements from `myvector` which are on EVEN positions (2, 4, 6, ...).

```{r exercise-even-positions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Get elements at even positions

```

```{r exercise-even-positions-hint}
# You can use recycling: c(FALSE, TRUE) will create FALSE, TRUE, FALSE, TRUE, FALSE
myvector[c(FALSE, TRUE)]
```

```{r exercise-even-positions-solution}
myvector <- c(15, 16, 17, 18, 20)
myvector[c(FALSE, TRUE)]  # Uses recycling: positions 2 and 4
```

---

<!-- NEW CONCEPT: FUNCTIONS -->
## Functions and Help

<!-- STEP 1: EXPLAIN THE CONCEPT -->
<div class="callout callout-concept">
<div class="callout-title">Built-in Functions</div>

R has thousands of **built-in functions** that perform common tasks:
- **Structure functions**: `length()`, `unique()`, `table()`
- **Math functions**: `sum()`, `mean()`, `sd()`, `median()`
- **Statistical functions**: `summary()`, `min()`, `max()`

Functions take input (arguments) and return output.
</div>

---

<!-- STEP 2: DEMONSTRATE - Structure Functions -->
### Demonstration: Vector Structure Functions

```{r func-structure-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector

# How many elements?
length(someothervector)

# What unique values?
unique(someothervector)

# Count each value
table(someothervector)
```

**What's happening here?**
- `length()` returns the number of elements (5)
- `unique()` removes duplicates (0 and 1)
- `table()` counts how many times each value appears

---

<!-- STEP 3: DEMONSTRATE - Math Functions -->
### Demonstration: Mathematical Functions

```{r math-functions-demo, exercise=TRUE, exercise.setup="shared-data"}
someothervector

# Calculate statistics
sum(someothervector)       # Add all elements
mean(someothervector)      # Average
sd(someothervector)        # Standard deviation
summary(someothervector)   # Quick overview
```

**What's happening here?**
- `sum()` adds all elements: 1+0+1+0+1 = 3
- `mean()` calculates average: 3/5 = 0.6
- `sd()` measures variability
- `summary()` gives min, max, median, mean, quartiles

---

<!-- STEP 4: EXERCISE -->
### Your Turn: Practice Using Functions

**Task:** 
Calculate the sum, mean, and standard deviation of `myvector`.

```{r exercise-functions, exercise=TRUE}
myvector <- c(15, 16, 17, 18, 20)

# Calculate sum


# Calculate mean


# Calculate standard deviation


```

```{r exercise-functions-hint}
# Use sum(), mean(), and sd() functions
sum(myvector)
mean(myvector)
sd(myvector)
```

```{r exercise-functions-solution}
myvector <- c(15, 16, 17, 18, 20)
sum(myvector)
mean(myvector)
sd(myvector)
```

---

<!-- STEP 5: GETTING HELP -->
### Getting Help in R

<div class="callout callout-info">
<div class="callout-title">Help Resources</div>

When you need to learn about a function:
- **`?function_name`** - Opens help documentation
- **`example(function_name)`** - Shows working examples
- **`help.search("keyword")`** - Searches help files

```r
?mean
example(mean)
help.search("average")
```
</div>

```{r help-demo, exercise=TRUE, eval=FALSE}
# Try these commands:
?mean
example(runif)
```

---

## Congratulations!

You've completed the Introduction to R tutorial!

### What You've Learned:

- Using R as a calculator  
- Creating and working with variables  
- Understanding vectors and data structures  
- Subsetting vectors by position and logical conditions  
- Vector operations and recycling  
- Comparison and logical operators  
- Useful R functions  
- How to get help in R

### Next Steps:

Continue exploring R programming and data analysis!

### Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/)
- [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)
- [R Documentation](https://www.rdocumentation.org/)

## Download Your Work

<div class="callout callout-tip">
<div class="callout-title">Save Your Work</div>

Save this completed tutorial as an HTML file for future reference.
</div>

```{r download-ui, echo=FALSE}
HTML('<button onclick="window.downloadTutorial(); return false;" class="btn-download-tutorial" style="display: block; margin: 20px auto;">📥 Download Tutorial as HTML</button>')
```
