---
title: "Your Tutorial Title"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Shared data for exercises

```

## Welcome

Welcome to [Your Workshop Name]

**Authors:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o., [Add other authors]

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## Section 1: Your Topic

Brief introduction to your topic.
Your First Section
### Demonstration

```{r demo-1, exercise=TRUE, exercise.setup="shared-data"}
# Add demonstration code here


```

### Exercise 1: Your Exercise

Instructions for your exercise.

```{r exercise-1, exercise=TRUE, exercise.setup="shared-data"}
# Students write their code here


```

```{r exercise-1-hint}
# Hint
```

```{r exercise-1-solution}
# Solution
```

---

## Congratulations!

You've completed [Tutorial Name]!

### What You've Learned:

✅ Topic 1  
✅ Topic 2  
✅ Topic 3  

### Next Steps:

Continue to **[Next Tutorial Name]**

---

## Resources

- [Resource 1](https://example.com)
- [Resource 2](https://example.com)
