---
title: "Your Tutorial Title"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-branding, echo=FALSE}
# Load Kuzman Consulting branding (CSS + logos)
source("../assets/setup_branding.R", local = knitr::knit_global())
setup_kuzman_branding()
```

```{r shared-data, exercise.setup=TRUE, include=FALSE}
# Shared data for exercises

```

```{r server-setup, context="server"}
# Register resource path for logos
source("../assets/setup_branding.R", local = TRUE)
setup_branding_server()
```

```{r download-ui, echo=FALSE}
# Download button
shiny::HTML('
<div style="text-align: center; margin: 30px 0;">
  <button onclick="window.downloadTutorial()" class="download-btn">
    📥 Download Your Tutorial as HTML
  </button>
</div>
')
```

## Welcome

Welcome to [Your Workshop Name]

**Authors:** dr.sc. Maja Kuzman - Kuzman Consulting d.o.o., [Add other authors]

### How to Use This Tutorial

- Click **"Run Code"** to execute R code
- Click **"Submit Answer"** to check your solutions
- Click **"Hint"** if you need help
- Progress is tracked as you complete exercises

---

## Section 1: Your Topic

Brief introduction to your topic.

<!-- Example: How to add callout boxes -->
<!-- Uncomment and customize the callout type you need: -->

<!-- Success callout (green) - use for positive outcomes, achievements -->
<!--
<div class="callout callout-success">
<div class="callout-title">Success Title</div>
Your success message or achievement here.
</div>
-->

<!-- Info callout (blue) - use for important information, definitions -->
<!--
<div class="callout callout-info">
<div class="callout-title">Info Title</div>
Your informational content here.
</div>
-->

<!-- Warning callout (yellow) - use for cautions, things to watch out for -->
<!--
<div class="callout callout-warning">
<div class="callout-title">Warning Title</div>
Your warning message here.
</div>
-->

<!-- Danger callout (red) - use for errors, common mistakes, critical issues -->
<!--
<div class="callout callout-danger">
<div class="callout-title">Danger Title</div>
Your danger/error message here.
</div>
-->

<!-- Tip callout (orange) - use for helpful tips, best practices -->
<!--
<div class="callout callout-tip">
<div class="callout-title">Tip Title</div>
Your helpful tip here.
</div>
-->

<!-- Concept callout (purple) - use for key concepts, theoretical explanations -->
<!--
<div class="callout callout-concept">
<div class="callout-title">Concept Title</div>
Your concept explanation here.
</div>
-->

### Demonstration

```{r demo-1, exercise=TRUE, exercise.setup="shared-data"}
# Add demonstration code here


```

### Exercise 1: Your Exercise

Instructions for your exercise.

```{r exercise-1, exercise=TRUE, exercise.setup="shared-data"}
# Students write their code here


```

```{r exercise-1-hint}
# Hint
```

```{r exercise-1-solution}
# Solution
```

---

<!-- Download Section: Placed before Congratulations so students can download their completed work -->
```{r download-section, echo=FALSE}
shiny::HTML('
<div style="text-align: center; margin: 40px 0; padding: 30px; background: linear-gradient(135deg, #667eea 0%, #764ba2 100%); border-radius: 15px; box-shadow: 0 10px 30px rgba(0,0,0,0.2);">
  <h2 style="color: white; margin-bottom: 15px; font-size: 28px;">📥 Save Your Work</h2>
  <p style="color: white; margin-bottom: 25px; font-size: 16px;">Download an HTML copy of this tutorial with all your code and results</p>
  <button onclick="window.downloadTutorial()" class="download-btn" style="background: white; color: #667eea; font-weight: bold; font-size: 18px; padding: 15px 40px; border: none; border-radius: 50px; cursor: pointer; box-shadow: 0 5px 15px rgba(0,0,0,0.3); transition: all 0.3s;" onmouseover="this.style.transform=\'scale(1.05)\'; this.style.boxShadow=\'0 8px 25px rgba(0,0,0,0.4)\'" onmouseout="this.style.transform=\'scale(1)\'; this.style.boxShadow=\'0 5px 15px rgba(0,0,0,0.3)\'">
    📥 Download Tutorial
  </button>
</div>
')
```

---

## Congratulations!

You've completed [Tutorial Name]!

### What You've Learned:

- Topic 1  
- Topic 2  
- Topic 3  

### Next Steps:

Continue to **[Next Tutorial Name]**

---

## Resources

- [Resource 1](https://example.com)
- [Resource 2](https://example.com)
